\metatitle{Monomial slide polynomials and fundamental slide polynomials}
\metadescription{An introduction to monomial slide polynomials and fundamental slide polynomials, their definitions, properties, and connections to quasisymmetric functions and Schubert polynomials.} 

\section[slideM]{Monomial slide polynomials}

\begin{polydata}{slideM}
  Name  & Monomial slide polynomials \\
  Space    & All \\
  Basis    & True \\
  Rating   & 1 \\
  Bib      & AssafSearles2016 \\
  Year     & 2016 \\
  Symbol   & $\slideM_\alpha(\xvec)$ \\
  Category & Schur \\
  Contains & qmonom | AssafSearles2016 \\
  StableLimit & qmonom | AssafSearles2016 \\
  PositiveIn & monomial | AssafSearles2016 | scope=nonsymmetric monomials; proof=definition \\
\end{polydata}

The \defin{monomial slide polynomials} were introduced by \name{Sami Assaf} and \name{Dominic Searles} in \cite{AssafSearles2016}.
The slide polynomials form a basis for the space of polynomials, and can be seen as a lift 
of the \hyperref[qmonom]{monomial quasisymmetric functions}.


\begin{definition}
For $\alpha$ being a weak composition, we set
\[
\slideM_\alpha(\xvec) \coloneqq \sum_{\substack{b \trianglerighteq a \\ \mathrm{flat}(b)=\mathrm{flat}(a)}} \xvec^b
\]
where $\mathrm{flat}(\beta)$ denotes the composition obtained by removing all 0s from $\beta$,
and $\trianglerighteq$ denotes \hyperref[partialOrderingsOnPartitions]{dominance order}.
\end{definition}

As an example (from \cite[Eq. 3.7]{AssafSearles2016}),
\[
\slideM_{(0,2,0,3)}(\xvec)
=
x_1^2x_2^3 + x_1^2x_3^3 + x_1^2 x_4^3
+ x_2^2 x_3^3  + x_2^2 x_4^3.
\]



The monomial slide polynomials have the \hyperref[qmonom]{monomial quasisymmetric functions}
as stable limit:
\[
 \lim_{m \to \infty} \slideM_{0^m\times \alpha}(\xvec) = \qmonom_{\mathrm{flat}(\alpha)}(\xvec).
\]


\section[slideF]{Fundamental slide polynomials}

\begin{polydata}{slideF}
  Name &  Fundamental slide polynomials \\
  Space    & All \\
  Basis    & True \\
  Rating   & 1 \\
  Bib      & AssafSearles2016 \\
  Year     & 2016 \\
  Symbol   & $\slideF_\alpha(\xvec)$ \\
  Category & Schur \\
  Contains & gessel | AssafSearles2016 \\
  StableLimit & gessel | AssafSearles2016 \\
  PositiveIn & slideM | AssafSearles2016 \\
  PositiveIn & fundamentalParticle | Searles2019PolynomialBases,HicksNiese2024x | proof=Thm. 3.13 and Cor. 3.11 \\
\end{polydata}

The \emph{fundamental slide polynomials} (or sometimes just \defin{slide polynomials})
were introduced by \name{Sami Assaf} and \name{Dominic Searles} in \cite{AssafSearles2016}.
The slide polynomials form a basis for the space of polynomials, and can be seen as a lift 
of the \hyperref[gessel]{Gessel quasisymmetric functions}.
\name{Oliver Pechenik} and \name{Dominic Searles} give a survey of
asymmetric function theory \cite{PechenikSearles2020}.  It is a useful entry
point for slide, glide, key, Demazure atom, and related nonsymmetric or
quasisymmetric families.

The $K$-theoretical analog of fundamental slide 
polynomials are the \emph{glide polynomials}, see \cite{PechenikSearles2017}.
They refine stable Grothendieck polynomials in the same way that fundamental
slide polynomials refine Schubert polynomials.
\todo{Add a small checked glide-polynomial example, preferably next to the
fundamental slide example and backed by Rust or a paper computation.}


The fundamental slide polynomials expand positively 
in the \hyperref[slideM]{monomial slide polynomials}.
Moreover, the \hyperref[schubert]{Schubert polynomials}
expand positively in the fundamental slide polynomials, \cite[Thm. 3.13]{AssafSearles2016}.
The main motivation for introducing the fundamental slide polynomials,
is that products of Schubert polynomials can be expanded (with a combinatorial formula) into
fundamental slide polynomials.

\begin{definition}
For $\alpha$ being a weak composition, \defin{fundamental slide polynomial} $\slideF_\alpha$
is defined as
\[
\slideF_\alpha(\xvec) \coloneqq \sum_{\substack{b \trianglerighteq a \\ \mathrm{flat}(b)\text{ refines }\mathrm{flat}(a)}} \xvec^b
\]
where $\mathrm{flat}(\beta)$ denotes the composition obtained by removing all 0s from $\beta$,
and $\trianglerighteq$ denotes \hyperref[partialOrderingsOnPartitions]{dominance order}.
\end{definition}


The fundamental slide polynomials have the \hyperref[gessel]{Gessel quasisymmetric functions}
as stable limit:
\[
 \lim_{m \to \infty} \slideF_{0^m\times \alpha}(\xvec) = \gessel_{\mathrm{flat}(\alpha)}(\xvec).
\]

\begin{example}
% Related Rust: sym-poly/multipoly/src/slide_polynomial.rs and
% sym-poly/multipoly/examples/slide_site_example.rs
For $\alpha=(1,0,2)$, the monomial slide polynomial is
\[
\slideM_{(1,0,2)}(\xvec)
=
x_1x_3^2+x_1x_2^2.
\]
The support can be seen directly from the weak-composition diagrams.  We record
the same information by listing the rows from top to bottom: row $3$, row $2$,
then row $1$.
\[
\begin{aligned}
b=(1,0,2)&:\quad
(\bullet\bullet,\ \emptyset,\ \bullet)
  &&\longmapsto x_1x_3^2,\\
b=(1,1,1)&:\quad
(\bullet,\ \bullet,\ \bullet)
  &&\longmapsto x_1x_2x_3,\\
b=(1,2,0)&:\quad
(\emptyset,\ \bullet\bullet,\ \bullet)
  &&\longmapsto x_1x_2^2.
\end{aligned}
\]
The first and last diagrams have flat composition $(1,2)$, so they contribute
to $\slideM_{(1,0,2)}$.  The middle diagram has flat composition $(1,1,1)$,
which refines $(1,2)$ but is not equal to it.
The fundamental slide polynomial permits the refinement $(1,1,1)$ of
$\mathrm{flat}(\alpha)=(1,2)$, so
\[
\slideF_{(1,0,2)}(\xvec)
=
x_1x_3^2+x_1x_2x_3+x_1x_2^2
=
\slideM_{(1,0,2)}(\xvec)+\slideM_{(1,1,1)}(\xvec).
\]
\end{example}

\subsection[fundamentalParticle]{Fundamental particle basis}

\begin{polydata}{fundamentalParticle}
  Name     & Fundamental particle basis \\
  Space    & All \\
  Basis    & True \\
  Rating   & 1 \\
  Bib      & Searles2019PolynomialBases \\
  Year     & 2020 \\
  Symbol   & $L_\alpha(\xvec)$ \\
  Keywords & divided-difference, 0-Hecke \\
  Category & Schur \\
  PositiveIn & monomial | HicksNiese2024x | scope=nonsymmetric monomials; proof=Thm. 3.9 and Cor. 3.11 \\
\end{polydata}

\name{Dominic Searles} introduced the \defin{fundamental particle basis}
\cite{Searles2019PolynomialBases}.  \name{Angela Hicks} and \name{Elizabeth Niese}
identify this basis with the fundamental Demazure atoms obtained by replacing
the usual Demazure atom operators with Hivert's quasisymmetric divided
difference operators \cite{HicksNiese2024x}.  In this form the basis has an
explicit monomial-positive formula, and the fundamental slide polynomials
expand positively in the fundamental particle basis.
\todo{Add a checked fundamental-particle expansion example, and compare it
with the fundamental slide expansion for a small weak composition.}


\subsection[slidePositive]{Slide positive families}

\hyperref[key]{Key polynomials} expand positively in the 
fundamental slide basis, \cite[Thm. 2.13]{AssafSearles2018}.
In \cite{ChoWilligenburg2022}, the authors determine for which $\alpha$,
the key polynomials $\key_\alpha$ expanded into slide polynomials are multiplicity free.

The \hyperref[pPartitionsFlagged]{flagged $(P,w)$-partition generating functions}
of \name{Sami Assaf} and \name{Nantel Bergeron} are slide-positive
\cite{AssafBergeron2019}.
In \cite{TewariWilsonZhang2022}, it is shown that certain polynomials similar to 
\hyperref[chromaticQuasisymmetric]{chromatic symmetric functions} are slide-positive.


See \cite{SmirnovTutubalina2021} for the notion of \emph{slide complexes}.

The \hyperref[kohnert]{Kohnert polynomial} page gives a diagrammatic model
which contains keys, Schubert polynomials, locks, and other Assaf--Searles
families as special cases.  For $K$-theoretic diagram models related to
Lascoux and glide phenomena, see the \hyperref[lascoux]{Lascoux polynomial}
page.


\section[forestPolynomials]{Forest polynomials}

\begin{polydata}{forestPolynomials}
  Name     & Forest polynomials \\
  Space    & All \\
  Basis    & True \\
  Rating   & 1 \\
  Bib      & NadeauTewari2023x \\
  Year     & 2023 \\
  Symbol   & $P_F(\xvec)$ \\
  Keywords & flagged-P-partitions, slide-positive, Schubert-positive \\
  Category & Schur \\
  PositiveIn & slideF | NadeauTewari2023x | proof=Thm. 1.3 \\
\end{polydata}

\name{Philippe Nadeau} and \name{Vasu Tewari}
\cite{NadeauTewari2023x} introduce \defin{forest polynomials}, indexed by
indexed forests.  These polynomials can be defined as generating functions for
labelings of forest posets with weak and strict inequalities along the two
types of edges.  This puts them close in spirit to flagged
\hyperref[pPartition]{P-partitions}.

Forest polynomials form a basis of the polynomial ring, and they expand
positively in the \hyperref[slideF]{fundamental slide basis}.  Conversely,
\hyperref[schubert]{Schubert polynomials} expand positively in the forest
polynomial basis.

\name{Philippe Nadeau}, \name{Hunter Spink}, and \name{Vasu Tewari}
\cite{NadeauSpinkTewari2024x} develop a quasisymmetric divided-difference
formalism in which forest polynomials play the role of Schubert polynomials.
Their trimming operators are indexed by forests and governed by the Thompson
monoid.  In this setting forest polynomials are characterized recursively by
the action of the trimming operators, and products of forest polynomials are
forest-positive.
\name{Annie Guo} and \name{Dora Woodruff} characterize when a Schubert
polynomial is itself a forest polynomial \cite{GuoWoodruff2026x}.  Their
criterion is that the indexing permutation avoids a specific set of six
patterns, giving another pattern-avoidance description of a special Schubert
class.


\section[lock]{Lock polynomials}

\begin{polydata}{lock}
  Name     & Lock polynomials \\
  Space    & All \\
  Basis    & True \\
  Rating   & 1 \\
  Bib      & AssafSearles2019 \\
  Year     & 2019 \\
  Symbol   & $\lock_\alpha(\xvec)$ \\
  Category & Schur \\
  Contains & schurS | AssafSearles2019 \\
  StableLimit & extSchur | AssafSearles2019 \\
  PositiveIn & slideF | AssafSearles2019 | proof=Thm. 6.11 \\
  PositiveIn & slideM | AssafSearles2019 | proof=Thm. 3.7 \\
\end{polydata}

Lock polynomials were introduced by \name{Sami Assaf} and \name{Dominic Searles} in \cite{AssafSearles2019}.
The Lock polynomials form a basis for the polynomial ring, 
and are indexed by weak compositions.
The combinatorial formula for these is
\[
\lock_\alpha(\xvec) \coloneqq \sum_{T \in LT(\alpha)} \xvec^T
\]
where the sum is over all \emph{lock tableaux}.

As an example (from \cite{Wang2020}), we have
\[
\lock_{(0,2,3)}(\xvec) = x_2^2 x_3^2 + x_1x_2x_3^3 + x_1^2x_3^3 + x_1x_2^2x_3^2 + 
x_1^2x_2x_3^2 + x_1^2x_2^2x_3 + x_1^2 x_2^3
\]

Whenever the non-zero parts of $\alpha$ are weakly decreasing,
we have that the lock polynomial coincides with a \hyperref[key]{key polynomial};
$\lock_\alpha(\xvec)=\key_\alpha(\xvec)$, see \cite[Thm. 6.12]{AssafSearles2019}.

There is a \hyperref[crystals]{crystal structure} on lock polynomials, 
explored by \name{George Wang} in \cite{Wang2020}.
This crystal structure embeds naturally into \hyperref[key]{Demazure crystals}.
