\metatitle{Chromatic quasisymmetric functions}
\metadescription{An introduction to chromatic quasisymmetric functions, their definition via proper colorings of graphs, properties such as symmetry, reciprocity, and power-sum positivity, and connections to LLT polynomials and Hessenberg varieties.}  

\section[chromaticQuasisymmetric]{Chromatic quasisymmetric functions}

\begin{polydata}{chromaticQuasisymmetric}
  Name   & Chromatic quasisymmetric functions \\
  Space  & Sym \\
  Basis  & False \\
  Rating & 7 \\
  Bib    & ShareshianWachs2012 \\
  Year   & 2012 \\
  PositiveIn & gessel | ShareshianWachs2012 | semiring=NN[q] \\
  PositiveIn & qPsi | AlexanderssonSulzgruber2019 | scope=after applying omega \\
  PositiveIn & powerSum | AlexanderssonSulzgruber2019,Ellzey2017,Athanasiadis2015 | scope=after applying omega, when symmetric \\
  PositiveIn & schurS | Gasharov1996,ShareshianWachs2012 | scope=(3+1)-free incomparability graphs at q=1 and unit interval graphs with q parameter \\
  PositiveIn & elementaryE | Hikita2024x | semiring=NN; scope=(3+1)-free incomparability graphs at q=1 \\
  Contains & completeMultipartite | CrewSpirkl2020x | scope=complete multipartite graphs at q=1 \\
  SignedIn & completeMultipartite | CrewSpirkl2020x | scope=chromatic symmetric functions at q=1; sign_rule=maximal stable partitions; proof=Cor. 3 \\
\end{polydata}

\name{G. Arunkumar} relates Stanley's chromatic symmetric functions to
Borcherds algebras \cite{Arunkumar2021}.  Another Hopf-algebraic viewpoint is
given by \name{Jean-Christophe Aval}, \name{Nantel Bergeron}, and
\name{John Machacek}, who introduce invariants of permutations, orders, and
graphs connected with the nabla operator \cite{AvalBergeronMachacek2020}.
\name{Jacob A. White} defines chromatic quasisymmetric class functions for
digraphs and, more generally, for combinatorial Hopf monoids
\cite{White2021,White2025}; these variants incorporate group actions into the
chromatic quasisymmetric setting.




\subsection[chromaticQuasisymmetricHistory]{Historical overview}

\name{Richard Stanley} introduced the \defin{chromatic symmetric function} of a graph $G$ in \cite{Stanley1995}.
This definition was later refined by \name{John Shareshian} 
and \name{Michelle Wachs} in \cite{ShareshianWachs2012} and \cite{ShareshianWachs2016},
where the graph $G$ is labeled and a parameter $q$ is introduced.

In Spring 2016, \name{Brittney Ellzey} \cite{Ellzey2017} presented a generalization where labeled graphs are replaced with \emph{oriented graphs}.
This new model is equivalent to the Shareshian--Wachs model when the orientation of the edges has no directed cycle.
See also  \name{Greta Panova}, \name{Per Alexandersson} \cite{AlexanderssonPanova2018}, who considered the same model from a different perspective.

The main open problem in this area concerns the conjectured positive 
expansion of chromatic symmetric functions in the elementary symmetric function basis,
and later the refined conjecture by \name{John Shareshian} 
and \name{Michelle Wachs} in \cite{ShareshianWachs2012,ShareshianWachs2016}.
These conjectures are referred to 
the \hyperref[chromaticQuasisymmetricElementaryExpansion]{Stanley--Stembridge conjecture and Shareshian--Wachs conjecture}, respectively.
Shareshian and Wachs also conjectured that $\omega \chrom_G(\xvec;q)$ is 
\hyperref[dotAction]{related to 
the equivariant cohomology of regular semisimple Hessenberg varieties}.



In \cite{Hwang2022x}, \name[B.-H. Hwang]{Byung-Hak Hwang} introduces an 
additional refinement on the chromatic quasisymmetric functions.
He defines a \enquote{flip} operator on colorings and show that connected 
components under such flips
are \hyperref[schurPositivity]{Schur-positive}. The ascent parameter introduced by Shareshian--Wachs
is preserved by such flips.
It is conjectured that each such component is positive in the elementary 
symmetric function basis.
See also \cite{Hwang2023x} for a connection with non-commutative functions.
A proof for Schur-positivity of the flip-components is also given in \cite{BlasiakErikssonPylyavskyySiegl2022x}.


For a representation-theoretical perspective, see \cite{Skandera2020},
where connections with immanents and characters is studied.

\name{Yusaku Nishimura} proves that the second Kneser chromatic function
$X_{K_{\setN,2}}(G)$ is a complete invariant for trees \cite{Nishimura2026}.
This fits into the Kneser chromatic functions of \name{Takayuki Hasebe},
\name{Hiroyuki Miezaki}, \name{Akihiro Munemasa}, \name{Yusaku Nishimura},
\name{Yusuke Sakuma}, and \name{Shuhei Tsujie}, where
$X_{K_{\setN,1}}(G)$ is Stanley's chromatic symmetric function and the whole
family is a complete invariant for finite graphs.

\name{Logan Crew}, \name{Evan Haithcock}, \name{Josephine Reynes}, and
\name{Sophie Spirkl} introduce a chromatic multisymmetric function for graphs
whose vertex set is partitioned into parts \cite{CrewHaithcockReynesSpirkl2024}.
They use homogeneous sets and homogeneous pairs to turn known relations for
ordinary chromatic symmetric functions into new ones, including a
generalization of Guay-Paquet's reduction of the Stanley--Stembridge
conjecture to unit interval graphs.

\name{Farid Aliniaeifard}, \name{Victor Wang}, and
\name{Stephanie van Willigenburg} define a centered chromatic symmetric
function $y_{G:v}$ using a quotient of noncommutative symmetric functions
\cite{AliniaeifardWangWilligenburg2024}.  This viewpoint gives new
$e$-positive unit interval graph families, operations preserving the stronger
centered positivity property, and noncommutative refinements of sink-counting
theorems for acyclic orientations.


\subsection[chromaticQuasisymmetricDefinition]{Definition}

Let $G$ be an \emph{oriented graph} with $n$ vertices. 
Given a vertex-coloring $\kappa:G\to \setP$,
let $\asc(\kappa)$ denote the number of directed edges $(u,v)$ in $G$ such that $\kappa(u) \lt \kappa(v)$.

The \defin{chromatic quasisymmetric function} of $G$ is defined as 
\[
\chrom_G(\xvec;q) \coloneqq  
\sum_{\substack{\kappa : G\to \setP \\ \kappa \text{ proper}} } x_{\kappa(1)}\dotsm x_{\kappa(n)} q^{\asc(\kappa)},
\]
where the sum runs over all proper colorings of the graph $G$. 
A coloring is proper if there are no edges whose endpoints receive the same color.
The function $\chrom_G(\xvec;q)$ is quasi-symmetric in $\xvec$.

From the definition, it is clear that
\[
\chrom_G(1^k;1)
\]
is the number of ways to color $G$ with $k$ colors, 
so the classical \defin{chromatic polynomial} $\chi_G(k)$ is given by the specialization $\chrom_G(1^k;1)$.
See \cite{SaganVatter2021} for a nice article with bijective proofs of properties of $\chi_G(k)$.
\name{Zhiyun Cheng}, \name{Ziyi Lei}, \name{Yitian Wang}, and
\name{Yanguo Zhang} construct a cohomology theory for the signed chromatic
polynomial of a signed graph \cite{ChengLeiWangZhang2022}.  Its graded Euler
characteristic is the signed chromatic polynomial, and its long exact sequence
categorifies signed deletion--contraction.
\name{Jean-Christophe Aval} and \name{Raquel Melgar} introduce chromatic
quasisymmetric functions for signed graphs \cite{AvalMelgar2025x}.  Their
construction is guided by Zaslavsky's signed-graph hyperplane arrangements and
leads to an algebra of signed quasisymmetric functions.


Note that every labeling induces an orientation, by orienting edges in the direction of largest label.
This labeled model is what was introduced in \cite{ShareshianWachs2012,ShareshianWachs2016}.



\subsection[chromaticQuasisymmetricClawFree]{Claw-free graphs}

If $G$ is a claw-free graph, \name{Vesselin Gasharov} (see \cite{Stanley1998}) conjectured that $\chrom_G(\xvec;1)$ is \hyperref[schurPositivity]{Schur-positive}.
There is no known $q$-analog of this conjecture.

Some progress in this direction has been made in \cite{ShelburneWilligenburg2024x},
using a type of involution inspired by the \hyperref[gesselSlinkyRule]{slinky rule}.
They prove Schur positivity for the family of graphs called \defin{generalized nets}.

\name{Randy Davila}, \name{Houston Schuerger}, and \name{Ben Small}
characterize claw-free graphs using zero forcing invariants
\cite{DavilaSchuergerSmall2024x}.  They prove that $G$ is claw-free if and
only if every induced subgraph $H$ satisfies $Z(H)=Z_+(H)$.

\name{Ethan Y. H. Li}, \name{Grace M. X. Li}, \name{Arthur L. B. Yang},
and \name{Zhong-Xue Zhang} introduce a strengthening of the nice property,
called the \defin{strongly nice property} \cite{LiLiYangZhang2024x}.
They prove that Schur positivity implies strong niceness, and that the
graphs whose induced subgraphs are all strongly nice are precisely the
claw-free graphs.  Thus strong niceness gives a necessary obstruction
which is well aligned with Gasharov's conjecture.

Distributive lattices give a separate family of negative tests when a poset
is viewed through its incomparability graph.  \name{Grace M. X. Li},
\name{Dun Qiu}, \name{Arthur L. B. Yang}, and \name{Zhong-Xue Zhang}
construct distributive lattices which are not nice and hence not
Schur-positive, and also distributive lattices which are nice but still not
Schur-positive \cite{LiQiuYangZhang2024x}.  Later,
\name{David G. L. Wang} and \name{K. Zhang} prove that the product lattices
$m \times 2$ and $m \times 3$ are not Schur-positive for $m \geq 8$, and that
$m \times 3$ is not strongly nice for $m \geq 44$ \cite{WangZhang2025x}.

\subsection[chromaticQuasisymmetricUnitIntervalGraph]{Unit interval graphs}

The following definition using area sequences is from \cite{AlexanderssonPanova2018}.

A \defin{circular unit arc digraph} (or circular Dyck path) 
$\Gamma_\avec$ is a directed graph with vertex set $[n]$ and edges
\begin{align}
(i-a_i) \to i, \; (i-a_i + 1) \to i,  \; (i-a_i + 2) \to i, \; \dotsc \; , (i-1) \to i
%   i \to i+1, \; i \to i+2,\; \dotsc, i \to i+a_i
\end{align}
for all $i = 1,\dotsc,n$, where indices are taken modulo $n$, 
and the integers $\avec=(a_1,\dotsc,a_n)$ satisfy

\begin{itemize}
\item $0 \leq a_i \leq n-1 $ for $1 \leq i \leq n$, 
\item $a_{i+1} \leq a_{i} + 1$ for $1 \leq i \leq n$, (index mod $n$).
\end{itemize}

Whenever $a_1=0$, $\Gamma_\avec$ is a \defin{unit interval graph}.
The sequence $\avec = (a_1,a_2,\dotsc,a_n)$ is called the \defin{area sequence} of the graph.

There are Catalan $\frac{1}{n+1}\binom{2n}{n}$ many unit interval graphs on $n$ vertices, 
and a total of 
\[
(n+2)\binom{2n-1}{n-1} - 2^{2n-1}
\]
circular unit arc digraph on $n$ vertices, see \cite{AlexanderssonLinussonPotka2019} for a proof.
This sequence appears as \oeis{A194460} in the OEIS.


\begin{example}
The area sequences $(0,1,2,1,1)$ and $(3,2,1,2,2)$ can be 
presented as \emph{Dyck diagrams}, or in the second circular case,
a \defin{circular Dyck diagram}:

\begin{figure}
\begin{ytableau}
\square & \square & \square &  & 5 \\
\square & \square &   & 4 \\
  &   &  3 \\
  &  2 \\
1
\end{ytableau}
\begin{ytableau}
\none & \none & \none & \none & \square & \square & & & 5 \\
\none & \none & \none & \square & \square & & & 4 \\
\none & \none & \square & \square & \square & & 3 \\
\none & \square & \square & & & 2 \\
\square &  & & & 1
\end{ytableau}
\end{figure}

The labeled boxes represent vertices, the empty boxes are edges 
(connecting the vertex in the same column with the vertex in the same row),
and the squares represent non-edges. The first graph has 5 edges, while the second has 9 edges.
Note that circular Dyck diagrams \enquote{wrap around} at top and bottom.
Edges are directed upwards (or rightwards) in these diagrams.
\end{example}


The \defin{Abelian unit interval graphs} on vertices $\{1,2,\dotsc,n\}$
are the unit interval graphs where there is no vertex $j$ such that both $(1,j)$ and $(j,n)$ are non-edges.
The number of Dyck paths of size $n$ with an Abelian unit interval graph is $2^{n-1}$.



\section[chromaticQuasisymmetricProperties]{Properties of chromatic quasisymmetric functions}


\subsection[chromaticQuasisymmetricSymmetric]{Symmetry}

Whenever $G$ is a unit arc digraph as above, $\chrom_G(\xvec;q)$ is symmetric 
(for labeled graphs, \cite{ShareshianWachs2012,ShareshianWachs2016} and for oriented graphs, \cite{Ellzey2017,AlexanderssonPanova2018}).
Furthermore, the coefficients in $\setZ[q]$ in the monomial basis are palindromic.
To be precise,
\[
\chrom_\avec(\xvec;q)=\sum_\lambda c_\lambda(q) \monomial_\lambda(\xvec) \qquad \Longrightarrow \qquad
c_\lambda(q) = q^{|\avec|}c_\lambda(1/q),
\]
where $|\avec|$ denotes the sum $a_1+a_2+\dotsb + a_n$.

\name{Maria Gillespie}, \name{Joseph Pappe}, and \name{Kyle Salois} study when
a chromatic quasisymmetric function is symmetric \cite{GillespiePappeSalois2024x}.
They show that symmetry of a product of quasisymmetric functions forces each
factor to be symmetric, reduce the problem to connected graphs, and prove that
directed trees other than directed paths have nonsymmetric chromatic
quasisymmetric functions.
\name{Laura Colmenarejo} and \name{Ian Klein} introduce total chromatic
quasisymmetric functions of graphs \cite{ColmenarejoKlein2026x}.  One variant
normalizes over all vertex labelings, while another is built from acyclic
orientations; the paper compares their monomial-quasisymmetric expansions and
gives explicit formulas for star graphs.


\subsection[chromaticQuasisymmetricReciprocity]{Reciprocity}

Let $\chi_G(k)$ be the chromatic polynomial associated with the graph $G$ on the vertex set $[n]$.
\name{Olivier Bernardi} and \name{Philippe Nadeau} \cite{BernardiNadeau2020} gave the following interpretation of
the specialization $(-1)^{n-i}\chi^{(i)}_G(-j)$.
\begin{theorem}
Let $i,j \geq 0$. Then $(-1)^{n-i}\chi^{(i)}_G(-j)$ counts the number of tuples
\[
\left( (V_1, \theta_1),\dotsc,(V_{i+j}, \theta_{i+j}) \right)
\]
such that 
\begin{itemize}
\item $V_1,\dotsc,V_{i+j}$ is a set-partition of the vertices of $G$,
\item $\theta_\ell$ is an acyclic orientation of the graph induced by $V_\ell$,
\item for all $\ell \in [i]$, $V_\ell$ is non-empty and $\theta$ has a unique source, given by $\min V_\ell$.
\end{itemize}
\end{theorem}





\subsection[chromaticQuasisymmetricLLT]{Relationship with LLT polynomials}

Carlson--Mellit \cite[Prop.3.4]{CarlssonMellit2017} proved the following \hyperref[plethysm]{plethystic} relationship between 
chromatic quasisymmetric functions and \hyperref[LLT]{unicellular LLT polynomials} indexed by unit interval graphs 
on $n$ vertices:
\[
(q-1)^{-n} \LLT_\avec[\xvec(q-1);q] =  \chrom_\avec(\xvec;q).
\]
This relationship is not always true in the circular case, that is, for area sequences where $a_1 \gt 0$.
The relationship is reminiscent of the technique 
introduced by \name{Jim Haglund}, \name{Mark Haiman} 
and \name{Nicholas Loehr} \cite{HaglundHaimanLoehr2005}.

This plethystic relationship is generalized in \cite[Thm. 6.2]{NovelliThibon2019},
where a non-commutative lift is given. \name[J.-C. Novelli]{Jean-Christophe Novelli} and \name[J.-Y. Thibon]{Jean-Yves Thibon} also introduce 
a non-commutative lift of the unicellular LLT polynomials.

\name{Michele D'Adderio}, \name{Roberto Riccardi}, and
\name{Viola Siconolfi} extend this picture from Dyck graphs to interval
graphs \cite{DAdderioRiccardiSiconolfi2023x}.  They introduce
quasisymmetric functions attached to increasing spanning forests and use them
to give a forest expansion for chromatic quasisymmetric functions of interval
graphs \cite[Thm.~7.2]{DAdderioRiccardiSiconolfi2023x}.  They also define
LLT quasisymmetric functions for ordered simple graphs, recovering the
unicellular LLT symmetric functions for Dyck graphs
\cite[Thm.~7.7]{DAdderioRiccardiSiconolfi2023x}, and prove an interval-graph
generalization of the Carlsson--Mellit plethystic identity
\cite[Thm.~10.1]{DAdderioRiccardiSiconolfi2023x}.


\subsection[chromaticQuasisymmetricGesselExpansion]{Fundamental quasisymmetric positivity}

It is easy to show that
\[
\chrom_G(\xvec;q) = \sum_{\theta \in AO(G)} q^{\asc(\theta)} K_{P(\theta)}(\xvec)
\]
where $K_{P(\theta)}(\xvec)$ is the generating function 
of \emph{strict} \hyperref[pPartition]{order-preserving maps} 
from the poset $P(\theta)$ given by the transitive closure of the directed edges in $\theta$,
to the set of positive integers.

From this, it follows that $\chrom_G(\xvec;q)$ is 
positive in the \hyperref[gessel]{fundamental quasisymmetric basis}.

\name{Vasu Tewari}, \name{Andrew Wilson}, and \name{Philip Zhang} study a
nonsymmetric refinement for Dyck graphs and prove positivity in the
\hyperref[slideF]{fundamental slide basis} \cite{TewariWilsonZhang2022}.


\subsection[chromaticQuasisymmetricPowerSumExpansion]{Power-sum positivity}

In \cite{AlexanderssonSulzgruber2019}, 
it is proved that $\omega \chrom_G(\xvec;q)$ is always $\qPsi$-positive, 
that is, positive in a \hyperref[qPsi]{quasisymmetric powersum} basis.
Consequently, $\omega \chrom_G(\xvec;q)$ 
is $\powerSum$-positive whenever $\chrom_G(\xvec;q)$ is symmetric. 
This was proved earlier in \cite{Ellzey2017}, and in the case of unit interval graphs $G$ in \cite{Athanasiadis2015}.
A combinatorial formula in the unit-interval case was conjectured earlier by Shareshian--Wachs.


We have the following expansion, proved in \cite{AlexanderssonSulzgruber2019}:
\begin{theorem}[Alexandersson, Sulzgruber (2019)]
Let $G$ be a graph on $n$ vertices.
The quasisymmetric power-sum expansion of $\omega \chrom_G(\xvec;q)$ is given by
\[
\omega \chrom_G(\xvec;q) = \sum_{\theta \in AO(G)} q^{\asc(\theta)} 
\sum_{f \in \opsurj^{\ast}(\theta)} \frac{\qPsi_{\type(f)}(\xvec)}{z_{\type(f)}}
\]
where the first sum ranges over acyclic orientations of $G$.
The set $\opsurj_{\alpha}^{\ast}(\theta)$
consists of order-preserving surjections $f$ from $\theta$ to some $[k]$,
such that $\theta$ restricted to the vertices $f^{-1}(j)$ has a unique sink.
The type of $f$ is the composition $\alpha$ given by $\alpha_j \coloneqq |f^{-1}(j)|$.
\end{theorem}
In particular, for $q=1$, we have that $\chrom_G(\xvec)  = \chrom_G(\xvec;1)$ is symmetric.
Thus, the formula above gives that
\begin{align}
\omega \chrom_G(\xvec) &= \sum_{\theta \in AO(G)} 
\sum_{f \in \opsurj^{\ast}(\theta)}
\frac{\powerSum_{\type(f)}(\xvec)}{z_{\type(f)}}.
\end{align}
This can further be simplified,
to give 
\begin{align}
\omega \chrom_G(\xvec) &=\sum_{\theta \in AO(G)} \powerSum_{\type(\theta)}(\xvec),
\end{align}
where $\type(\theta)$ is given by the \emph{source-components}
of $\theta$, see \cite{BernardiNadeau2020}.
This formula is very similar to the stronger result by \name[C. A. Athanasiadis]{Christos Athanasiadis} \cite{Athanasiadis2015},
which treats the case with general $q$.


%
% For a given $\theta$, we define an ordered set partition $\pi(\theta)$
% as follows (compare with the $\elementaryE$-expansion
% of \hyperref[unicellularLLTEExpansion]{unicellular LLT polynomials}).
% Each sink $s$ of $\theta$ lies in a unique part of $\pi(\theta)$.
% Furthermore, vertex $j$ is in the same part as $s$,
% if $s$ is the largest sink $j$ can reach via a directed path.
% Hence, each part of $\pi(\theta)$ induces an
% orientation with a unique sink --- the largest element in the part.
%
% Recall also that $z_\alpha = \prod_j m_j! j^{m_j}$,
% where there are $m_j$ parts equal to $j\geq  1$ in $\alpha$.
%
% &= \sum_{\theta \in AO(G)} \powerSum_{\type(\theta)}(\xvec)
% where the type of $\theta$ is the partition given by the part sizes of $\pi(\theta)$.


\subsection[chromaticQuasisymmetricSchurExpansion]{Schur positivity}

In \cite{Gasharov1996}, \name[Gasharov]{Vesselin Gasharov} proved that the $\chrom_G(\xvec;1)$ 
are \hyperref[schurPositivity]{Schur-positive} with an explicit expansion,
whenever $G$ is the incomparability graph of a $3+1$-free poset.
In particular, this implies the unit-interval case, as those are given as the incomparability graphs
of $3+1$ and $2+2$-free posets. Gasharov's proof uses a kind of sign-reversing involution.

Gasharov's result was later extended to the Shareshian--Wachs setting (with a $q$-parameter),
by using a different involution.
That is, $\chrom_G(\xvec;q)$ is \hyperref[schurPositivity]{Schur-positive} for unit-interval graphs $G$,
see \cite[Thm. 6.3]{ShareshianWachs2012}.
\name{Dongkwan Kim} and \name{Pavlo Pylyavskyy} give a
Robinson--Schensted correspondence for unit interval orders which refines the
Schur expansion through dual equivalence graphs \cite{KimPylyavskyy2021}.

\begin{example}
We have the following Schur expansions of chromatic quasisymmetric functions, indexed by area sequences:
\begin{align*}
\chrom_{000}(x;q)&=\schurS_{111}+2 \schurS_{21}+\schurS_{3} \\
\chrom_{001}(x;q)&=(1+q) \schurS_{111}+(1+q) \schurS_{21} \\
\chrom_{011}(x;q)&=(1+2 q+q^2) \schurS_{111}+q \schurS_{210} \\
\chrom_{111}(x;q)&=(3 q+3 q^2) \schurS_{111} \\
\chrom_{012}(x;q)&=(1+2 q+2 q^2+q^3) \schurS_{111} \\
\chrom_{112}(x;q)&=(q+4 q^2+q^3) \schurS_{111} \\
\chrom_{122}(x;q)&=(3 q^2+3 q^3) \schurS_{111} \\
\chrom_{222}(x;q)&=6 q^3 \schurS_{111} 
\end{align*}
\end{example}


\begin{problem}
Prove that $\chrom_G(\xvec;q)$ is \hyperref[schurPositivity]{Schur-positive} (with coefficients in $\setN[q]$)
in the case of \emph{circular} unit arc digraphs.
This conjecture has been verified for all circular unit arc digraphs up to 10 vertices.

Find a bijective proof of Schur-positivity (using \hyperref[schurPositivityCrystals]{crystals} or
a version of \hyperref[schurPositivityRSK]{RSK}).
Gasharov's result provides a hint on how a crystal graph structure might be defined.

A step in the direction of a crystal structure is taken in \cite{Ehrhard2022x},
but the \enquote{crystals} described there are not crystals in the usual sense.
\end{problem}


In \cite{WangWang2020}, the authors give a (signed) formula 
for the Schur coefficients of the chromatic symmetric function of any graph.
They do this by inverting the matrix of Kostka coefficients.
The approach is reminiscent of how one can turn a fundamental quasisymmetric expansion
into a Schur expansion, using the \hyperref[gesselSlinkyRule]{slinky rule}.
The authors also characterize all Schur positive complete tripartite graphs.



\subsection[chromaticQuasisymmetricElementaryExpansion]{Elementary symmetric expansion}

The Stanley--Stembridge conjecture (1993) is a long-standing open problem in 
algebraic combinatorics, first formulated in \cite{StanleyStembridge1993} 
and \cite{Stanley1995} (for $q=1$). A refinement using the general $q$-parameter was presented 
in \cite[Conj. 4.9]{ShareshianWachs2012}. 
Positivity in the elementary symmetric function basis (for $q=1$) was recently proved 
by \name{Tatsuyuki Hikita}.
For a survey on expansions in the elementary symmetric function basis, see \hyperref[chromaticEexpansion]{this page}.

The \hyperref[def:hookFilter]{hook filter} on the Gessel fundamental basis is
useful for extracting hook-shaped Schur coefficients from chromatic symmetric
functions; it is one of the tools already present in Stanley's original
development \cite{Stanley1995}.


\subsection[chromaticDeletionContraction]{Deletion contraction property}

There is an extension to \hyperref[extendedChromatic]{extended chromatic symmetric functions}
which admits a deletion-contraction recurrence.

\name{Rosa Orellana} and \name{Geoffrey Scott} prove a triangle-deletion
recurrence and use it to construct and compare graphs with equal chromatic
symmetric functions \cite{OrellanaScott2014}.


\subsection[rookTheory]{Connection with rook placements}

Chromatic quasisymmetric functions of unit interval graphs have a strong connection with rook placements.
Acyclic orientations and rook placements are in bijection, see for example \cite{AlexanderssonPanova2018}.
This connection is expanded upon in \cite{ColmenarejoMoralesPanova2023}.
In particular, some new results in the Abelian case using $q$-hit numbers are given.

\subsection[alphaChromaticSymmetric]{Alpha-chromatic symmetric functions}

\name{Jim Haglund}, \name{Jaeseong Oh}, and \name{Meesue Yoo}
introduce \defin{$\alpha$-chromatic symmetric functions}, which add a
real parameter $\alpha$ to the chromatic quasisymmetric functions of
natural unit interval graphs \cite{HaglundOhYoo2025}.  These functions
interpolate between chromatic symmetric functions and unicellular
\hyperref[unicellularLLT]{LLT polynomials}: the coefficient of the first
$q$-falling-factorial term recovers the chromatic symmetric function, while
the large-$\alpha$ limit recovers the Schur coefficients of the
unicellular LLT polynomial up to a $q$-factorial factor.

Their main results include a positive monomial formula in an
$\alpha$-binomial basis \cite[Thm.~3.1]{HaglundOhYoo2025}, an expansion
into a falling-factorial basis whose coefficients are sums of ordinary
chromatic symmetric functions \cite[Thm.~4.4]{HaglundOhYoo2025}, and a
rook-theoretic solution of the Garsia--Remmel $q$-hit problem for Ferrers
boards associated with Dyck paths \cite[Cor.~6.3]{HaglundOhYoo2025}.


\section[treesChromaticQuasisymmetric]{Chromatic symmetric functions of trees}


\name{Richard Stanley} conjectures \cite{Stanley1995} that the chromatic symmetric function of a tree, 
uniquely determines the tree. In \cite{WangYubZhang2023x} it is shown that trees with exactly 
two vertices of degree greater than 2, are distinguished by their chromatic symmetric functions.

For chromatic symmetric functions (CSF) for trees in the star basis, see \cite{PrietoMierOrellanaZamora2023,GonzalezOrellanaTomba2024x}.
It is shown that trees of diameter less than 5 can be reconstructed from their CSF.
Caterpillar graphs are known to be reconstructed from their CSF, \cite{LoeblSereni2019}.
Certain $q$-caterpillar trees can be reconstructed from their CSF, \cite{ArunkumarNarayananBVSawant2023x}.

\name{Nima Amoei Mobaraki}, \name{Yasaman Gerivani}, and
\name{Sina Ghasemi Nezhad} develop a framework for calculating chromatic
symmetric functions in forest bases \cite{MobarakiGerivaniNezhad2024x}.
Their method recovers the equivalence of the chromatic symmetric function and
the $U$-polynomial on forests, and gives formulas for extracting monomial
coefficients from induced subgraphs.

In \cite[Thm. 1.3]{HasebeTsujie2017}, it is proved that \emph{ordered trees} are distinguished by
their order quasisymmetric function (where the rooted tree is viewed as a poset, colors are increasing away from the root).
Both the weakly increasing, or strictly increasing version of the quasisymmetric function does the job.
The work by \name{Takahiro Hasebe} and \name{Shuhei Tsujie} is continued in \cite{Zhou2020x},
where one is interested in reconstructing the tree from the quasisymmetric function.

It is still an open problem (posed by Hasebe and Tsujie) to prove that
the \hyperref[pPartition]{P-partition quasisymmetric function} distinguishes
posets whose Hasse diagrams are trees.

In \cite{AlexanderssonSulzgruber2019}, we conjecture that in the family of directed trees,
the chromatic quasisymmetric function uniquely determines the directed tree.
This conjecture would follow from the conjecture by Hasebe and Tsujie.
See also \cite{FoleyKazdanKrollAlbergaMelnykTenenbaum2021x}.


There is a variant of Stanley's CSF defined for \emph{rooted trees},
and this rooted version is shown to distinguish non-isomorphic trees, see \cite{LoehrWarrington2024}.


The CSF associated with trees are in general not \hyperref[schurPositivity]{Schur positive},
see the counterexamples presented in \cite{EmmanuellaSandratraRambeloson2020}.
In fact, no tree on 20 vertices with maximal degree 10 is \hyperref[schurPositivity]{Schur positive}.


See also \cite{PrietoMartinWagnerZamora2024x} for various results on polynomial invariants of trees.


\section[chromaticQuasisymmetricGeneralGraphs]{Other families of graphs}

By writing the CSF in different bases, one can obtain various classical statistics of graphs,
see \cite{ChanCrew2023}. For coefficients in the elementary basis in relation to sinks, see \cite{CrewZhang2022x}.


The chromatic symmetric functions for certain \enquote{spider graphs} 
are shown to be \hyperref[schurPositivity]{Schur positive} in \cite{ThibonWang2023x}.
The approach uses non-commutative ideas.

\name{Esme Bajo}, \name{Matthias Beck}, and
\name{Andrés R. Vindas-Meléndez} define $q$-chromatic polynomials by
weighting proper colorings with $q^{\sum_v \lambda_v c(v)}$
\cite{BajoBeckVindasMelendez2024x}.  These polynomials specialize to the
ordinary chromatic polynomial at $q=1$ and, for $\lambda=(1,\dotsc,1)$, to a
principal specialization of Stanley's chromatic symmetric function.  Their
results connect this invariant with $q$-Ehrhart polynomials, order
polytopes, and strengthened tree-distinguishing questions.



\section[chromaticQuasisymmetricHessenberg]{Connection with Hessenberg varieties}

\label{chromaticQuasisymmetricMomentGraph}
In \cite[Conj. 5.3]{ShareshianWachs2012}, a connection with 
cohomology of regular semisimple Hessenberg varieties was conjectures.
This conjecture was resolved in 2015, where \name{Patrick Brosnan} 
and \name{Timothy Chow} proved that for unit-interval graphs $\avec$, 
the symmetric function $\omega \chrom_\avec(\xvec;q)$ 
is the \hyperref[frobeniusCharacteristic]{Frobenius characteristic} of 
a symmetric group action on the cohomology of regular semisimple Hessenberg varieties.
This was also proved independently by Guay-Paquet \cite{GuayPaquet2016x}.

For more details on this symmetric group action, see the \hyperref[dotAction]{dot action page}.

\name{Lucas Gagnon} gives a finite-group analogue of this result:
inducing the trivial character from certain unipotent subgroups of
$\GL_n(\mathbb{F}_q)$ realizes the chromatic quasisymmetric function, up to the
expected scalar factor \cite[Thm.~3.1]{Gagnon2024}.  The same paper relates
these character values to point counts of nilpotent Hessenberg varieties over
finite fields, and realizes vertical-strip LLT polynomials as unipotent
constituents of induced characters \cite[Thm.~5.1]{Gagnon2024}.

\name{Alex Abreu}, \name{Antonio Nigro}, and \name{Samrith Ram} give a
finite-field point-counting formula for arbitrary Hessenberg varieties
\cite{AbreuNigroRam2024x}.  If $T$ is a linear operator on
$\mathbb{F}_q^n$, then the number of $\mathbb{F}_q$-rational points of the
Hessenberg variety $H(m,T)$ is
\[
|H(m,T)|
=
\left\langle F_T(\xvec),\omega\chrom_{G(m)}(\xvec;q)\right\rangle,
\]
where $F_T(\xvec)$ is an invariant-flag generating function expressible using
modified Hall--Littlewood polynomials \cite[Thm.~1.1]{AbreuNigroRam2024x}.
For triangulable $T$, this also gives Poincare polynomials of the corresponding
complex Hessenberg varieties \cite[Cor.~3.8]{AbreuNigroRam2024x}.

\name{Alex Abreu} and \name{Antonio Nigro} relate characters of
Kazhdan--Lusztig elements of the Hecke algebra, including chromatic symmetric
functions of indifference graphs, to intersection cohomology of parabolic
Lusztig varieties \cite{AbreuNigro2025Parabolic}.  The resulting
decomposition through partial flag varieties points toward recursive
approaches to Haiman's positivity conjecture and the Stanley--Stembridge
conjecture.

\name{Young-Hoon Kiem} and \name{Donggun Lee} introduce generalized
Hessenberg varieties and formulate a weighted-graph version of the
Shareshian--Wachs conjecture \cite{KiemLee2024}.  They prove the
generalized conjecture, so the corresponding weighted chromatic
quasisymmetric functions have Schur coefficients which are unimodal
polynomials with nonnegative coefficients \cite[Thm.~4.6 and
Cor.~4.8]{KiemLee2024}.  They also propose a generalized
Stanley--Stembridge question for elementary positivity in this weighted
setting.


\section[chromaticQuasisymmetricMisc]{Other results related to chromatic symmetric functions}

In \cite{LinPierson2025x}, the authors study a different multiplicative 
structure (Hopf algebra) on chromatic symmetric functions and show that graph
complement (of unweighted, triangle-free graphs) corresponds naturally to a Hopf algebra operation.
