\metatitle{Miscellaneous cyclic sieving}
\metadescription{A list of miscellaneous cyclic sieving phenomena}
\metakeywords{cyclic sieving,rowmotion,root posets,minuscule posets,
frieze patterns,lattice polytopes,Hamming codes}

See the \hyperref[cyclicSieving]{cyclic sieving phenomenon}
page for the definition and related theorems.

\section[cspMisc]{Miscellaneous}


\subsection[cspListPosetsRoot]{Root posets}


Let $\Phi$ be a \hyperref[rootSystem]{root system} and $\Phi^+$ its poset of positive roots.
Armstrong--Stump--Thomas \cite{ArmstrongStumpThomas2013} studied the action of
the \hyperref[rowmotion]{rowmotion} operator, also called the
\enquote{Fon-Der-Flaass/Panyushev map}, on the \hyperref[antichain]{antichains}
of $\Phi^+$.
Addressing conjectures of Panyushev and Bessis--Reiner, they showed that this action is
in equivariant bijection with the Kreweras complementation action on
the $\Phi$-noncrossing partition lattice.
Moreover, they showed that the $q$-$\Phi$-Catalan polynomial is a cyclic sieving polynomial
for the action of rowmotion on the antichains of $\Phi^+$ (equivalently, for
the Kreweras complement acting on $\Phi$-noncrossing partitions).
Note that the Type $A$ case of this CSP is equivalent to
the \hyperref[cspNonCrossingMatchingsI]{noncrossing matchings} CSP.


Rowmotion can be realized as a certain composition of toggles on
\hyperref[orderIdeal]{order ideals}.
By making these toggles piecewise-linear, one obtains a piecewise-linear action
of rowmotion on height $m$ \hyperref[pPartition]{$P$-partitions} of a poset $P$.
Hopkins (see \cite{Hopkins2024}) conjectured a CSP for the action
of piecewise-linear rowmotion on these $P$-partitions when $P$ is a
root poset \enquote{of coincidental type}, where the sieving polynomial is
the so-called \emph{$q$-$\Phi$-multi-Catalan number}.
This conjecture extends the aforementioned result of Armstrong--Stump--Thomas, which is the case $m=1$.

\name{Dongkwan Kim} studies $q$-Kreweras numbers for coincidental Coxeter
groups attached to limit symbols \cite{Kim2020x}.  This gives another family
of $q$-analogues in the same root-poset and noncrossing-partition setting.

\name{Darij Grinberg} and \name{Tom Roby} prove a birational rowmotion
periodicity theorem for rectangular posets over suitable noncommutative rings
\cite{GrinbergRoby2023}.  Their result extends the field-valued periodicity
theorem for products of two chains and includes a noncommutative antipodal
reciprocity formula.


\subsection[cspListFrieze]{Polygonal dissections and Frieze patterns}

In \cite{AdamsBanian2025x}, the authors prove a CSP for the set $A_\mu$,
consisting of noncrossing dissections of an $(n+2)$-gon, where the number of
$j$-gons in the dissection is $\mu_j$.  The group action is rotation of the
$(n+2)$-gon.  This gives
\[
\left( A_\mu, C_{n+2},  \frac{1}{[n+1]_q} \qbinom{n+k}{k}\qbinom{k}{\mu_1,\mu_2,\dotsc,\mu_n} \right)
\]
where $k-1$ is the number of noncrossing diagonals in the dissection.

There is then a certain bijection between noncrossing dissections of
$(n+2)$-gons and classes of frieze patterns, see \cite{HolmJorgensen2018}.



\subsection[cspListPosetsMinuscule]{Minuscule posets}



In \cite{RushShi2012}, the authors consider \hyperref[rowmotion]{rowmotion}
acting on the \hyperref[antichain]{antichains} of minuscule posets.
The corresponding CSP-polynomial is the rank-generating function of the poset,
which for minuscule posets enjoys some particularly nice properties, see
\cite[Ex.~170]{StanleyEC1}.

For the product of two chains, this action is a special case of the
\hyperref[cspPlanePartitions]{plane partitions CSP}.

Hopkins (see \cite{Hopkins2024}) conjectured a CSP for the action of piecewise-linear
rowmotion on height $m$ $P$-partitions when $P$ is a minuscule poset, where the
sieving polynomial is the rank-generating function of $P\times[m]$,
which again has nice properties (in particular, a product formula).
This conjecture extends the aforementioned result of Rush--Shi, which is the case $m=1$.

For \hyperref[planePartitions]{plane partitions} with additional symmetries
and related cyclic-sieving conjectures, see \cite{Hopkins2020}.



\subsection[csp01Matrices]{BiCSP on $0$--$1$-matrices and Hall--Littlewood}

In \cite{BarceloReinerStanton2008}, a bi-cyclic sieving phenomenon on
\hyperref[permutationMatrix]{permutation matrices} is described.
Let $X_n$ be the set of $n\times n$ permutation matrices,
and let $\grpc_n \times \grpc_n$ act on $X_n$ by cyclic shifts of rows and columns.
Then
\[
\left(X_n, \grpc_n \times \grpc_n, \epsilon(q,t)\sum_{\lambda \vdash n} f^\lambda(q) f^\lambda(t) \right)
\]
exhibits the bi-cyclic sieving phenomenon.
Here, $\epsilon(q,t)$ is $(qt)^{n/2}$ if $n$ is even, and $1$ if $n$ is odd.
In fact, they prove a much stronger statement regarding complex reflection groups.

This is further generalized in \cite{Rhoades2010b} to $0$--$1$-matrices,
where the column sums and row sums in the matrices are given by fixed integer partitions.

\name{Brendon Rhoades} further generalizes this to matrices with nonnegative
integer entries, and proves a bi-cyclic phenomenon using
\[
\epsilon(q,t)\sum_{\lambda \vdash n} K_{\lambda,\mu}(q) K_{\lambda,\nu}(t)
\]
as a biCSP polynomial.  Note that this is closely
related to the \hyperref[rsk]{Robinson--Schensted--Knuth algorithm}.



\subsection[cspTensors]{Tensor products}

In \cite{Westbury2016}, cyclic sieving phenomena are created via representation theory
and crystals, generalizing the promotion CSP by \name{Brendon Rhoades}.
This is related to the principal specialization of
\hyperref[frobeniusCharacteristic]{Frobenius image} of
characters of the symmetric group.



\subsection[cspLatticePolytopes]{Lattice polytopes}

\name{Jim Propp} asks for a solution to the following problem.
Let $P \subset \setR^d$ be a \hyperref[latticePolytope]{lattice polytope} and
$g$ be a \emph{linear map} $\setR^d \to \setR^d$, such that
$\langle g \rangle$ is a cyclic group of order $n$ where $g(P) = P$.

Can we find a linear function $\sigma:\setZ^d \to \setZ$ such that for the
$q$-analogue of the \hyperref[ehrhart]{Ehrhart function} of $P$, defined as
\[
p_m(q) \coloneqq \sum_{x \in mP \cap \setZ^d} q^{\sigma(x)},
\]
such that $(mP \cap \setZ^d, \langle g \rangle, p_m(q))$ is a CSP-triple
for all $m \in \setN$?
That is, we get a family of cyclic sieving phenomena on the lattice points of dilations of $P$.

\name{Dongchun Han}, \name{Xuan Wang}, \name{Hanbin Zhang}, and
\name{Shiwen Zhang} study zero-sum polytopes associated with finite abelian
groups, including reciprocity, rigidity, and cyclic sieving phenomena
\cite{HanWangZhangZhang2026x}.


\subsection[cspIndependentSetsGraphs]{Independent sets of graphs}

\name{Jacob A. White} gives cyclic sieving phenomena for independent sets of
fixed size in graphs equipped with cyclic group actions \cite{White2026x}.
The main examples are powers of cycle graphs, and the same work gives closed
formulas for the number of independent sets of fixed size in powers of cycles
and paths.  Whiskering gives a way to build new CSP examples from old ones,
and recursive methods handle independent sets in gear graphs, helm graphs,
and book graphs.


\subsection[cspStrongDichotomy]{Strong dichotomy classes}

\name{Octavio A. Agust{\'i}n-Aquino} studies cyclic sieving in the
enumeration of strong dichotomy classes, that is, bicolour self-complementary
and rigid patterns in $\setZ/2k\setZ$ \cite{AgustinAquino2026x}.  The paper
proves that, for odd $k$, evaluating the rigid pattern-inventory polynomial
at $-1$ gives the negative of the number of strong classes.


\subsection[cspCyclicCodes]{Hamming codes}

\name{Alexander Mason}, \name{Vic Reiner} and \name{Shruthi Sridhar} 
have some results on cyclic codes and (dual) Hamming codes in \cite{MasonReinerSridhar2020x}.


\subsection[cspMaximalWeights]{Dominant maximal weights}

A cyclic sieving phenomenon on dominant maximal weights in affine Kac--Moody
algebras is described in \cite{KimOhOh2020}.
