\metatitle{Gessel quasisymmetric functions}
\metadescription{An introduction to Gessel quasisymmetric functions, including
their definition, product rules, specializations, 0-Hecke interpretation, and
relationships with the power-sum and quasisymmetric power-sum bases.}
\metakeywords{Gessel quasisymmetric functions,fundamental quasisymmetric
functions,product rule,specializations,power-sum basis,quasisymmetric
power-sum basis,0-Hecke algebra}

\section[gessel]{Gessel quasisymmetric functions}

\begin{polydata}{gessel}
  Name   & Gessel quasisymmetric functions \\
  Space    & QSym \\
  Basis    & True \\
  Rating   & 4 \\
  Bib      & Gessel1984 \\
  Year     & 1984 \\
  Symbol   & $\gessel_\alpha(x)$ \\
  Category & QuasiElementary \\
  PositiveIn & qmonom | Gessel1984 \\
\end{polydata}


The \defin{Gessel quasisymmetric functions} $\gessel_\alpha(\xvec)$,
also called the \defin{fundamental quasisymmetric functions},
were introduced by \name{Ira Gessel} in \cite{Gessel1984}.
They are indexed by integer compositions and constitute 
a basis for the space of \hyperref[quasiSymmetricFunctions]{quasisymmetric functions}. 


\subsection[gesselDefinitionCompositions]{Definition (compositions)}

For $\alpha \vDash n$, we define
\begin{equation*}
 \gessel_\alpha(\xvec) = \sum_{\beta \leq \alpha} M_\beta(\xvec) =
 \sum_{\beta \leq \alpha}
 \sum_{i_1 \lt \dotsb \lt i_{\length(\beta)}}
 x_{i_1}^{\beta_1} \dotsm x_{i_{\length(\beta)}}^{\beta_{\length(\beta)}}
\end{equation*}
where $\leq$ is the \hyperref[compositionRefinementOrder]{refinement partial
order}.
There is a second way to index the Gessel quasisymmetric functions, by using
subsets.

\subsection[gesselDefinitionSubsets]{Definition (subsets)}

Let $S \subseteq [n-1]$.
Then we define $\gessel_{n,S}(\xvec)$ as
\begin{equation*}
 \gessel_{n,S}(\xvec) =
 \sum_{\substack{1 \leq b_1 \leq \dotsb \leq b_n \\
 i \in S \Rightarrow b_i \lt b_{i+1} }}
 x_{b_1} \dotsm x_{b_n}
\end{equation*}

To translate between the two definitions, note that
$\gessel_\alpha(\xvec) = \gessel_{n,S_\alpha}(\xvec)$, where, for
$\ell=\length(\alpha)$,
$S_\alpha = \{\alpha_1, \alpha_1+\alpha_2,\dotsc,
\alpha_1+\dotsb + \alpha_{\ell-1}\}$.
Conversely, $\gessel_{n,S}(\xvec) = \gessel_{comp(S)}$, where
$comp(S) = (s_1,s_2-s_1,s_3-s_2,\dotsc,s_{\ell} - s_{\ell-1},n-s_{\ell})$.

In terms of the \hyperref[qmonom]{monomial quasisymmetric functions},
we have
$\gessel_{\alpha}(\xvec)
= \sum_{\beta \preceq \alpha} \qmonom_\beta(\xvec)$, where $\preceq$ denotes
\hyperref[compositionRefinementOrder]{composition refinement}.

\section[gesselProperties]{Properties}

See \hyperref[standardQuasiInvolution]{the page on quasisymmetric functions}
for how the \hyperref[standardInvolution]{standard involution on symmetric functions}
that sends $\schurS_\lambda$ to $\schurS_{\lambda'}$ 
can be extended to quasisymmetric functions.


\subsection[gesselProduct]{Product rule}

The product $\gessel_{m,S} \gessel_{n,T}$ has the following positive rule.
Choose $\sigma \in \symS_m$ and $\tau \in \symS_n$ such that the
\hyperref[descentSet]{descent sets} of the permutations satisfy
$S = \DES(\sigma)$ and $T = \DES(\tau)$.
A \defin{shuffle}\label{shuffleOfPermutations} of $\sigma$ and $\tau$ is a
permutation in $\symS_{m+n}$ such that entries $1,\dotsc,m$ appear in the same
relative order as in $\sigma$, and the entries $m+1,\dotsc,m+n$ appear in the
same relative order as in $\tau$ after $m$ is subtracted from these entries.
For example, the shuffles of the permutations $21$ and $12$ are given by
\[
2134, 2314, 3214, 2341, 3241, 3421.
\]
There are $\binom{m+n}{m}$ shuffles in total.
We then have
\[
\gessel_{m,S}(\xvec) \gessel_{n,T}(\xvec) =
\sum_{\pi \text{ shuffle of } \sigma,\tau }
\gessel_{m+n,\DES(\pi)}(\xvec).
\]


\subsection[gesselSpecializations]{Specializations}

For $\length(S)\lt m$, we have
\begin{equation*}
 \gessel_{n,S}(1^m) = \binom{n+m-\length(S)-1}{n}
\end{equation*}
and for $\length(S)\lt m$,
\begin{equation*}
 \gessel_{n,S}(1,q,q^2,\dotsc,q^m) = q^{m \length(S)+m-|S|-1}\qbinom{n+m-1-\length(S)}{m}_q
\end{equation*}
where $|S|$ denotes the sum of the elements of $S$.


\subsection[gesselSlinkyRule]{Egge--Loehr--Warrington and the slinky rule}

In \cite{EggeLoehrWarrington2010}, the following lifting from fundamental
quasisymmetric functions to Schur functions was given.
An alternative interpretation of this rule is proved in \cite{GarsiaRemmel2018}.
A short proof by \name{Ira Gessel} is given in \cite{Gessel2018OnTheSchurFunction}.

Suppose $X(\xvec)$ is a symmetric function with fundamental quasisymmetric
expansion
\[
 X(\xvec) = \sum_\alpha c_\alpha \gessel_\alpha(\xvec)
\]
Then
\[
 X(\xvec) = \sum_\alpha c_\alpha \schurS_\alpha(\xvec).
\]
The composition indices in $\schurS_\alpha(\xvec)$ should be 
computed via the \hyperref[schurJacobiTrudi]{Jacobi--Trudi identity}
using the \hyperref[completeH]{complete homogeneous symmetric functions}.
By using determinant formulas, the compositions can be straightened into
partitions.  This can be done using the \defin{slinky rule}: pictorially, the
parts of the compositions fall down when $\alpha$ is illustrated in French
notation.  See \cite{EggeLoehrWarrington2010}.

\begin{example*}[The slinky rule]

We use the slinky rule on the composition $(4,3,8)$.

\begin{figure}
\begin{ytableau}
*(lightblue) & *(lightblue) & *(lightblue) & *(lightblue) & *(lightblue) & *(lightblue) & *(lightblue) & *(lightblue) \\
*(gray) & *(gray) & *(gray) \\
*(lightgreen) & *(lightgreen) & *(lightgreen) & *(lightgreen)
\end{ytableau}

\begin{ytableau}
*(lightblue) & *(lightblue) & *(lightblue) & *(lightblue) \\
*(gray) & *(gray) & *(gray) & *(lightblue) & *(lightblue) & *(lightblue) & *(lightblue) \\
*(lightgreen) & *(lightgreen) & *(lightgreen) & *(lightgreen)
\end{ytableau}

\begin{ytableau}
*(lightblue) & *(lightblue) & *(lightblue) & *(lightblue) \\
*(gray) & *(gray) & *(gray) & *(lightblue) & *(lightblue)  \\
*(lightgreen) & *(lightgreen) & *(lightgreen) & *(lightgreen) & *(lightblue) & *(lightblue) 
\end{ytableau}
\end{figure}

The resulting shape is $(6,5,4)$ and the sign is $(-1)^2$ as the height 
of the ribbon of size $8$ ends two levels below its starting row.
That is, the height of a ribbon is computed in the same 
manner as in the \hyperref[schurMurnaghanNakayama]{Murnaghan--Nakayama rule}.
Hence, $\schurS_{438}=\schurS_{654}$.


As a second example, $\schurS_{25115} = -\schurS_{43322}$ according to the
slinky rule.

\begin{figure}
\begin{ytableau}
*(lightblue) & *(lightblue) & *(lightblue) & *(lightblue) & *(lightblue) \\
\; \\
\; \\
*(gray) & *(gray) & *(gray) & *(gray) & *(gray) \\
*(lightgreen) & *(lightgreen)
\end{ytableau}

\begin{ytableau}
*(lightblue) & *(lightblue) \\
\; & *(lightblue)  \\
\; & *(lightblue)  & *(lightblue) \\
*(gray) & *(gray) & *(gray) \\
*(lightgreen) & *(lightgreen)  & *(gray) & *(gray)
\end{ytableau}
\end{figure}

Note that the sign is $(-1)^{2}(-1)^{1} = -1$.

Finally, $\schurS_{12} = 0$, since there are not enough boxes to perform a
slinky move.
Similarly, $\schurS_{2115} = 0$, since the slinky rule does not give a partition.
\begin{figure}
\begin{ytableau}
*(lightblue) & *(lightblue) & *(lightblue) & *(lightblue) & *(lightblue) \\
\; & \none \\
\; & \none \\
*(lightgreen) & *(lightgreen) 
\end{ytableau}
\begin{ytableau}
*(lightblue) & *(lightblue) \\
\; & *(lightblue)  \\
\; & *(lightblue) & *(lightblue)  \\
*(lightgreen) & *(lightgreen) 
\end{ytableau}
\end{figure}

\end{example*}


\subsection[gesselPowerSum]{Relationship with the power-sum basis}


\name{Christos A. Athanasiadis} \cite{Athanasiadis2015}, also implicitly in
\cite{AdinRoichman2015}, proves the following result.
Suppose $X(\xvec)$ is a symmetric function such that
\begin{equation*}
X(\xvec) = \sum_{S \subseteq [n-1]} a_S \gessel_{n,S}(\xvec).
\end{equation*}
Then
\begin{equation*}
X(\xvec) =
\sum_{\lambda \vdash n} z_{\lambda}^{-1} \powerSum_{\lambda}(\xvec)
\sum_{S\in U_\lambda} (-1)^{|S\setminus S_\lambda|} a_S,
\end{equation*}
where $U_\lambda$ is the set of
\defin{$\lambda$-unimodal subsets}\label{lambdaUnimodalSubset} of $[n-1]$
and $S_\lambda \coloneqq \{ s_1,s_2,\dotsc,s_{\ell-1} \}$.

Here, $s_i$ is defined as $\lambda_1+\dotsb + \lambda_{i}$
and a set $A \subseteq [n-1]$ is $\lambda$-unimodal if for $0\leq i \lt \ell$
the intersection of $A$ with the intervals
$\{s_i+1,s_i+2,\dotsc,s_{i+1}-1\}$
is a prefix of the latter.


\subsection[gesselQuasiPowerSum]{Relationship with a quasisymmetric power-sum basis}

In \cite{BallantineDaughertyHicksMason2020}, two versions of quasisymmetric
power-sum bases are given.
They give the fundamental quasisymmetric expansion of these two quasisymmetric
power-sum bases.

In \cite{AlexanderssonSulzgruber2019}, we provide the
\hyperref[qPsi]{quasisymmetric power-sum} expansion of $\gessel_\alpha(\xvec)$.
We have
\begin{equation*}
\gessel_{n,S}(\xvec) =
\sum_{\alpha} \frac{\qPsi_\alpha(\xvec)}{z_\alpha}
(-1)^{|S\setminus S_\alpha|}
\end{equation*}
where the sum is taken over all compositions $\alpha$ such that the set $S$ is
\hyperref[lambdaUnimodalSubset]{$\alpha$-unimodal}.
This formula implies Athanasiadis's formula above.


\name{Austin Roberts} gives a criterion for promoting symmetric
\hyperref[gessel]{fundamental-positive} expansions to
\hyperref[schurPositivity]{Schur-positive} expansions by grouping the
fundamental terms into suitable equivalence classes \cite{Roberts2016}.

\name{Sarah K. Mason} and \name{Elizabeth Niese} introduce a super analogue
of the fundamental quasisymmetric basis in their study of
\hyperref[hookSchur]{quasisymmetric hook Schur functions}
\cite{MasonNiese2018}.  Their quasisymmetric hook Schur functions expand
positively in this super Gessel fundamental basis.


\section[zeroHeckeAlgebra]{The 0-Hecke algebra}

The \defin{0-Hecke algebra}\label{zeroHeckeAlgebraDefinition} $H_n(0)$ in type
$A$ is the $\setC$-algebra generated by $T_i$, $i=1,\dotsc,n-1$, subject to
the relations
\[
T_i^2 = T_i \qquad T_iT_{i+1}T_i = T_{i+1}T_iT_{i+1}\quad 
T_{j}T_i = T_{i}T_j \text{ whenever } |i-j|\geq 2.
\]
Note that $T_{s_1}\dotsc T_{s_{\ell}}$ is independent of the choice of
\hyperref[reducedWord]{reduced word} $w = s_1\dotsm s_{\ell}$, with
$w \in \symS_n$.
It follows that the dimension of $H_n(0)$ is $n!$.

As an example of such operators, the divided difference operators 
generating the \hyperref[keyDefinitionOperators]{key polynomials} are such operators.
This also highlights the connection with the \hyperref[macdonaldE]{non-symmetric Macdonald polynomials}.

Similar to how $\symS_n$-modules decompose into irreducible modules indexed by
partitions, we can completely characterize indecomposable $H(0)$-modules; see
\cite{KrobThibon1997}.
There are $2^{n-1}$ irreducible representations $F^\alpha$, indexed by integer
compositions of $n$.

The map $\frobChar$ on a $H_n(0)$-module $M$, sending $F^\alpha \to \gessel_\alpha$ 
is the \defin{quasisymmetric characteristic}\label{quasisymmetricCharacteristic} of $M$.
This is an analogue of the \hyperref[frobeniusCharacteristic]{Frobenius characteristic}.
Under this characteristic, the simple modules correspond to the fundamental
quasisymmetric functions.  The projective indecomposable modules give the
dual noncommutative symmetric side: the Grothendieck group of projective
$0$-Hecke modules is identified with $\mathrm{NSym}$, while the Grothendieck
group of finite-dimensional modules is identified with $\spaceQSym$
\cite{KrobThibon1997}.  Induction and restriction then recover the dual Hopf
algebra structures on these two spaces.

For example, in \cite{TewariWilligenburg2015}, the authors show that
\hyperref[qSchur]{quasisymmetric Schur functions} are given as the
quasisymmetric characteristic of certain $H_n(0)$-modules.
\name{Seung-Il Choi}, \name{Young-Hun Kim}, \name{Sun-Young Nam}, and
\name{Young-Tak Oh} construct $0$-Hecke modules from standard permuted
composition tableaux \cite{ChoiKimNamOh2021}.

See also \cite{Huang2013} for some examples of graded vector spaces.
\name{Young-Hun Kim} and \name{Dominic Searles} construct type $B$ poset
modules for the type $B$ 0-Hecke algebra \cite{KimSearles2026x}.  Their
quasisymmetric characteristics give a representation-theoretic interpretation
of type $B$ $P$-partition enumerators, and the Grothendieck group they obtain
is isomorphic to the space of type $B$ quasisymmetric functions as a
$\spaceQSym$-module and comodule.
