\metatitle{Gelfand--Tsetlin patterns and polytopes}
\metadescription{Definition of Gelfand--Tsetlin patterns and Gelfand--Tsetlin polytopes. Bijection with semistandard Young tableaux. Related results.}


\section[gtpatterns]{Gelfand--Tsetlin patterns}

Gelfand--Tsetlin (sometimes spelled Cetlin or Zetlin) patterns
were introduced by \name{Israel Gelfand} and \name{Michael Tsetlin} in \cite{GelfandTsetlin1950},
motivated by the close relationship with representation theory of $GL_n$.
A short introduction to GT-patterns can be found in \cite[p.~103]{StanleyEC2}.

For a brief background, see the Gelfand--Tsetlin pattern collection in
FindStat \cite{FindStatGelfandTsetlinPatterns}.

\name{Ilse Fischer} and \name{Florian Schreier-Aigner} prove product formulas
for $(-1)$-enumerations of arrowed Gelfand--Tsetlin patterns
\cite{FischerSchreierAigner2023x}.  Their proofs use a recent
Littlewood-type identity and determinant evaluations via LU decompositions,
Sister Celine's algorithm, and creative telescoping.


A \defin{Gelfand--Tsetlin pattern}\label{gtPattern} or \defin{GT-pattern}
is a triangular (or sometimes parallelogram) arrangement of non-negative integers,
with certain constraints.
The triangular patterns are indexed as follows.
\[
\begin{matrix}
x_{n1} & & x_{n2} & & \cdots & & \cdots & & x_{nn} \\
 & \ddots & & \ddots &  & &  \iddots &  & &  \\
%   &  & x_{13} &  & x_{23}  &  & x_{33}\\
%   &  &  &  x_{12} &   & x_{22}  \\
   &  &   x_{21} &  & x_{22}\\
   &  &    & x_{11}
\end{matrix}
\]
The less common parallelogram patterns are indexed as follows.
\[
\begin{matrix}
x_{n1} & & x_{n2} & & \cdots & & \cdots & & x_{nm} \\
 & \ddots & & \ddots &  & &   & & & \ddots  \\
%   &  & x_{13} &  & x_{23}  &  & x_{33}   &  & \\
%   &  &  &  x_{12} &   & x_{22}  &     &  & \\
   &  &   x_{11} &  & x_{12} & & \cdots & & \cdots & & x_{1m} \\
   &  &    & x_{01} &    & x_{02} & & \cdots & & \cdots & & x_{0m}
\end{matrix}
\]
The entries must satisfy the conditions
\begin{equation*}
x_{i+1,j} \geq x_{ij} \text{ and } x_{ij} \geq x_{i+1,j+1}
\end{equation*}
for all values of $i$, $j$ where the indexing is defined.
The inequalities simply state that horizontal rows are weakly decreasing,
down-right diagonals are weakly decreasing and down-left diagonals are weakly increasing.

The \defin{weight}\label{gtPatternWeight} $w(G) = (w_1,\dotsc,w_n)$ of a
GT-pattern $G$ is defined as the vector where $w_i$ is the difference of the
sum of row $i$ and $i-1$.
By convention, $x_{0j}=0$ if these entries are not specified.


\subsection[gtpatternsAsSSYT]{Bijection with semistandard Young tableaux}

The conditions ensure every row in the pattern defines an integer partition.
Furthermore, any two adjacent rows define a skew shape.
These properties enable us to biject triangular GT-patterns with
\hyperref[prelimTableaux]{Young tableaux} and parallelogram GT-patterns with
\hyperref[skewShapes]{skew Young tableaux}.

The skew shape defined by row $j$ and $j+1$ in a GT-pattern $G$
describes which boxes in a tableau $T$ have content $j$.
In particular, if $T$ has shape $\lambda/\mu$ then the first row in $G$ is $\mu$
and the last row is $\lambda$.


\begin{example}
The following triangular GT-pattern
\[
\begin{matrix}
5 &   & 4 &   & 2 &   & 1 &   & 1 &  & 0\\
 & 5 &   & 3 &   & 2 &   & 1 &   & 0\\
 &  & \color{blue} 3 &   & \color{blue} 3 &   & \color{blue} 2 &   & \color{blue} 1  \\
 &  &  & 3 &   & 3 &   & 1 &  \\
 &  &  &  & 3 &   & 2 &  \\
 &  &  &  &  & 3 & \\
\end{matrix}
\]
is in bijection with the semistandard Young tableau below.

\begin{figure}
\begin{ytableau}
*(lightblue)1 & *(lightblue)1 & *(lightblue)1 & 5 & 5 \\
*(lightblue)2 & *(lightblue)2 & *(lightblue)3 & 6 \\
*(lightblue)3 & *(lightblue)4 \\
*(lightblue)4 \\
6
\end{ytableau}
\end{figure}

Notice that the fourth row in the GT-pattern is $3321$ and
the shape of all entries less than or equal to four in the tableau is also $3321$.
\end{example}

\begin{example}
The skew Young tableau
\begin{figure}
\begin{ytableau}
\none  & \none & 1 & 3 \\
\none  & 1 & 2 \\
\none  & 4 \\
2
\end{ytableau}
\end{figure}
is in bijection with the GT-pattern
\[
\begin{matrix}
4 &   & 3 &   & 2 &   & 1\\
 & 4 &   & 3 &   & 1 &   & 1\\
 &  & 3 &   & 3 &   & 1 &   & 1\\
 &  &  & 3 &   & 2 &   & 1 &   & 0\\
 &  &  &  & 2 &   & 1 &   & 1 &   & 0\\
\end{matrix}
\]
Observe that in any GT-pattern, $x_{i+1,j} - x_{i,j}$ counts the
number of boxes with content $i$ in row $j$ in the corresponding tableau.
Thus, the weight of a GT-pattern is the same as the weight of the corresponding tableau.
\end{example}

There is a \hyperref[crystals]{crystal graph structure} on semistandard Young tableaux.
The corresponding crystal graph on GT-patterns is described explicitly in
\cite{HartwigKingston2026}.


\subsection[gtpatternsTokuyama]{Tokuyama's formula}

Let $SGT(\lambda)\subset \gtp(\lambda)$ be the set of
\defin{strict Gelfand--Tsetlin patterns}\label{strictGTPattern}.
This is the subset of patterns where each row is \emph{strictly decreasing}.
Consider tuples $(a,b,c)$ in the pattern arranged as
$\begin{smallmatrix}
a && b\\ & c \end{smallmatrix}$.
Let $S(G)$ be the number of entries $c$ such that $a\gt c \gt b$,
and let $L(G)$ be the number of entries $c$ such that $a=c$.

Let $\rho = (n-1,n-2,\dotsc,1,0)$. Then (see \cite{Tokuyama1988})
\begin{equation*}
\schurS_{\lambda}(x_1,\dotsc,x_n) \cdot \prod_{1 \leq i \lt j \leq n} (x_i + t x_j)  =
\sum_{G \in SGT(\lambda + \rho)} (1+t)^{S(G)} t^{L(G)}\xvec^{w(G)}.
\end{equation*}

Note that this interpolates between the \hyperref[schurWeylformula]{Weyl character formula}, and the SSYT definition of \hyperref[schurS]{Schur polynomials}.


See also \cite{Stanley1986} for some special cases, and connection with
\hyperref[planePartitions]{plane partitions} and
\hyperref[alternatingSignMatrix]{alternating sign matrices}.
\name{Yulia Alexandr}, \name{Patricia Commins}, \name{Alexandra Embry},
\name{Sylvia Frank}, \name{Yutong Li}, and \name{Alexander Vetter} study
type $B$ analogues via ice models \cite{AlexandrComminsEmbryFrankLiVetter2018x}.
They construct models in bijection with Gelfand--Tsetlin-type patterns and
shifted tableaux for the Sundaram rules, obtaining partition functions that
deform the Weyl character formula for $SO(2n+1,\setC)$.

\section[gtpolytopes]{Gelfand--Tsetlin polytopes}

By specifying the top row in a triangular GT-pattern as $\lambda$
and imposing the inequalities above, we obtain a convex polytope.
This is a \defin{Gelfand--Tsetlin polytope}\label{gtPolytope},
$\gtp_{\lambda} \subset \setR^{n(n+1)/2}$.
By construction, the integer lattice points in $\gtp_{\lambda}$
are in bijection with $\SSYT(\lambda)$.
The analogous statement holds for parallelogram GT-patterns where the bottom
row $\mu$ is also a fixed integer partition.
These polytopes are denoted $\gtp_{\lambda\mu} \subset \setR^{nm}$.
Below, we use $d$ to denote the dimension ($n(n+1)/2$ or $mn$) of the ambient space.


\subsection[gtpolytopesMarkedPoset]{Unimodular triangulation and marked posets}

The polytopes $\gtp_{\lambda\mu}$ admit \emph{unimodular triangulations}.
This means that the polytopes can be decomposed into simplices with integer
vertices and normalized volume $1$. For a short proof, see
\cite{Alexandersson2019CounterExamples}.

As a consequence, Gelfand--Tsetlin polytopes have integer vertices,
and have the \hyperref[polytopeIDP]{integer decomposition property} (IDP).
This means that
if $p$ is a lattice point in $k \gtp_{\lambda\mu}$ for some integer $k\geq 1$,
then $p$ can be expressed as $p_1 + \dotsb + p_k$ where each $p_i$ is a lattice point in $\gtp_{\lambda\mu}$.
This can be proved directly by bijecting to semistandard tableaux,
see \cite[Subsec.~2.3]{Alexandersson2016GTPoly}.  Under this bijection,
elementwise addition of GT-patterns corresponds to concatenating tableaux:
a tableau of shape $k\lambda/k\mu$ decomposes into $k$ tableaux of shape
$\lambda/\mu$.

\name{Akiyoshi Tsuchiya} generalizes the notion of IDP to $k$-tuples of
polytopes (the $k=2$ case was introduced in \cite{HaaseHofmann2017}).
The following theorem follows very easily from the bijection with SSYT.

\begin{theorem}[Alexandersson, 2018]
Let $\gtp_{\lambda^1},\dotsc,\gtp_{\lambda^k}$
be GT-polytopes, embedded in the same ambient space
by padding the partitions with zeros if needed.
Then
\[
\setZ^d \cap \left( \gtp_{\lambda^1} + \dotsb + \gtp_{\lambda^k} \right)
=
\left( \setZ^d \cap \gtp_{\lambda^1} \right) + \dotsb + \left(\setZ^d \cap \gtp_{\lambda^k} \right).
\]
By letting all $\lambda^i$ be equal, we recover the usual IDP.
\end{theorem}

This theorem does not generalize to skew shapes. The pair $\gtp_{3}$
and $\gtp_{2/1}$ yields a counterexample.


Gelfand--Tsetlin polytopes are special cases of
\hyperref[markedOrderPolytope]{marked order polytopes},
which generalize \hyperref[orderPolytope]{order polytopes} considered by
\name{Richard Stanley} \cite{Stanley86TwoPosetPolytopes}.
In \cite{ArdilaBliemSalazar2011}, the authors generalize the bijection
between order polytopes and \hyperref[chainPolytope]{chain polytopes} and
give a \defin{marked chain polytope}\label{markedChainPolytope} analog
of GT-polytopes.
In type $A$ these analogs are called
\defin{Feigin--Fourier--Littelmann polytopes}\label{fflvPolytope}.

\name{Igor Makhlin} studies
\defin{marked chain-order polytopes}\label{markedChainOrderPolytope}
for the \defin{Gelfand--Tsetlin poset}\label{gtPoset} \cite{Makhlin2024}.
He realizes their toric varieties
as explicit Gröbner/SAGBI degenerations of the \hyperref[flagVarieties]{flag
variety}, using \hyperref[schubertPipeDream]{pipe dreams} to construct the
Plücker-coordinate embeddings, and also obtains standard monomial theories and
PBW-type monomial bases.
\name{Xin Fang} and \name{Gleb Koshevoy} compare Lusztig polytopes and
FFLV polytopes \cite{FangKoshevoy2020x}.

\name{Oliver Clarke}, \name{Akihiro Higashitani}, and
\name{Fatemeh Mohammadi} relate Gelfand--Tsetlin and FFLV polytopes for
Grassmannians to
\defin{matching-field polytopes}\label{matchingFieldPolytope}
\cite{ClarkeHigashitaniMohammadi2024}.
They construct a sequence of combinatorial mutations between these polytopes
which stays inside the class of matching-field polytopes, giving toric
degenerations and Newton--Okounkov bodies for the
\hyperref[grassmannianVarieties]{Grassmannians}.

\name{Robert G. Donnelly} and \name{Molly W. Dunkum} construct
Gelfand--Tsetlin-type weight bases for all special linear Lie algebra
representations whose characters are skew Schur functions
\cite{DonnellyDunkum2022}.  Their bases give another representation-theoretic
version of the GT-pattern picture for skew Schur objects.
\name{Markus Fulmek} enumerates symmetric Gelfand--Tsetlin patterns by linear
algebra \cite{Fulmek2019x}.
\name{Hideya Watanabe} and \name{Keita Yamamura} give an explicit type $A$
crystal isomorphism between the Lenart--Postnikov alcove-path model and the
Gelfand--Tsetlin pattern model \cite{WatanabeYamamura2021}.

\name{Spencer Daugherty}, \name{Nicolle González}, \name{Bárbara Muniz},
\name{Pablo S. Ocal}, \name{Jianping Pan}, and \name{Jacinta Torres} study
the monoid algebra of \hyperref[prelimTableaux]{semistandard tableaux}
\cite{DaughertyGonzalezMunizOcalPanTorres2025x}.
For $m=n$ this algebra agrees with the
\defin{Gelfand--Tsetlin semigroup ring}\label{gtSemigroupRing}.  They
prove structural properties such as being Noetherian, reduced, Koszul, and
Cohen--Macaulay, describe maximal ideals and irreducible modules, and relate
the associated variety to toric degenerations of
\hyperref[flagVarieties]{partial flag varieties}.

The Gelfand--Tsetlin polytopes are also special cases of
\hyperref[flowPolytope]{flow polytopes}; see
\cite{LiuMeszarosDizier2019flow}.  In fact,
\name{Ricky Ini Liu}, \name[Karola Mészáros]{Karola Mészáros}, and
\name{Avery St. Dizier} develop a general connection between marked order
polytopes and flow polytopes, with GT-polytopes as a central example.
\name{Jihyeug Jang} and \name[J. S. Kim]{Jang Soo Kim} compute volumes of
flow polytopes related to caracol graphs \cite{JangKim2020}.


\subsection[birkhoffPolytopeAsGelfandTsetlin]{Birkhoff polytope as a GT-polytope}

There is a natural correspondence between \defin{contingency matrices}\label{contingencyMatrices}
and certain GT-polytopes \cite{Alexandersson2016GTPoly}.
A contingency matrix is a matrix of non-negative integers with specified row sums and column sums.

Given a skew shape $\lambda/\mu$ and the GT-polytope for weight
$\onevec = (1,1,\dotsc,1)$, each lattice point corresponds to a
semistandard Young tableau.
Let $a_{ij}$ denote the number of boxes with entry $j$ in row $i$ of the tableau.
This gives an $m \times n$ matrix $A = (a_{ij})$ where each column sums to $1$
and row $i$ sums to $\lambda_i - \mu_i$.
The change of variables between GT-patterns and these matrices is an integral affine transformation,
so lattice points in the GT-polytope are in bijection with contingency matrices
with column sums $1$ and row sums $\lambda_i - \mu_i$.

The special case $\lambda_i - \mu_i = 1$ with $n = m$ recovers the
\hyperref[birkhoffPolytope]{Birkhoff polytope}, which is the convex hull of
all $n \times n$ permutation matrices.
Since GT-polytopes are compressed (have the \hyperref[polytopeIDP]{IDP}),
it follows that the Birkhoff polytope is also compressed,
a fact proved independently in \cite{StanleyDecompositions80}.
There are several open questions about Birkhoff polytopes,
such as how to compute their volumes; see \cite{Pak2000}.




\subsection[gtEhrhart]{Ehrhart polynomials}

For a partition of length $m$,
\[
 \left| \setZ^d \cap \gtp_{\lambda} \right| = \schurS_{\lambda}(1^m)  =
 \prod_{1 \leq i \lt j \leq m } \frac{\lambda_i - \lambda_j + j-i}{j-i},
\]
where the last identity is due to \name{Richard Stanley}'s
\hyperref[schurSpecializations]{hook-content formula}.
Furthermore, the \hyperref[ehrhart]{Ehrhart polynomial} of $\gtp_{\lambda}$ is given by
\[
Ehr(\gtp_{\lambda};k) = \prod_{1 \leq i \lt j \leq m } \frac{k(\lambda_i - \lambda_j) + j-i}{j-i}.
\]


\begin{conjecture}[Alexandersson, Alhajjar \cite{AlexanderssonAlhajjar2018}]
The Ehrhart polynomial of $\gtp_{\lambda\mu}$ has non-negative coefficients.
Equivalently, the function
\[
f(k) = \schurS_{k\lambda/k\mu}(1^m)
\]
is a polynomial in $k$ with non-negative coefficients.
\end{conjecture}
For example, $m=3$, $\lambda=321$ and $\mu=1$ gives the Ehrhart polynomial
\[
Ehr(\gtp_{321/1},k) = \frac{1}{2} (k^5+6 k^4+14 k^3+16 k^2+9 k+2).
\]

Adding an extra row to $\lambda/\mu$ gives
\[
\schurS_{k\lambda/k\mu}(1^n) = K_{k\nu/k\tau,(km)^n},
\]
for appropriate $\nu/\tau$ and $m = |\lambda/\mu|$.
The shape $\lambda/\mu$ can be filled freely, and the remaining boxes in the
added row are then filled uniquely so that the weight becomes $m^n$.
A similar construction works for flagged skew Schur
functions and \hyperref[schurFlagged]{flagged skew Kostka coefficients}.

\name{Margaret Bayer}, \name{Bennet Goeckner}, \name{Su Ji Hong},
\name{Tyrrell McAllister}, \name{McCabe Olsen}, \name{Casey Pinckney},
\name{Julianne Vega}, and \name{Martha Yip} study Newton polytopes arising
from Schur polynomials and inflated symmetric
\hyperref[grothendieck]{Grothendieck polynomials}
\cite{BayerGoecknerHongMcAllisterOlsenPinckneyVegaYip2021}.  These polytopes
have the \hyperref[polytopeIDP]{integer decomposition property} by
\hyperref[newtonPolytopes]{saturated Newton polytope} results, and the paper
characterizes when they are \hyperref[reflexivePolytope]{reflexive}.



\begin{question}
Do all Kogan faces of Gelfand--Tsetlin polytopes have non-negative Ehrhart polynomial?
There is an example (see \cite{Alexandersson2019CounterExamples})
of a (non-Kogan) face where the Ehrhart polynomial
has a negative coefficient. However, this polynomial is of degree 20.
\end{question}


\subsection[gtpolytopesFVectors]{Vertices, $f$-vectors and differential equations}

There has been some effort in counting the number of vertices of GT-polytopes.
In \cite{GusevKiritchenkoTimorin2013}, the authors
describe a generating function that enumerates the vertices of $\gtp_{\lambda}$.

Let $V(1^{i_1},2^{i_2},\dotsc,k^{i_k})$ denote the number of vertices of
$\gtp_{\lambda}$ where $\lambda$ is the partition with \hyperref[prelimPartitions]{type} $(i_1,i_2,\dotsc,i_k)$.
Consider the generating function
\[
E_k(z_1,\dotsc,z_k) \coloneqq
\sum_{i_1,\dotsc,i_k \geq 0} V(1^{i_1},2^{i_2},\dotsc,k^{i_k}) \frac{z_1^{i_1}}{i_1!} \dotsm \frac{z_k^{i_k}}{i_k!}.
\]
\begin{theorem}[See \cite{GusevKiritchenkoTimorin2013}]
The formal power series $E_k(z_1,\dotsc,z_k)$ satisfies the partial differential equation
\[
\left( \frac{\partial^k}{\partial z_1 \dotsm \partial z_k} -
\left(\frac{\partial}{\partial z_1}-\frac{\partial}{\partial z_2}\right)
\dotsm
\left(\frac{\partial}{\partial z_{k-1}}-\frac{\partial}{\partial z_k}\right) \right)E_k = 0
\]
\end{theorem}

This result is later improved in \cite[Thm.~1.14]{AnChoKim2018},
where the authors provide a differential equation
for the full $f$-vector of GT-polytopes.
They consider the generating function
\[
F_k(z_1,\dotsc,z_k;t) \coloneqq
\sum_{d\geq 0} \sum_{i_1,\dotsc,i_k \geq 0} t^d f_d(1^{i_1},2^{i_2},\dotsc,k^{i_k}) \frac{z_1^{i_1}}{i_1!} \dotsm \frac{z_k^{i_k}}{i_k!}.
\]
where $f_d$ is the number of $d$-dimensional faces of $\gtp_{\lambda}$.


\subsection[gtpolytopesTypeBCD]{Type $B$ and $C$}

There are analogs of GT-polytopes for type $B$ and $C$, see \cite{ArdilaBliemSalazar2011}.


\section[gtKostka]{Weight-restricted Gelfand--Tsetlin polytopes}

Let $\gtp_{\lambda,w}\subseteq \gtp_{\lambda}$ be the
\defin{weight-restricted Gelfand--Tsetlin polytope}\label{weightRestrictedGTPolytope}
defined as $\gtp_{\lambda}$ but with the additional constraint that the
difference of the row sums of rows $j+1$ and $j$ is exactly $w_j$.

By construction, lattice points in $\gtp_{\lambda,w}$ are in bijection with $\SSYT(\lambda,w)$.
Therefore, the number of lattice points in $\gtp_{\lambda,w}$ is
given by the \hyperref[schurCombinatorialFormula]{Kostka coefficient} $K_{\lambda w}$.

We do the analogous definitions for parallelogram GT-patterns and skew shapes.


\begin{theorem}[See \cite[p.~179]{Kirillov2001}]
If $\lambda$ has $r$ parts and $\mu$ has $s$ parts, then the dimension of the
polytope $\gtp_{\lambda,\mu}$ (assuming non-empty) is given by
\[
 \dim \gtp_{\lambda,\mu} = (r-1)(s-1) - \binom{r}{2} - \sum_{i=1}^{r} \binom{\lambda'_i - \lambda'_{i+1}}{2}.
\]
\end{theorem}


\subsection[gtKostkaVertices]{Non-integrality and vertices}

Contrary to $\gtp_{\lambda}$, the polytopes (also called Gelfand--Tsetlin polytopes) $\gtp_{\lambda,w}$
are not always integral. In fact, the denominators of the
vertices can get arbitrary large, see \cite{DeLoeraMcAllister2004}.

In \cite{Alexandersson2016GTPoly}, we prove that for $\lambda \vdash n$,
the polytope $\gtp_{\lambda,1^n}$ is non-integral unless $\lambda=(2,2)$ or a hook shape.


\subsection[gtKostkaSaturation]{Saturation theorem}

The following theorem is a GT-polytope analog of the celebrated
\defin{Saturation theorem}\label{lrSaturationTheorem},
proved by \name{Allen Knutson} and \name{Terence Tao}
\cite{KnutsonTau1999,KnutsonTaoWoodward2004}.
They proved a similar statement for so-called
\defin{Berenstein--Zelevinsky polytopes}\label{berensteinZelevinskyPolytope}.

\begin{theorem}[See \cite{Alexandersson2015KSaturation}]

Every \emph{non-empty} Gelfand--Tsetlin polytope $\gtp_{\lambda/\mu,w}$
contains at least one vertex with integer coordinates.

\end{theorem}



\subsection[gtKostkaEhrhart]{Ehrhart polynomials and period collapse}

Since the $\gtp_{\lambda/\mu,w}$ are not lattice \hyperref[polytope]{polytopes},
one does not expect the \hyperref[ehrhart]{Ehrhart function} to be a polynomial.
However, these polytopes exhibit a \defin{period collapse}\label{periodCollapse}
\cite{HaaseMcAllister2007}
and they do have a polynomial \hyperref[ehrhart]{Ehrhart function}.
See \cite{Rassart2004} for a proof of this statement.


\begin{conjecture}[See \cite[p.~170]{Kirillov2001} and later \cite{KingTolluToumazet2004}]
The coefficients of the Ehrhart polynomial of $\gtp_{\lambda/\mu,w}$ are non-negative.
\end{conjecture}

See also \cite{Alexandersson2015KSaturation} for related open conjectures.


We could also consider a fixed face of the Gelfand--Tsetlin polytope intersected
with the hyperplane defined by a weight.
Suppose $P$ is the polytope obtained from the non-skew shape $\lambda = (2,2,2)$
and $w = (1,1,\dotsc,1)$, intersected with the face requiring
that $3$ is not in the first row of the corresponding SSYTs.
Then the number of lattice points in $kP$ is
\[
  3, 7, 13, 22, 34, 50, 70, 95, 125, 161, 203 \dotsc, \text{ for } k=1,2,\dotsc,
\]
which is a quasipolynomial of degree 3 with period 2.





\subsection[gtKostkaInterior]{Interior lattice points and strict inequalities}

The \defin{interior} of a Gelfand--Tsetlin polytope consists of
all points satisfying the interlacing inequalities strictly:
\[
x_{i+1,j} > x_{ij} > x_{i+1,j+1}.
\]
Integer points in the interior are strict GT-patterns,
where every interlacing condition holds with strict inequality
(cf.\ $SGT(\lambda)$ in \hyperref[gtpatternsTokuyama]{Tokuyama's formula}).

By \hyperref[ehrhartMacdonaldReciprocity]{Ehrhart--Macdonald reciprocity},
the number of interior lattice points in the $t$-th dilate satisfies
\[
\left| t \cdot \gtp^{\circ} \cap \setZ^d \right| = (-1)^d \cdot Ehr(\gtp; -t),
\]
where $d = \dim \gtp$.
This provides a method for computing the Ehrhart polynomial by
evaluating lattice-point counts at both positive dilates (regular GT-patterns)
and negative dilates (strict GT-patterns, via reciprocity).

For weight-restricted polytopes $\gtp_{\lambda/\mu,w}$,
the interior lattice points satisfy strict interlacing
together with the weight equalities (which are unchanged),
and their count is again governed by
\hyperref[ehrhartMacdonaldReciprocity]{Ehrhart--Macdonald reciprocity}.


\begin{example*}[Reciprocity]
Consider $\lambda = (5,3,0)$ with the triangular GT-pattern
\[
\begin{matrix}
5 && 3 && 0 \\
& a && b \\
&& c
\end{matrix}
\qquad
5 \geq a \geq 3,\; 3 \geq b \geq 0,\; a \geq c \geq b.
\]
This polytope has dimension $3$ and contains $\schurS_{(5,3)}(1,1,1) = 42$ lattice points.
For the interior, the strict conditions
$5 > a > 3$, $3 > b > 0$, $a > c > b$
force $a = 4$ and $b \in \{1, 2\}$:
\begin{itemize}
\item $b=1$: $c \in \{2, 3\}$;
\item $b=2$: $c = 3$;
\end{itemize}
giving $3$ interior lattice points.
The Ehrhart polynomial is
\[
Ehr(\gtp_{(5,3,0)};t) = \frac{(2t+1)(5t+2)(3t+1)}{2},
\]
and one verifies $(-1)^3 \cdot Ehr(\gtp_{(5,3,0)};-1) = -\frac{(-1)(-3)(-2)}{2} = 3$.
\end{example*}


\begin{example*}[Reciprocity in the Kostka case]
Consider instead the weight-restricted polytope
$\gtp_{(3,2,1),(2,2,2)}$, whose lattice points are
GT-patterns with top row $(3,2,1)$ and weight $(2,2,2)$:
\[
\begin{matrix}
3 && 2 && 1 \\
& a && b \\
&& c
\end{matrix}
\qquad
\begin{aligned}
& 3 \geq a \geq 2,\; 2 \geq b \geq 1,\; a \geq c \geq b, \\
& a + b = 4,\; c = 2.
\end{aligned}
\]
The weight conditions fix $c=2$ and $a+b=4$.
Combined with $3 \geq a \geq 2$ and $2 \geq b \geq 1$,
there are two solutions: $(a,b,c) \in \{(3,1,2),\; (2,2,2)\}$,
confirming $K_{(3,2,1),(2,2,2)} = 2$.
The strict conditions $3 > a > 2$, $2 > b > 1$, $a > c > b$
have no integer solutions, so there are no interior lattice points.
This reflects that the polytope, after intersecting with the weight hyperplanes,
is zero-dimensional.
\end{example*}



\section[gtLR]{Littlewood--Richardson coefficients}

The \hyperref[lrTableaux]{Littlewood--Richardson coefficient} $c^\lambda_{\mu\nu}$
counts skew semistandard Young tableaux of shape $\lambda/\mu$
and content $\nu$ whose reverse reading word
(read right to left across each row, from top to bottom)
is a \defin{lattice word}\label{latticeWord}:
at every position in the word, the number of occurrences of letter $i$
weakly exceeds the number of occurrences of letter $i+1$, for all $i$.

In terms of GT-patterns, a skew tableau of shape $\lambda/\mu$ and content $\nu$
corresponds to a chain of partitions
$\mu = \alpha^{(0)} \subseteq \alpha^{(1)} \subseteq \dotsb \subseteq \alpha^{(n)} = \lambda$,
where each $\alpha^{(k)}/\alpha^{(k-1)}$ is a horizontal strip of size $\nu_k$.
The quantity $\alpha^{(k)}_j - \alpha^{(k-1)}_j$ counts the cells with
letter $k$ in row $j$ of the tableau.


\subsection[gtLRInequalities]{Yamanouchi inequalities on GT-patterns}

The lattice word condition on the reverse reading word translates to the
\defin{Yamanouchi inequalities}\label{yamanouchiInequalities}, the following
linear inequalities on the GT-pattern entries.

\begin{proposition}
A skew semistandard Young tableau with GT-pattern
$\alpha^{(0)} \subseteq \alpha^{(1)} \subseteq \dotsb \subseteq \alpha^{(n)}$
is a Littlewood--Richardson tableau if and only if
\begin{equation*}
\sum_{l=1}^{j} \bigl(\alpha^{(k+1)}_l - \alpha^{(k)}_l\bigr)
\;\leq\;
\sum_{l=1}^{j-1} \bigl(\alpha^{(k)}_l - \alpha^{(k-1)}_l\bigr)
\end{equation*}
for all $k = 1, \dotsc, n-1$ and $j = 1, \dotsc, m$,
where $m$ is the number of parts of $\lambda$.
\end{proposition}

Setting $j=1$ gives $\alpha^{(k+1)}_1 \leq \alpha^{(k)}_1$ for $k \geq 1$,
which combined with the interlacing condition
$\alpha^{(k+1)}_1 \geq \alpha^{(k)}_1$
forces $\alpha^{(k+1)}_1 = \alpha^{(k)}_1$ for all $k \geq 1$.
This recovers the well-known fact that the first row of a
Littlewood--Richardson tableau is filled entirely with letter $1$.

These additional linear constraints, imposed on $\gtp_{\lambda/\mu,\nu}$,
cut out a polytope whose integer lattice points are in bijection with
\hyperref[lrTableaux]{Littlewood--Richardson tableaux}.  In fact, this gives the
description of \hyperref[bzPatterns]{Berenstein--Zelevinsky patterns}.


\begin{example*}[A Littlewood--Richardson case]
Consider $c^{(3,2,1)}_{(2,1),(2,1)}$.
The skew shape $(3,2,1)/(2,1)$ has three cells,
at positions $(1,3)$, $(2,2)$, and $(3,1)$:
\begin{figure}
\begin{ytableau}
\none & \none & {} \\
\none & {} \\
{}
\end{ytableau}
\end{figure}
Since each row and column has at most one cell,
every filling is semistandard.
With content $(2,1)$ (two $1$s and one $2$), there are three fillings:
\begin{figure}
\begin{ytableau}
\none & \none & 1 \\
\none & 1 \\
2
\end{ytableau}
%
\begin{ytableau}
\none & \none & 1 \\
\none & 2 \\
1
\end{ytableau}
%
\begin{ytableau}
\none & \none & 2 \\
\none & 1 \\
1
\end{ytableau}
\end{figure}
so $K_{(3,2,1)/(2,1), (2,1)} = 3$.
The corresponding GT-patterns have
$\alpha^{(0)} = (2,1,0)$, $\alpha^{(2)} = (3,2,1)$,
and the three possibilities for the middle row $\alpha^{(1)}$
are $(3,2,0)$, $(3,1,1)$, and $(2,2,1)$.

Applying the Yamanouchi inequality with $k=1$, $j=1$
gives $\alpha^{(2)}_1 - \alpha^{(1)}_1 \leq 0$:
\begin{itemize}
\item $\alpha^{(1)} = (3,2,0)$: $3 - 3 = 0 \leq 0$. Satisfied.
\item $\alpha^{(1)} = (3,1,1)$: $3 - 3 = 0 \leq 0$. Satisfied.
\item $\alpha^{(1)} = (2,2,1)$: $3 - 2 = 1 > 0$. Violated.
\end{itemize}
The third pattern is eliminated: it corresponds to the tableau
with reverse reading word $2, 1, 1$,
which fails the lattice word condition at the first letter.
The remaining inequalities (with $j=2,3$) are satisfied
for the first two patterns,
confirming $c^{(3,2,1)}_{(2,1),(2,1)} = 2$.
This agrees with the product decomposition
\[
\schurS_{(2,1)} \cdot \schurS_{(2,1)}
= \schurS_{(4,2)} + \schurS_{(4,1,1)} + \schurS_{(3,3)}
 + 2\,\schurS_{(3,2,1)} + \schurS_{(3,1,1,1)}
 + \schurS_{(2,2,2)} + \schurS_{(2,2,1,1)}.
\]
\end{example*}


\begin{example*}[A larger example]
For a larger example, consider $c^{(4,3,2)}_{(2,1),(3,2,1)}$,
where the skew shape $(4,3,2)/(2,1)$ has six cells and content $\nu = (3,2,1)$.
The GT-pattern has four rows:
\[
\alpha^{(0)} = (2,1,0) \subseteq \alpha^{(1)} \subseteq \alpha^{(2)} \subseteq \alpha^{(3)} = (4,3,2).
\]
\begin{itemize}
\item At step $0 \to 1$ (letter $1$, strip size $3$), there are three extensions:
\[
\alpha^{(1)} \in \{(4,2,0),\; (4,1,1),\; (3,2,1)\}.
\]
\item At step $1 \to 2$ (letter $2$, strip size $2$),
the Yamanouchi condition $\alpha^{(2)}_1 \leq \alpha^{(1)}_1$ eliminates
extensions that place letter $2$ in the first row.
\item At step $2 \to 3$ (letter $3$, strip size $1$),
the Yamanouchi condition is checked against the cumulative
row differences from the previous step.
\end{itemize}
After applying all constraints, one finds
$c^{(4,3,2)}_{(2,1),(3,2,1)} = 2$,
compared to the skew Kostka number
$K_{(4,3,2)/(2,1),(3,2,1)} = 6$.
The Yamanouchi conditions eliminate four of the six GT-patterns.
\end{example*}




\section[alternatingSignMatrix]{Monotone triangles and alternating sign matrices}

\defin{Monotone triangles}\label{monotoneTriangle} are special triangular GT-patterns with additional
constraints.
First, the top row is $(n,n-1,\dotsc,1)$ and all rows must be \emph{strictly} decreasing.

Monotone triangles of size $n$ are in bijection with $n \times n$ \emph{alternating sign matrices}.
\defin{Alternating sign matrices}\label{alternatingSignMatrixDefinition}
are matrices with entries in $\{-1,0,1\}$
such that in each row and column, the nonzero elements alternate in sign.
Furthermore, the sum of the entries in each row (and column) is $1$.

\begin{example}
Consider the following $5\times 5$ alternating sign matrix, where we omit the zeros,
and $-1$ is written as $\color{blue}{\overline{1}}$.
The rows are accumulated from the bottom, so that row $k$ from the bottom
is the sum of the first $k$ rows.
\[
\begin{bmatrix}
 &1& & & \\
1&\color{blue}{\overline{1}}& &1& \\
 &1& &\color{blue}{\overline{1}}&1\\
 & &1& & \\
 & & &1&
\end{bmatrix}
\to
\begin{bmatrix}
1&1&1&1&1 \\
1& &1&1&1 \\
 &1&1& &1 \\
 & &1&1&  \\
 & & &1&
\end{bmatrix}
\]
For each row, we then record the columns that contain a $1$
and write the column indices in a decreasing fashion.
The result is a monotone triangle.
\[
\begin{matrix}
5&&4&&3&&2&&1 \\
&5&&4&&3&&1 \\
&&5&&3&&2 \\
&&&4&&3 \\
&&&&4& \\
\end{matrix}
\]
\end{example}

\name{Doron Zeilberger} \cite{Zeilberger1995} proved that the number of
monotone triangles of size $n$, or equivalently, number of $n\times n$
\hyperref[alternatingSignMatrix]{alternating sign matrices} is given by
\[
\prod_{j=0}^{n-1}\frac{(3j+1)!}{(n+j)!}.
\]

A generalization of this result is given in \cite{Fischer2006},
where difference operators are used in a formula to
count the number of monotone triangles with prescribed top row $\lambda$.


\section[gtApplications]{Applications}

\subsection[gtSchur]{Schur polynomials}

The close connection between GT-patterns and \hyperref[schurS]{Schur polynomials}
is quite apparent. The Schur polynomial $\schurS_\lambda(x_1,\dotsc,x_n)$ can be expressed
as a sum over $GT(\lambda)$, the set of triangular GT-patterns with top row
equal to $\lambda$.
\[
\schurS_\lambda(x_1,\dotsc,x_n) = \sum_{G \in GT(\lambda) } \xvec^{w(G)}
\]
where $w$ depends on the row-sums of the pattern $G$.

\subsection[gtHL]{Hall--Littlewood polynomials}

The identity for Schur functions can be
generalized to \hyperref[hallLittlewoodP]{Hall--Littlewood polynomials},
via a weighted version of Brion's theorem.
\begin{theorem}[See \cite{FeiginMakhlin2016}]
Let $\lambda$ be a partition. Then
\begin{equation*}
 \hallLittlewoodP_\lambda(\xvec;t) = \sum_{G \in GT(\lambda) } p_G(t) \xvec^{w(G)}
\end{equation*}
where $p_G(t)$ is a certain statistic that depends on $G$.
\end{theorem}
This theorem can also be proved via other means, see \cite{Macdonald1995}.



\subsection[gtKey]{Key polynomials}

The \hyperref[key]{key polynomials} generalize the \hyperref[schurS]{Schur polynomials}.
In \cite{KiritchenkoSmirnovTimorin2010}, a formula for key polynomials is proved,
where the sum runs over lattice points in a union of certain faces of $GT(\lambda)$,
see \hyperref[keyDefinitionGT]{key polynomials from GT-patterns} for details.


\subsection[gtSchubert]{Schubert polynomials}

\name{Ricky Ini Liu}, \name{Karola Mészáros}, and \name{Avery St. Dizier}
prove that many \hyperref[schubertGTFormula]{Schubert polynomials} can be
obtained by projecting integer point transforms of Minkowski sums of
GT-polytopes \cite{LiuMeszarosDizier2019}.  More precisely, if $D$ is a
column-convex diagram, then the character of the flagged Schur module of $D$
is obtained from the integer point transform of a convex polytope
$P_D=GT(\lambda^{(1)})+\dotsb+GT(\lambda^{(n)})$ by a monomial
specialization.  When $D$ is the Rothe diagram of a column-convex permutation,
this character is the corresponding Schubert polynomial.
