\metatitle{Symmetricfunctions.com}
\metadescription{Landing page with links.}

\begin{cssvars}
    --col-main-start: 2;
    --col-main-end:   8;
    --toc-display: none;
    --header-border: none;
\end{cssvars}


% TODO: Search bar here?


\begin{topicsection}{Preliminaries}

\topiccard{preliminaries}{Partitions and tableaux}{Basic definitions, notation and conventions for partitions, Young diagrams and Young tableaux.}

\topiccard{permutationMatrix}{Permutations}{Permutation matrices, Rothe diagrams, Lehmer codes, reduced words, Bruhat order.}

\topiccard{permutationPattern}{Permutation patterns}{Pattern containment, avoidance and avoidance classes.}

\topiccard{namedPermutations}{Permutation families}{Named families of permutations: alternating, vexillary, Baxter, boolean, and more.}

\topiccard{notationAffinePermutations}{Permutation generalizations}{Affine permutations, signed permutations, colored permutations, Stirling permutations.}

\topiccard{qAnalogs}{$q$-analogs}{Classical $q$-analogs, $q$-binomial coefficients and related identities.}

\topiccard{representationTheory}{Representation theory}{General representation-theoretic background used throughout the site.}

\topiccard{kTheory}{K-theory}{Grothendieck groups, vector bundles, coherent sheaves, and K-theoretic Schubert calculus.}

\topiccard{lagrangeInversion}{Lagrange inversion}{The Lagrange inversion formula and some standard combinatorial applications.}

\topiccard{transferMatrixMethod}{Transfer matrix method}{Finite-state enumeration, rational generating functions, and transfer matrices.}

\topiccard{lgvLemma}{Lindström--Gessel--Viennot lemma}{Non-intersecting lattice paths and determinantal formulas via the LGV lemma.}

\end{topicsection}


\begin{topicsection}{Symmetric functions}

\topiccard{polynomialIndex}{List of symmetric functions and generalizations}{Index of all functions.}

\topiccard{plethysm}{Plethysm}{Plethysm and plethystic operations.}

\topiccard{symmetricFunctionsSpace}{Symmetric function operators}{Fock space, Hall inner product, standard involution, adjoint operators, and Hopf structure.}

\topiccard{schurPositivity}{Schur positivity}{Ways to prove Schur positivity.}

\topiccard{diagonalHarmonics}{Diagonal harmonics}{Diagonal harmonics, parking functions and related graded $S_n$-modules.}

\topiccard{latticeModel}{Vertex models and Yang--Baxter}{Vertex models, the Yang--Baxter equation and their applications in algebraic combinatorics.}

\end{topicsection}



\begin{topicsection}{Tableaux and crystal combinatorics}

\topiccard{rsk}{Robinson--Schensted--Knuth correspondence}{The RSK correspondence and its combinatorial and representation-theoretic consequences.}

\topiccard{tableauOperations}{Operations on Young tableaux}{Classical operations on tableaux such as Bender--Knuth moves, promotion, evacuation and jeu-de-taquin.}

\topiccard{crystals}{Crystals}{Crystals, crystal graphs and their connections to tableaux and representation theory.}

\end{topicsection}



\begin{topicsection}{Cyclic sieving}

\topiccard{cyclicSieving}{The cyclic sieving phenomenon}{Definitions, examples and general methods related to the cyclic sieving phenomenon (CSP).}

\topiccard{cspWord}{CSP on words}{Cyclic sieving on words and permutations.}

\topiccard{cspPath}{CSP on paths}{Cyclic sieving on lattice paths.}

\topiccard{cspCatalan}{CSP on Catalan objects}{Cyclic sieving on Catalan objects.}

\topiccard{cspMatch}{CSP on matchings}{Cyclic sieving on matchings and crossings.}

\topiccard{cspTableau}{CSP on tableaux}{Cyclic sieving on tableaux and fillings.}

\topiccard{cspMisc}{Miscellaneous}{Cyclic sieving on various sets.}

\end{topicsection}





\begin{topicsection}{Combinatorial objects}


\topiccard{polytope}{Polytopes}{Polytopes, Ehrhart functions and $h^*$-vectors.}

\topiccard{gtpatterns}{Gelfand--Tsetlin polytopes}{Gelfand--Tsetlin patterns, their polytopes and connections to representation theory.}

\topiccard{planePartitions}{Plane partitions}{Plane partitions and their enumerative and representation-theoretic properties.}

\topiccard{parkingFunctions}{Parking functions}{Parking functions, bijections and generalizations.}

\topiccard{matroid}{Matroids}{Matroids, their basic properties and selected examples from algebraic combinatorics.}

\topiccard{rootSystem}{Root systems}{Root systems, Weyl groups and their combinatorial features.}

\end{topicsection}



\begin{topicsection}{Structure coefficients}

\topiccard{kostantPartitionFunction}{Kostant partition function}{The Kostant partition function and its role in weight multiplicities.}

\topiccard{kostkaFoulkes}{Kostka--Foulkes polynomials}{Kostka and Kostka--Foulkes polynomials and their combinatorial interpretations.}

\topiccard{riggedConfigurations}{Rigged configurations}{Rigged configurations and their connections to Kostka-type polynomials and crystals.}

\topiccard{borderStripTableaux}{Border-strip and ribbon tableaux}{Border-strip tableaux, ribbon tableaux and their roles in character formulas.}

\topiccard{littlewoodRichardsonModels}{Littlewood--Richardson coefficients}{Models for Littlewood--Richardson coefficients and their combinatorial interpretations.}

\topiccard{schubertCalculus}{Schubert calculus}{Grassmannians, Schubert classes, Pieri and Giambelli computations, and enumerative examples.}

\end{topicsection}



\begin{topicsection}{Real-rooted polynomials}

\topiccard{polynomialRootIntro}{Unimodality and real-rootedness}{Unimodality, log-concavity, real-rootedness, totally non-negative matrices and Hadamard products.}

\topiccard{interlacingPolynomials}{Interlacing polynomials}{Interlacing roots, stable polynomials, Sturm sequences, and linear operators preserving real-rootedness.}

\topiccard{stablePolynomial}{Stable polynomials}{Multivariate stability, same-phase stability, the half-plane property, and connections to real-rootedness.}

\topiccard{lorentzianPolynomial}{Lorentzian polynomials}{Lorentzian polynomials, ultralog-concavity, and the symmetric decomposition of $h^*$-polynomials.}

\topiccard{gammaPositivity}{Gamma-positive polynomials}{Gamma-positivity, palindromic polynomials, the Foata--Strehl action, and Gal's conjecture.}

\topiccard{realRootedCatalan}{Catalan families}{Narayana polynomials, ballot polynomials, Motzkin polynomials, and other Catalan-type real-rooted families.}
\topiccard{realRootedWords}{Interlacing examples}{Eulerian polynomials, type $B$ Eulerian, derangement polynomials, peak polynomials, and other permutation statistics.}
\topiccard{realRootedTableaux}{Tableaux and polytopes}{Descent polynomials of standard Young tableaux, Ehrhart polynomials, and set partition statistics.}
\topiccard{realRootedMatching}{Graph polynomials}{Matching polynomials, independence polynomials of claw-free graphs, rook polynomials, and Chow polynomials.}



\end{topicsection}


\begin{topicsection}{Miscellaneous}

\topiccard{unitIntervalAO}{Acyclic orientations and Touchard--Riordan}{Touchard--Riordan-type polynomials.}

\topiccard{resources}{More reading and links}{Additional reading and links related to algebraic combinatorics.}

\topiccard{personalMinorProblems}{Minor research problems}{Minor problems and conjectures related to my own research.}

\end{topicsection}
