\metatitle{Lah symmetric functions}
\metadescription{An introduction to Lah symmetric functions, their combinatorial definition via forests of increasing ordered trees, and their positivity in the elementary symmetric function basis.}
\metakeywords{Lah symmetric functions,monomial-positive,e-positive}

\section[lah]{Lah symmetric functions}

\begin{polydata}{lah}
  Name   & Lah symmetric functions \\
  Space    & Sym \\
  Basis    & False \\
  Rating   & 1 \\
  Bib      & PetreolleSokal2019x \\
  Year     & 2019 \\
  Symbol   & $\lah_{n,k}(\xvec)$ \\
  Keywords & monomial-positive, e-positive \\
  Category & Other \\
  PositiveIn & elementaryE | PetreolleSokal2019x \\
\end{polydata}



The \emph{Lah symmetric functions} were introduced in \cite{PetreolleSokal2019x}.
They are indexed by two integers, $n \geq k \geq 1$.
The combinatorial description is as follows.

We first need the notion of an \emph{increasing ordered tree} on vertex set $[n]$.
Each node has label smaller than its children (so vertex labeled $1$ is a root),
and the children are ordered linearly.
The numbers of such trees on $n$ nodes are $1$, $1$, $3$, $15$, $105,\dotsc$,
see \oeis{A001147}.

An \emph{unordered forest of increasing ordered trees}
is a forest on vertex set $[n]$, where each component is an increasing ordered tree.
The number of such forests on $n$ vertices with $k$ trees is given by \oeis{A001497}:
\begin{array}{lllll}
\toprule
 n \backslash k & \textbf{1} & \textbf{2} & \textbf{3} & \textbf{4} & \textbf{5} \\
\midrule
\textbf{1}& 1 \\
\textbf{2}& 1 & 1 \\
\textbf{3}& 3 & 3 & 1\\
\textbf{4}& 15 & 15 & 6 & 1  \\
\textbf{5}& 105 & 105 & 45 & 10 & 1 \\
\bottomrule
\end{array}
A closed-form formula is given by $\frac{(2n-k-1)!}{2^{n-k}(n-k)!(k-1)!}$,
see \cite[Eq.~7.11]{PetreolleSokal2019x}.

The \defin{Lah symmetric function} is then defined as
\[
 \lah_{n,k}(\xvec) \coloneqq \sum_{F \in \mathrm{Forests}(n,k)} \prod_{j=1}^n
\elementaryE_{c(j)}(\xvec)
\]
where $c(j)$ is the number of children of vertex $j$,
and the sum is taken over all unordered forests of increasing ordered trees,
with $n$ vertices and exactly $k$ trees.
By definition, the $\lah_{n,k}(\xvec)$ are positive
in the \hyperref[elementaryE]{elementary symmetric function basis}.
This also implies \hyperref[schurPositivity]{Schur positivity}, since each
$\elementaryE_d$ is the Schur function $\schurS_{1^d}$ and products of Schur
functions are Schur-positive.

\emph{Note:} the $\lah_{n,k}(\xvec)$ is denoted $L^{(\infty)+}_{n,k}(X)$
and $L^{(\infty)-}_{n,k}(X) = \omega \lah_{n,k}(\xvec)$ in 
\cite{PetreolleSokal2019x}.

\begin{example*}[Lah symmetric functions for $n=4$]
% Related Rust: sym-poly/sym/src/lah.rs and
% sym-poly/sym/examples/lah_site_example.rs
We have that
\begin{align*}
\lah_{4,1} &= 6 \elementaryE_3 + 8 \elementaryE_{21} + \elementaryE_{111} =
\monomial_{3} + 11\monomial_{21}+36\monomial_{111}, \\
\lah_{4,2} &= 8 \elementaryE_2 + 7 \elementaryE_{11} =
7\monomial_{2} + 22\monomial_{11}, \\
\lah_{4,3} &= 6 \elementaryE_{1}= 6\monomial_{1}, \\
\lah_{4,4} &= 1.
\end{align*}
For example, the three terms in $\lah_{4,1}$ count increasing ordered trees
whose child-count partition is respectively $(3)$, $(2,1)$, and $(1,1,1)$.
\end{example*}

\subsection[lahForestBasis]{The connected Lah forest basis}

There is a natural multiplicative basis obtained by taking connected Lah
forests as generators.  Define
\[
  \ell_d(\xvec) \coloneqq \lah_{d+1,1}(\xvec)
  \qquad\text{and}\qquad
  \ell_\lambda(\xvec) \coloneqq
  \ell_{\lambda_1}(\xvec)\ell_{\lambda_2}(\xvec)\dotsm .
\]
Since
\[
  \ell_d = d!\elementaryE_d
  + \text{terms involving products of smaller elementary functions},
\]
the functions $\ell_\lambda$ form a basis of $\mathrm{Sym}_{\setQ}$.
The multiplication rule is simply
\[
  \ell_\lambda \ell_\mu = \ell_{\lambda \cup \mu}.
\]

The ordinary Lah symmetric functions are positive in this basis.  If
$m_a(\lambda)$ denotes the multiplicity of the part $a$ in $\lambda$, then
\[
  \lah_{n,k} =
  \sum_{\substack{\lambda \vdash n-k \\ \ell(\lambda) \leq k}}
  \frac{n!}
  {(k-\ell(\lambda))!\prod_i(\lambda_i+1)!
   \prod_{a\geq 1} m_a(\lambda)!}
  \ell_\lambda .
\]
Indeed, once the non-singleton component sizes are fixed, the coefficient
counts the ways to choose their vertex sets inside $[n]$; each component then
contributes one connected Lah generator after standardizing its labels.

\begin{example*}[Small connected Lah-basis expansions]
% Related Rust: sym-poly/sym/src/lah.rs and
% sym-poly/sym/examples/lah_forest_basis_site_example.rs
In degree at most $3$, the connected generators begin
\[
  \ell_1=\elementaryE_1,\qquad
  \ell_2=\elementaryE_{11}+2\elementaryE_2,\qquad
  \ell_3=\elementaryE_{111}+8\elementaryE_{21}+6\elementaryE_3.
\]
For $n=4$, the Lah symmetric functions expand as
\[
  \lah_{4,1}=\ell_3,\qquad
  \lah_{4,2}=4\ell_2+3\ell_{11},\qquad
  \lah_{4,3}=6\ell_1,\qquad
  \lah_{4,4}=1.
\]
\end{example*}

\begin{remark*}
The connected Lah forest basis is not a positive basis for all elementary
positive symmetric functions.  In particular, the complete graph $K_2$ is a
\hyperref[chromaticQuasisymmetricUnitIntervalGraph]{unit interval graph}, but
its chromatic symmetric function satisfies
\[
  \chrom_{K_2}(\xvec) = 2\elementaryE_2 = \ell_2-\ell_{11}.
\]
Thus unit interval graph chromatic symmetric functions are not all positive
in the connected Lah forest basis.
\end{remark*}
