\metatitle{Macdonald E polynomials}
\metadescription{An introduction to Macdonald E polynomials and their $q$-Whittaker specialization, including definitions, properties, and connections to fillings and key polynomials.}

\section[macdonaldE]{Macdonald E polynomials}

\begin{polydata}{macdonaldE}
  Name     & Macdonald E polynomials \\
  Space    & All \\
  Basis    & True \\
  Rating   & 2 \\
  Bib      & Opdam1995\\
  Year     & 1995\\
  Symbol   & $\macdonaldE_\mu(\xvec;q,t)$ \\
  Keywords & fillings \\
  Category & Schur \\
  SpecializesTo & macdonaldEspec | Opdam1995 | map=t=0 \\
  SpecializesTo & atom | Mason2009 | map=q,t -> infinity \\
  SpecializesTo & macdonaldP | Macdonald1987 | scope=symmetric case \\
  SpecializesTo & key | RamYip2011 | map=q=t=0 \\
  StableLimit & macdonaldP | Macdonald1987 | scope=symmetric limit \\
\end{polydata}

The non-symmetric Macdonald polynomials were introduced in \cite{Opdam1995},
\cite{Macdonald1996} and \cite{Cherednik1995nonsymmetric}.
The non-symmetric Macdonald polynomials are closely related with \emph{affine root systems},
and the \emph{double affine Hecke algebra}. Their definition is rather indirect,
and does not give an efficient way of computing these non-symmetric polynomials.

In \cite{HaglundHaimanLoehr2008}, \name{James Haglund}, \name{Mark Haiman} and \name{Nicholas Loehr} 
found an explicit formula for the non-symmetric Macdonald polynomials in type $A$,
as a sum over \emph{non-attacking fillings}.
This model is the basis for the \hyperref[macdonaldEperm]{permuted basement Macdonald polynomials},
and we refer to that page for definitions.
The specialization $q=t=0$ gives \hyperref[key]{key polynomials}, while the
limit $q,t\to\infty$ gives \hyperref[atom]{Demazure atoms}
\cite{Mason2009}.

Later, an alternative combinatorial formula using alcove walks was
proved by \name{Arun Ram} and \name{Martha Yip} \cite{RamYip2011}, which works for all Lie types.
At $q=t=0$, their model reduces to the \emph{Littelmann path model}.

In \cite{BorodinWheeler2019}, an integrable vertex model is used to 
give an alternative formula for non-symmetric Macdonald polynomials.
\todo{Later figures/examples pass: explain the Borodin--Wheeler vertex model
with one small lattice configuration and its monomial weight.  Use a figure
checked against the paper or generated from code rather than a hand-drawn
guess.}

See the book \cite{qtCatalanBook} for more background on the type $A$
non-symmetric Macdonald polynomials.
My personal research with non-symmetric Macdonald polynomials and further generalizations 
can be found in 
\cite{Alexandersson2015gbMacdonald,AlexanderssonSawhney2017,AlexanderssonSawhney2019}.

\name{Charles F. Dunkl} evaluates families of nonsymmetric Macdonald
superpolynomials at special points \cite{Dunkl2021SuperMacdonald}.  These
superpolynomials take values in Hecke-algebra modules built from
anti-commuting variables, and the evaluations involve $(q,t)$-hook products.

\name[Milo Bechtloff Weising]{Milo Bechtloff Weising} constructs stable-limit
nonsymmetric Macdonald functions in type $A$ \cite{Weising2023x}.  They form a
simultaneous eigenbasis for the limit Cherednik operators in Ion--Wu's
stable-limit double affine Hecke algebra representation, connecting stable
DAHA theory with the Carlsson--Mellit double Dyck path algebra.
\name[Milo Bechtloff Weising]{Milo Bechtloff Weising} and
\name{Alexander E. Black} prove that the supports of nonsymmetric Macdonald
polynomials are $M$-convex \cite{WeisingBlack2025x}.  This proves the
saturated Newton polytope property conjectured by Monical--Tokcan--Yong, and
also gives $M$-convex supports for affine Demazure characters of type
$\mathrm{GL}$.

A model using multiline queues is introduced in \cite{CorteelMandelshtamWilliams2018x},
and further explored in \cite{CorteelMandelshtamMasonWilliams2019x}.
\todo{Later figures/examples pass: add a small multiline-queue diagram for a
type $A$ nonsymmetric Macdonald polynomial, including the weight contribution
and a note on the ASEP/TASEP specialization when reliable.  Generate or verify
the queue data with Rust code if possible.}

\name{Olya Mandelshtam} and \name{Jerónimo Valencia-Porras} use twisted
multiline queues to model Macdonald polynomials at $t=0$
\cite{MandelshtamValenciaPorras2024x}.  This adds another multiline-queue
realization to the non-symmetric Macdonald polynomial circle.

\name{Arvind Ayyer}, \name{James Martin}, and \name{Lauren Williams} connect
ASEP polynomials and nonsymmetric Macdonald polynomials at $q=1$ to the
stationary distribution of an inhomogeneous multispecies $t$-PushTASEP on a
ring \cite{AyyerMartinWilliams2024x}.  The normalizing partition function is
the corresponding Macdonald $P$ polynomial at $q=1$, and the proof uses
multiline diagrams.

\name{Evgeny Feigin}, \name{Anton Khoroshkin}, and
\name{Ievgen Makedonskyi} introduce parasymmetric Macdonald polynomials,
interpolating between symmetric and nonsymmetric Macdonald polynomials and
depending on a parabolic subalgebra \cite{FeiginKhoroshkinMakedonskyi2023x}.
They categorify specializations of these polynomials using cyclic modules for
parahoric subalgebras of affine Kac--Moody Lie algebras.


\subsection[macdonaldEMonks]{Monk's rule}

\name{Wendy Baratta} \cite{Baratta2009,Baratta2011} gives Monk type
rules (terminology from \hyperref[schubertMonksRule]{Schubert calculus}) for products of the form 
\[
x_j \macdonaldE_\mu(x_1,\dotsc,x_n;q,t), \quad \elementaryE_1(\xvec) \cdot \macdonaldE_\mu(x_1,\dotsc,x_n;q,t), \quad 
 \elementaryE_{r}(\xvec) \cdot\macdonaldE_\mu(x_1,\dotsc,x_n;q,t),
\]
expanded again in the $\{\macdonaldE_\alpha\}$ basis. 
Baratta uses interpolation Macdonald polynomials in the proof.

In \cite{HalversonRam2022x}, the authors prove Monk type formulas for 
\[
x_j \macdonaldE_\mu, \qquad (x_1+\dotsb + x_j) \macdonaldE_\mu, \qquad 
x_j^{-1} \macdonaldE_\mu, \qquad (x_j^{-1}+\dotsb + x^{-1}_j) \macdonaldE_\mu.
\]
Halverson--Ram uses a different method based on intertwiners.

\subsection[macdonaldERecurrences]{Recurrence relations}

The HHL filling formula is often proved by checking the
\defin{Knop--Sahi recurrence}.  There are two common indexing
orientations.  Haglund--Haiman--Loehr, and the skyline convention used by
Moura--Mandelshtam, index a composition by column heights.  This website
follows Alexandersson's English-diagram convention; when $\sigma=\omega_0$,
our index $\alpha$ corresponds to the HHL index $\rev(\alpha)$.
Thus the cyclic recurrence is reversed compared with the HHL display.
In the website convention, let $\alpha=(\alpha_1,\dotsc,\alpha_n)$ and
\[
\widehat{\alpha}\coloneqq(\alpha_2,\dotsc,\alpha_n,\alpha_1+1),\qquad
\Psi f\coloneqq x_1f(x_2,\dotsc,x_n,q^{-1}x_1).
\]
Then
\[
\macdonaldE_{\widehat{\alpha}}(\xvec;q,t)
  = q^{\alpha_1}\Psi\macdonaldE_\alpha(\xvec;q,t)
\]
in Alexandersson's convention \cite[Sec.~3]{Alexandersson2015gbMacdonald}.
Equivalently, after setting $\mu=\rev(\alpha)$, this is the HHL recurrence
\cite[Sec.~2.1]{HaglundHaimanLoehr2008}.

In HHL skyline indexing, if $\mu_i>\mu_{i+1}$ and
$u=(i,\mu_{i+1}+1)$ is the corresponding cell in the column diagram, then
\[
\macdonaldE_{s_i\mu}(\xvec;q,t)
  =
  \left(
    T_i+
    \frac{1-t}{1-q^{\leg(u)+1}t^{\arm(u)}}
  \right)
  \macdonaldE_\mu(\xvec;q,t),
\]
where $T_i$ is the affine Hecke operator in the HHL convention.  Translating
this adjacent-part recurrence to the website convention reverses the shape
positions.  Together with $\macdonaldE_{(0,\dotsc,0)}=1$, the cyclic
recurrence and the special cases of the adjacent recurrence uniquely
determine all non-symmetric Macdonald polynomials
\cite[Sec.~2.1]{HaglundHaimanLoehr2008}.

For \hyperref[macdonaldEperm]{permuted-basement Macdonald polynomials},
Demazure--Lusztig operators give a second recursive viewpoint.  Let
% Related Rust: sym-poly/multipoly/src/nonsymmetric_macdonald.rs checks the
% q=0 operator recursion against the non-attacking filling formula.
\[
\widetilde{\pi}_i\coloneqq (1-t)\pi_i+t s_i,\qquad
\widetilde{\theta}_i\coloneqq (1-t)\theta_i+t s_i,
\]
where $\pi_i$ and $\theta_i$ are the usual Demazure character and atom
operators.  If $\gamma_i$ is the length of the row whose basement label is
$i$, then Alexandersson's basement-permuting formulas state
\cite[Prop.~15]{Alexandersson2015gbMacdonald} that
\[
\widetilde{\theta}_i\macdonaldE^\sigma_\alpha
 =
\begin{cases}
t\macdonaldE^{\sigma s_i}_\alpha, & \text{if } \ell(\sigma s_i)<\ell(\sigma)
   \text{ and } \gamma_i\leq\gamma_{i+1},\\
\macdonaldE^{\sigma s_i}_\alpha, & \text{if } \ell(\sigma s_i)<\ell(\sigma)
   \text{ and } \gamma_i>\gamma_{i+1},
\end{cases}
\]
and
\[
\widetilde{\pi}_i\macdonaldE^\sigma_\alpha
 =
\begin{cases}
t\macdonaldE^{\sigma s_i}_\alpha, & \text{if } \ell(\sigma s_i)>\ell(\sigma)
   \text{ and } \gamma_i<\gamma_{i+1},\\
\macdonaldE^{\sigma s_i}_\alpha, & \text{if } \ell(\sigma s_i)>\ell(\sigma)
   \text{ and } \gamma_i\geq\gamma_{i+1}.
\end{cases}
\]

There are also shape-permuting versions.  For example, if
$\alpha_j<\alpha_{j+1}$ and the adjacent basement labels are
$\sigma_j=i+1$, $\sigma_{j+1}=i$, then
\[
\macdonaldE^\sigma_{s_j\cdot\alpha}
  =
  \left(
    \widetilde{\theta}_i+
    \frac{1-t}{1-q^{\leg(u)+1}t^{\arm(u)}}
  \right)\macdonaldE^\sigma_\alpha,
\]
where $u=(j+1,\alpha_j+1)$ is computed in the diagram of
$\alpha$ \cite[Prop.~17]{Alexandersson2015gbMacdonald}.
The inverse case $\alpha_j>\alpha_{j+1}$ is given in
\cite[Prop.~18]{Alexandersson2015gbMacdonald}.  The more recent
\hyperref[macdonaldEPermShape]{shape-changing identities} of
Moura--Mandelshtam remove the adjacent-basement restriction and give a
straightening rule in the full permuted-basement family.


\section[macdonaldEspec]{Non-symmetric q-Whittaker polynomials}

\begin{polydata}{macdonaldEspec}
  Name     & Non-symmetric q-Whittaker polynomials \\
  Space    & All \\
  Basis    & True \\
  Rating   & 2 \\
  Bib      & Opdam1995\\
  Year     & 1995\\
  Symbol   & $\macdonaldE_\mu(\xvec;q,0)$ \\
  Keywords & fillings, key-positive \\
  Category & Schur \\
  Contains & qWhittaker | Opdam1995 | scope=symmetric/partition case \\
  SpecializesTo & nonSymmetricElementary | AlexanderssonSawhney2019 | map=q=1 \\
  PositiveIn & key | AlexanderssonSawhney2017,Assaf2018Kostka,AssafGonzalez2018 | semiring=NN[q] \\
  PositiveIn & atom | MandelshtamNiergarthSingh2026x | semiring=NN[q]; proof=Cor. 4.7 \\
\end{polydata}

The \defin{non-symmetric $q$-Whittaker polynomials} are obtained by letting $t=0$ in $\macdonaldE_\mu(\xvec;q,t)$.
The result can be thought of as a non-symmetric analogue of the \hyperref[hallLittlewoodT]{transformed Hall--Littlewood polynomials},
and generalizes the \hyperref[qWhittaker]{$q$-Whittaker functions}.
We keep this family distinct from the
\hyperref[hallLittlewoodH]{modified Hall--Littlewood polynomials}: the former
comes from the $t=0$ Macdonald specialization, while the latter is the $q=0$
specialization of \hyperref[macdonaldH]{modified Macdonald polynomials}.
There are nevertheless direct comparisons between their filling models; see
\cite{Ratheesh2024x,Bhattacharya2025}.

The polynomials $\macdonaldE_\mu(\xvec;q,0)$ expand positively in \hyperref[key]{key polynomials},
and the coefficients are \hyperref[kostkaFoulkes]{Kostka--Foulkes polynomials},
see \cite{AlexanderssonSawhney2017,Assaf2018Kostka,AssafGonzalez2018}.
\name{Sami Assaf} and \name{Nicolle González} construct an affine Demazure
crystal on semistandard key tabloids whose characters realize these
specialized nonsymmetric Macdonald polynomials \cite{AssafGonzalez2021}.
Removing the affine edges recovers their finite Demazure crystals.

Models using \emph{quantum alcove walks} and \emph{quantum Lakshmibai--Seshadri path} 
are considered in \cite{LenartNaitoSagakiSchillingShimozono2017}.


\subsection[nonSymmetricElementary]{Non-symmetric elementary polynomials}

\begin{polydata}{nonSymmetricElementary}
  Name     & Non-symmetric elementary polynomials \\
  Space    & All \\
  Basis    & True \\
  Rating   & 1 \\
  Bib      & AlexanderssonSawhney2019 \\
  Year     & 2019 \\
  Symbol   & $\macdonaldE_\lambda(\xvec;1,0)$ \\
  Keywords & fillings, key-positive \\
  Category & Elementary \\
  Generalizes & elementaryE | AlexanderssonSawhney2019 | scope=partition shapes \\
  PositiveIn & key | Assaf2018Kostka,AlexanderssonSawhney2019 | proof=Kostka numbers \\
\end{polydata}

The \defin{non-symmetric elementary polynomials} are the specialization
$\macdonaldE_\lambda(\xvec;1,0)$, where $\lambda$ ranges over weak
compositions.  For a fixed number of variables, these polynomials form a basis
of the polynomial ring and naturally extend the ordinary
\hyperref[elementaryE]{elementary symmetric functions}
\cite[Sec.~4.1]{AlexanderssonSawhney2019}.
More precisely, if $\lambda$ is a partition, then the corresponding
permuted-basement polynomial $\macdonaldE^\sigma_\lambda(\xvec;1,t)$ is
independent of $\sigma$ and $t$, and is equal to
$\elementaryE_{\lambda'}(\xvec)$ \cite[Thm.~18]{AlexanderssonSawhney2019}.

They expand positively in \hyperref[key]{key polynomials}, with coefficients
given by ordinary Kostka numbers \cite{Assaf2018Kostka}.  This is the
$q=1$ specialization of the key-positive expansion of
$\macdonaldE_\lambda(\xvec;q,0)$ by Kostka--Foulkes polynomials.
