\metatitle{Families of permutations}
\metadescription{Named families of permutations: pattern-avoidance classes, alternating permutations, derangements, and more.}


\section[namedPermutations]{Named families of permutations}

See the \hyperref[permutationPattern]{permutation patterns page} 
for background on pattern containment and avoidance.

\begin{itemize}

\item \defin{Up-down}\label{permUpDown} ---  in one-line notation, we have $\pi_1 \lt \pi_2 \gt \pi_3 \lt \pi_4 \dotsb$.
\defin{Down-up} permutations are of the form $\pi_1 \gt \pi_2 \lt \pi_3 \gt \pi_4 \dotsb$.
Enumerated by the Euler numbers \oeis{A000111}.

\item \defin{André} --- Either up-down or down-up, \oeis{A001250}.

\item \defin{Derangements} --- permutations with no fixed point, \oeis{A000166}.

\item \defin{Dominant} --- 132-avoiding. Enumerated by the Catalan numbers, \oeis{A000108}.

\item \defin{Baxter} --- avoids $2-14-3$ and $3-14-2$, see \url{https://doi.org/10.1016/S0012-365X(97)00112-X}.

\item \defin{Boolean} --- 123-avoiding and 3412-avoiding, see \cite{Tenner2007,GunawanPanRussellTenner2022x}.
A permutation is boolean iff the order ideal it defines in the \hyperref[bruhatOrder]{Bruhat order}, is a Boolean lattice, see \cite[Thm. 4.3]{Tenner2007}.
Enumerated by odd-indexed Fibonacci numbers, \oeis{A001519}.

\item \defin{Fireworks} --- Same as 3-12 avoiding, see
\cite{ChouSetiabrata2025x}. This means that the initial terms of decreasing
runs are increasing.

\item \defin{Fully commutative} --- 321-avoiding, see \cite{BilleyJockuschStanley1993,GunawanPanRussellTenner2022x}.
Enumerated by the Catalan numbers, \oeis{A000108}.

\item A \defin{simsun permutation} is a permutation such that restricting to entries $1,2,\dotsc,k$,
it does not have a double descent, i.e., $\pi(i)\gt \pi(i+1) \gt \pi(i+2)$, see \cite{MaYeh2016}.
The number of simsun permutations in $\symS_n$ is given by the Euler number $E_{n+1}$, see \oeis{A000111}.

\item \defin{King permutations} are permutations where no two adjacent entries differ by $1$ in one-line notation.
This can be described via mesh-pattern avoidance, see \oeis{A002464}.

\item \label{vexillary}
A permutation is \defin{vexillary} if it avoids the pattern 2143, see \oeis{A005802}.

\item A permutation is \defin{Grassmann} (or \defin{Grassmannian}) if it has at most one descent.
These are special cases of vexillary permutations.
There are $2^n-n$ Grassmann permutations in $\symS_n$, see \oeis{A000325}.
See \hyperref[grassmannianPermutation]{the Schubert polynomial page} for more details.

\item A permutation $\pi$ is \defin{bi-Grassmann} if both $\pi$ and $\pi^{-1}$ are Grassmann. Equivalent with avoiding 2413.
This set has size $1+\binom{n+1}{3}$.

\item \defin{Separable} permutations are those which avoids $2413$ and $3142$, see \oeis{A006318}.
For reference, see \cite{West1995}.

\item \defin{Skew-merged} permutations avoid $2143$ and $3412$. These are exactly the ones
created by a union of an increasing sequence, and a decreasing sequence, see \cite[Thm.2.9]{Stankova1994}, \oeis{A029759}.
See \cite{Atkinson1998} for exact enumeration and connection with RSK.

\item \defin{Flattened permutations}, where runs are in lex order.
Enumerated by Bell numbers \oeis{A000110}, see \cite[Prop. 16]{NabawandaRakotondrajaoBamunoba2020}.

\item \defin{Parity-alternating permutations} are permutations which sends odd numbers to odd numbers,
and even numbers to even, see \cite{KebedeRakotondrajao2021}.
This can be generalized to \defin{mod-k-alternating permutations},
where we demand that $\pi(i) \equiv i \mod k$.
For pattern-avoidance in mod-k-alternating permutations, see \cite{AlexanderssonFufaGetachewQiu2022x}.


\item The \defin{Richardson permutation} with parameters $(k_1,k_2,\dotsc,k_m)$
is given by the following in two-line notation.
\[
\begin{pmatrix}
1 & 2 &       \dotsc & k_1   & k_1+1 & k_1+2 & \dotsc & k_1 + k2 & k_1+k_2+1 & \dotsc \\
k_1 & k_1-1 & \dotsc,   1		 & k_1+k_2 & k_1+k_2-1 & \dotsc & k_1+1 & k_1+k_2+k_3 & \dotsc \\
\end{pmatrix}
\]
Basically, we take the unique longest permutation in each of the sets $\symS_{k_1},\dotsc,\symS_{k_m}$,
and produce the \hyperref[directSumPermutations]{direct sum} of these.

\item \defin{Wachs permutations} --- \oeis{A359039}.
These are the permutations in $\symS_n$ defined as
\[
 W_n \coloneqq \{ \sigma \in \symS_n : |\sigma^{-1}(i)-\sigma^{-1}(i^*)|\leq 1 \text{ for all } i \in [n-1] \},
\]
where $i^*$ is defined as
\[
 i^* \coloneqq \begin{cases}
 i-1 &\text{if $i$ is even}  \\
 i+1 &\text{if $i$ is odd and $i+1 \leq n$}  \\
 n & \text{otherwise.}
 \end{cases}
\]
These permutations are studied in \cite{BrentiSentinelli2022x}, and there are generalizations to
other types.

\end{itemize}
