\metatitle{Polytopes}
\metadescription{An introduction to lattice polytopes, Ehrhart polynomials,
order polytopes, hypersimplices, generalized permutohedra, flow and root
polytopes, IDP, Gorenstein and reflexive polytopes, and magic positivity.}
\metakeywords{lattice polytope, Ehrhart polynomial, h-star polynomial,
order polytope, chain polytope, hypersimplex, generalized permutohedron,
flow polytope, root polytope, reflexive polytope, Gorenstein polytope}



\section[polytope]{Polytopes}


A polytope is called a \defin{lattice polytope}\label{latticePolytope}
or \defin{integral polytope}\label{integralPolytope}
if all its vertices have integer coordinates.
If its vertices have rational coordinates, we call it a
\defin{rational polytope}\label{rationalPolytope}.
A convex polytope is called a \defin{regular polytope}\label{regularPolytope}
if its Euclidean symmetry group acts transitively on its flags, that is, on
maximal chains of faces.

\name{Marie-Charlotte Brandenburg}, \name{Jesús A. De Loera}, and
\name{Chiara Meroni} study exact algorithms for cutting \defin{polytopal
measures}\label{polytopalMeasure}, meaning Lebesgue measure restricted to
finite unions of full-dimensional convex polytopes
\cite{BrandenburgLoeraMeroni2026x}.  For fixed ambient dimension, their
algorithms describe all ham-sandwich cuts and more general
prescribed-proportion cuts as semialgebraic sets in polynomial time.  They also
compute centerpoints of rational polytopal measures, and show that the
centerpoints of a convex polytope are exactly its floating body at level
$1/(d+1)$.



\section[ehrhartPolytopes]{The Ehrhart polynomial and $h^*$-vector}

Let $P$ be a lattice polytope.  The number of lattice points in the dilation
$n\cdot P$ is a polynomial in $n$.
The \defin{Ehrhart polynomial of $P$}\label{ehrhartPolynomialDefinition}
is the polynomial
$E_P(n) \coloneqq |n P \cap \setZ^d|$.
The \defin{Ehrhart series}\label{ehrhartSeries} is
\[
\sum_{n \geq 0} |nP \cap \setZ^d| t^n = \frac{H_P(t)}{(1-t)^{d+1}}
\]
where $H_P(t) = h^*_0 + h^*_1 t + \dotsb + h^*_d t^d$ is the
\defin{$h^*$-polynomial}\label{hStarPolynomial} of $P$.
The vector $(h^*_0 , h^*_1, \dotsc,  h^*_d )$ is the
\defin{$h^*$-vector}\label{hStarVector}; all its entries are nonnegative
integers, see \cite{Stanley1993} and \cite{Beck2007}.
For inequalities that every \(h^*\)-vector must satisfy, see
\hyperref[hStarInequalities]{the \(h^*\)-inequalities subsection} on the
Ehrhart theory page, which gives a more focused overview of this material.

For a $d$-dimensional polytope, we have the following relation between
entries in the \hyperref[hStarVector]{$h^*$-vector} and the
\hyperref[ehrhartPolynomialDefinition]{Ehrhart polynomial}:
\[
 E_P(t) = \sum_{j=0}^d \binom{t+d-j}{d} h^*_j.
\]
There is a similar relation to express $H_P(t)$ in terms of the
Ehrhart polynomial coefficients involving the \hyperref[eulerianPolynomial]{Eulerian polynomials}.

\name{Victor Reiner} and \name{Brendon Rhoades} introduce a $q$-deformation
of the Ehrhart series using harmonic spaces and Macaulay inverse systems for
finite point configurations \cite{ReinerRhoades2024x}.  They conjecture that
this $q$-Ehrhart series is rational and construct a bigraded algebra whose
Hilbert series recovers the same series.


For a \emph{rational} polytope $P$ with denominator $q$
(the smallest positive integer such that $qP$ is a lattice polytope),
the lattice-point counting function $E_P(t) = |tP \cap \setZ^d|$ is a
\defin{quasi-polynomial}\label{ehrhartQuasiPolynomial} of degree $d$
and period dividing $q$.
The \hyperref[ehrhartSeries]{Ehrhart series} then satisfies
\[
 \sum_{n \geq 0} E_P(n) z^n = \frac{h^*(P;z)}{(1-z^q)^{d+1}},
\]
where $h^*(P;z)$ is a polynomial of degree at most $q(d+1)-1$
with nonnegative integer coefficients, see \cite{Beck2007} and
\cite{Stanley1993}.

\name{Jesús A. De Loera}, \name{Laura Escobar}, \name{Nathan Kaplan}, and
\name{Chengyang Wang} develop a
\defin{weight-lifting}\label{weightLifting} construction for sums
of weighted lattice points of polytopes \cite{LoeraEscobarKaplanWang2024x}.
For many quasi-polynomial weights, the weighted sum over lattice points in
$P$ is transformed into an ordinary lattice-point count in a higher-dimensional
polytope.  Their applications include representation-theoretic multiplicities
such as \hyperref[kostkaCoefficient]{Kostka coefficients}.

\name{Esme Bajo}, \name{Robert Davis}, \name{Jesús A. De Loera},
\name{Alexey Garber}, \name{Sofía Garzón Mora}, \name{Katharina Jochemko},
and \name{Josephine Yu} develop a weighted Ehrhart theory extending
Stanley's nonnegativity theorem \cite{BajoDavisLoeraGarberMoraJochemkoYu2024}.
They prove nonnegativity for weighted $h^*$-polynomials when the weights are
homogeneous polynomials decomposable into products of linear forms which are
nonnegative on the polytope.

Some rational polytopes exhibit \hyperref[periodCollapse]{period collapse},
meaning the Ehrhart quasi-polynomial is in fact a polynomial.
Certain \hyperref[gtKostkaEhrhart]{Gelfand--Tsetlin polytopes} have this property.
However, not every such polynomial arises as the
\hyperref[ehrhartPolynomialDefinition]{Ehrhart polynomial} of a lattice
polytope.
For example, \cite{HaaseMcAllister2007} construct a rational polygon $Q$
whose Ehrhart function is a polynomial with linear coefficient $1$,
but no integral polygon has this property.
\todo{Expand the Haase--McAllister period-collapse example
\cite{HaaseMcAllister2007}: spell out the associated \(h^*\)-polynomial and
the dimension-two obstruction showing that it is not the \(h^*\)-polynomial
of any lattice polygon, equivalently not the Ehrhart polynomial of any
same-dimensional lattice polytope.  Check whether the related
De Loera--McAllister GT-polytope examples \cite{DeLoeraMcAllister2004}
should be mentioned separately.}
\todo{Add Clarke--Higashitani--Zaffalon, \emph{Restricted Chain-Order
Polytopes via Combinatorial Mutations}, arXiv:2211.07995.  Explain that
combinatorial mutations preserve Ehrhart series, so restricted chain-order
polytopes give large period-collapse families.  Also note that period collapse
for many GT-type rational polytopes follows from the flagged skew Schur
function interpretation of the corresponding stretching functions.}

\name{Dave Anderson} and \name{Aniket Shah} introduce a weighted
$q$-enumeration of lattice points in a polytope \cite{AndersonShah2023x}.
They prove a Brion-type formula for the corresponding generating function and
relate the construction to toric arc schemes of normal fans.


\subsection[ehrhartMacdonaldReciprocity]{Ehrhart--Macdonald reciprocity}

Let $P$ be a $d$-dimensional rational polytope with Ehrhart polynomial
$E_P(t)$.
The \defin{Ehrhart--Macdonald reciprocity} states that
\[
	E_P(-t) = (-1)^d E_{P^\circ}(t)
\]
where $E_{P^\circ}(t) = |t P^\circ \cap \setZ^d|$ counts the lattice points
in the \emph{interior} of $tP$.
In terms of the Ehrhart series, this is equivalent to the relation
$H_P(t) = t^{d+1} H_P(1/t)$, meaning the
\hyperref[hStarVector]{$h^*$-vector} is palindromic if and
only if the polytope is \hyperref[gorenstein]{Gorenstein}.
The reciprocity can be seen as a discrete analogue of Poincar\'e duality.
This was first proved by \name{Ian G. Macdonald} \cite{Macdonald1971} for
lattice polytopes; see \cite{Beck2007} for an alternative proof in the
rational case.



\subsection[stanleyMonotonicity]{Stanley's monotonicity theorem}

\begin{theorem}[Stanley Monotonicity, {\cite{Stanley1993}}]
Suppose that $P \subseteq Q$ are rational polytopes with $qP$
and $qQ$ integral (for minimal possible $q \in \setZ_{>0}$).
Define the $h^*$-polynomials via
\[
 \operatorname{Ehr}(P;z) = \frac{h^*(P;z)}{(1-z^q)^{\dim(P)+1}}
 \qquad \text{and} \qquad
 \operatorname{Ehr}(Q;z) = \frac{h^*(Q;z)}{(1-z^q)^{\dim(Q)+1}}.
\]
Then $h^*(P;z) \leq h^*(Q;z)$ coefficient-wise.
\end{theorem}
See \cite{Beck2007} for an alternative proof.


\section[polytopeFamilies]{Families of polytopes}


\subsection[hypersimplex]{Hypersimplices}

For $0<k<n$, the \defin{hypersimplex}\label{hypersimplexDefinition}
$\Delta_{k,n}$ is
\[
  \Delta_{k,n}
  \coloneqq
  \left\{ x\in[0,1]^n : x_1+x_2+\dotsb+x_n=k \right\}.
\]
It is the \hyperref[matroidBasePolytope]{base polytope} of the uniform
matroid $U_{k,n}$, and hence a
\hyperref[generalizedPermutohedron]{generalized permutohedron}.
Its normalized volume is the Eulerian number $A_{k,n-1}$.

\name{Nan Li} proved a conjecture of \name{Richard P. Stanley} giving a
descent--exceedance interpretation of the
\hyperref[hStarPolynomial]{$h^*$-polynomial} of hypersimplices
\cite{Li2012}.  More precisely, for the half-open hypersimplex
\[
  \widetilde{\Delta}_{k,n}
  \coloneqq
  \left\{
    x\in[0,1]^{n-1} :
    k-1 < x_1+x_2+\dotsb+x_{n-1} \leq k
  \right\},
\]
one has
\[
  h^*(\widetilde{\Delta}_{k,n};t)
  =
  \sum_{\substack{w\in\symS_{n-1}\\
                  \operatorname{exc}(w)=k-1}}
  t^{\operatorname{des}(w)}.
\]
The proof uses a shelling of a unimodular triangulation of the hypersimplex.
A second shelling gives another formula using descents and Li's cover
statistic, and the construction extends to generalized hypersimplices related
to algebras of Veronese type \cite[Thms.~1.2, 1.3 and 7.3]{Li2012}.


\subsection[orderPolytope]{Order polytopes}

Given a \hyperref[poset]{partial order} $P$ on the set $[n]$, we associate a
polytope $\mathcal{O}(P)$, called the
\defin{order polytope}\label{orderPolytopeDefinition} of $P$.
It is the polytope in $\setR^n$ defined via the inequalities
\begin{align}
x_i \leq x_j \text{ if } i <_P j
\qquad \text{and} \qquad
0 \leq x_i \leq 1 \text{ for } 1\leq i \leq n.
\end{align}

There are order polytopes with negative Ehrhart coefficients,
see \cite[Fig.~3.87]{StanleyEC1} and \cite{Alexandersson2019CounterExamples}.


Let $Z_n$ be the zig-zag poset with cover relations
$z_1<z_2>z_3<z_4>\dotsb$.  The
\hyperref[hStarPolynomial]{$h^*$-polynomial} of the order polytope
$\mathcal{O}(Z_n)$ has a combinatorial interpretation in terms of the swap
statistic on alternating permutations.  This also gives another proof that
its coefficients are symmetric and \hyperref[unimodalPolynomial]{unimodal};
see \cite{CoonsSullivant2023}.
\name{Teemu Lundström} and \name{Leonardo Saud Maia Leite} study the order
polytopes of crown posets \cite{LundstromLeite2026}.  They give formulas for
$f$-vectors, recurrences for
\hyperref[ehrhartPolynomialDefinition]{Ehrhart polynomials}, and an
interpretation of the \hyperref[hStarPolynomial]{$h^*$-polynomial} using
cyclically alternating permutations.

\name{Benjamin Braun} and \name{Aryaman Jal} prove that order polytopes of
generalized snake posets have \hyperref[realRootedPolynomial]{real-rooted}
\hyperref[hStarPolynomial]{$h^*$-polynomials}
\cite{BraunJal2026x}.  These posets are width-two distributive lattices, and
the proof uses Stanley's interpretation of the $h^*$-polynomial as a
\hyperref[pEulerianPolynomial]{$P$-Eulerian polynomial}, its realization as a
non-nesting rook polynomial, and interlacing recurrences.  The ladder and
regular snake cases recover the Narayana and Delannoy polynomials,
respectively.


\subsection[chainPolytope]{Chain polytopes}

Given a finite poset $P$, its
\defin{chain polytope}\label{chainPolytopeDefinition}
$\mathcal{C}(P)$ is the
polytope in $\setR_{\geq 0}^{P}$ cut out by
\[
  x_{p_1}+x_{p_2}+\dotsb+x_{p_k}\leq 1
\]
for every chain $p_1<_P p_2<_P\dotsb<_P p_k$.  Stanley introduced chain
polytopes together with
\hyperref[orderPolytopeDefinition]{order polytopes}
\cite{Stanley86TwoPosetPolytopes}; the two have the same
\hyperref[ehrhartPolynomialDefinition]{Ehrhart polynomial}.



\subsection[markedOrderPolytope]{Marked order polytopes}

The \defin{marked order polytope}\label{markedOrderPolytopeDefinition}
generalizes the \hyperref[orderPolytopeDefinition]{order polytope}; see
\cite{ArdilaBliemSalazar2011}.
It is obtained by replacing the conditions $0 \leq x_i \leq 1$ with
$a_i \leq x_i \leq b_i$ for $1\leq i \leq n$,
where the $a_i$ and $b_i$ are constants.

\name{Jan Stricker} studies $2$-levelness for marked order, marked chain, and
marked chain-order polytopes, and gives an exact formula for the
\hyperref[ehrhartPolynomialDefinition]{Ehrhart polynomial} of marked order
polytopes \cite{Stricker2024x}.  Via transfer
between the marked models, the same polynomial also gives the Ehrhart
polynomial for the corresponding marked chain and marked chain-order
polytopes.

The \hyperref[gtpolytopes]{Gelfand--Tsetlin polytopes} are marked order polytopes.


\subsection[crossPolytope]{Cross-polytopes}

The \defin{cross-polytope}\label{crossPolytopeDefinition}
in dimension $d$ is
\[
  \left\{ x\in\setR^d : |x_1|+\dotsb+|x_d|\leq 1 \right\}.
\]
\name{Krishna Menon} and \name{Emil Verkama} give a combinatorial
interpretation for scaled Ehrhart coefficients of cross-polytopes, and more
generally for pyramids over cross-polytopes, in terms of colored permutations
\cite{MenonVerkama2025x}.  In particular, this gives a combinatorial proof
that these Ehrhart coefficients are positive.


\subsection[rootPolytope]{Root polytopes}

For a finite \hyperref[rootSystem]{root system} $\Phi\subseteq\setR^d$, the
\defin{root polytope}\label{rootPolytopeDefinition} is the convex hull
$\operatorname{conv}(\Phi)$.  In type
$A$, one often also considers graphical root polytopes: for a graph $G$ on
$[n]$ with oriented edges $i\to j$, the associated root polytope is
\[
  Q_G\coloneqq
  \operatorname{conv}\left(\{0\}\cup\{e_i-e_j : i\to j \text{ is an edge of }G\}\right).
\]


\subsection[flowPolytope]{Flow polytopes}

Let $G=(V,E)$ be a directed graph and let $a\in\setR^V$ satisfy
\[
  \sum_{v\in V} a_v=0.
\]
The \defin{flow polytope}\label{flowPolytopeDefinition}
$\mathcal{F}_G(a)$ is the set of nonnegative
flows $f:E\to\setR_{\geq 0}$ such that, for every vertex $v$,
\[
  \sum_{e=(v,w)} f(e)-\sum_{e=(u,v)} f(e)=a_v.
\]
Thus the vector $a$ records the netflow at each vertex.  Flow polytopes are
closely related to \hyperref[rootPolytopeDefinition]{root polytopes},
\hyperref[kostantPartitionFunction]{Kostant partition functions}, and
\hyperref[gtpolytopes]{Gelfand--Tsetlin polytopes}; see for example
\cite{LiuMeszarosDizier2019flow}.
\name[R. S. González D’León]{Rafael S. González D’León},
\name{Christopher R. H. Hanusa}, and \name{Martha Yip} introduce permutation
flows as a combinatorial model for triangulations of flow polytopes
\cite{DLeonHanusaYip2025x}.  Their construction gives a recursive
triangulation model in which top-dimensional simplices are indexed by
permutation flows.


\subsection[permutohedron]{The permutohedron}

The \defin{permutohedron}\label{permutohedronDefinition} is the polytope
\[
  \operatorname{Perm}_n
  \coloneqq
  \operatorname{conv}
  \left\{ (w(1),w(2),\dotsc,w(n)) : w\in\symS_n \right\}.
\]
Equivalently, it is the convex hull of all coordinate permutations of
$(1,2,\dotsc,n)$.  Its face structure is governed by ordered set partitions
of $[n]$, and it is a basic example of a generalized permutohedron; see
\cite{ZieglerLecturesOnPolytopes95}.


\subsection[generalizedPermutohedron]{Generalized permutohedra}

A \defin{generalized permutohedron}\label{generalizedPermutohedronDefinition}
is a polytope whose normal fan coarsens the
braid fan.  Equivalently, every edge is parallel to $e_i-e_j$ for some
$i,j$.  Many generalized permutohedra can be described by
\hyperref[submodularFunction]{submodular functions}; in particular,
\hyperref[matroidBasePolytope]{matroid base polytopes} are generalized
permutohedra.


\subsection[birkhoffPolytope]{Birkhoff polytopes}

The \defin{Birkhoff polytope}\label{birkhoffPolytopeDefinition}
$\mathcal{B}_n$ is the polytope of
$n\times n$ doubly stochastic matrices:
\[
  \mathcal{B}_n
  \coloneqq
  \left\{
    X\in\setR_{\geq 0}^{n\times n} :
    \sum_j X_{ij}=1 \text{ for all } i,\,
    \sum_i X_{ij}=1 \text{ for all } j
  \right\}.
\]
Equivalently, by the Birkhoff--von Neumann theorem, $\mathcal{B}_n$ is
the convex hull of the $n\times n$ permutation matrices.  See
\cite{Pak2000} for several combinatorial questions about this polytope.
The Birkhoff polytope also appears as a
\hyperref[birkhoffPolytopeAsGelfandTsetlin]{Gelfand--Tsetlin polytope}.

\name{Esther Banaian}, \name{Sunita Chepuri}, \name{Emily Gunawan}, and
\name{Jianping Pan} define $c$-Birkhoff polytopes by taking convex hulls of
permutation matrices indexed by $c$-singletons for a Coxeter element $c$
\cite{BanaianChepuriGunawanPan2025x}.  In type $A$, these polytopes are
integrally equivalent to order polytopes of heap posets.  Their type $B$
version embeds signed permutations in $B_n$ into $S_{2n}$ and gives the
same conclusion: for every Coxeter element $c_B$ of $B_n$, the corresponding
type $B$ $c$-Birkhoff polytope is integrally equivalent to the order
polytope of the heap of the longest $c_B$-sorting word
\cite[Thm.~3.16]{BanaianChepuriGunawanPan2026x}.


\subsection[edgePolytope]{Edge polytopes}

Let $G=([n],E)$ be a finite simple graph.  The
\defin{edge polytope}\label{edgePolytopeDefinition} of $G$ is
\[
  P_G \coloneqq \operatorname{conv}\{e_i+e_j : \{i,j\}\in E\}
  \subseteq \setR^n.
\]
The related \defin{symmetric edge polytope}\label{symmetricEdgePolytope} is
\[
  P_G^{\pm} \coloneqq
  \operatorname{conv}\{\pm(e_i-e_j): \{i,j\}\in E\}.
\]
Symmetric edge polytopes often appear in
\hyperref[ehrhart]{Ehrhart-theoretic} questions, for
example in gamma-positivity and magic positivity problems.


\subsection[perfectMatchingPolytope]{Perfect matching polytopes}

For a graph $G=(V,E)$, the
\defin{perfect matching polytope}\label{perfectMatchingPolytopeDefinition}
is the convex hull
in $\setR^E$ of the indicator vectors of perfect matchings of $G$.  For the
complete bipartite graph $K_{n,n}$, this is the
\hyperref[birkhoffPolytope]{Birkhoff polytope}.


\subsection[associahedron]{Associahedra}

The \defin{associahedron}\label{associahedronDefinition},
also called the \defin{Stasheff polytope}\label{stasheffPolytope}, is the
polytope whose vertices correspond to triangulations of a convex polygon and
whose edges correspond to diagonal flips.  Equivalently, its face poset
records partial triangulations of the polygon.


\subsection[parkingFunctionPolytope]{Parking function polytopes}

Let $u=(u_1,\dotsc,u_n)\in\setR_{\geq 0}^n$ satisfy
$u_1\leq\dotsb\leq u_n$.  A vector
$a=(a_1,\dotsc,a_n)\in\setR_{\geq 0}^n$ is a \defin{$u$-parking function}
if its non-decreasing rearrangement $b_1\leq\dotsb\leq b_n$ satisfies
$b_i\leq u_i$ for all $i$.  The
\defin{parking function polytope}\label{parkingFunctionPolytopeDefinition}
$\operatorname{PF}(u)$ is the convex hull of all $u$-parking functions.

\name{Fu Liu} and \name{Warut Thawinrak} study these polytopes, including
their normal fans, face posets, $h$-polynomials, volumes, and Ehrhart
polynomials \cite{LiuThawinrak2025x}.  In this nonnegative convention,
the classical parking functions correspond to $u=(0,1,\dotsc,n-1)$.

\name{Mitsuki Hanada}, \name{John Lentfer}, and
\name[Andrés R. Vindas-Meléndez]{Andrés R. Vindas-Meléndez} study the
subfamily coming from $\xvec=(a,b,\dotsc,b)$ \cite{HanadaLentferVindasMelendez2023}.
They relate these parking-function polytopes to Pitman--Stanley polytopes and
partial permutahedra, and obtain a closed formula for the volume.


\section[polytopeIDP]{The integer decomposition property}

An integral polytope $\mathcal{P} \subset \setR^d$ is said to have the
\defin{integer decomposition property}\label{integerDecompositionProperty} (IDP)
if for every positive integer $k$ and $\xvec \in k \mathcal{P} \cap \setZ^d$,
we can find $\xvec_1,\xvec_2,\dotsc,\xvec_k \in \mathcal{P} \cap \setZ^d$
such that $\xvec_1 + \dotsb + \xvec_k = \xvec$.

In the following list, each entry implies the next:
\begin{itemize}
	\item $\mathcal{P}$ has a unimodular triangulation.
	\item $\mathcal{P}$ has the integer decomposition property.
	\item $\mathcal{P}$ is integral.
\end{itemize}





\section[gorenstein]{Reflexive and Gorenstein polytopes}\label{reflexivePolytope}

A lattice polytope $P \subset \setR^d$ containing the origin in its interior
is called \defin{reflexive}\label{reflexivePolytopeDefinition}
if the dual polytope
\[
 P^* \coloneqq \left\{ y \in \setR^d : \langle x, y \rangle \leq 1 \text{ for all } x \in P \right\}
\]
is again a lattice polytope.
Equivalently, $h^*_P(t)$ is palindromic of degree $d$, see \cite{Hibi92}.

A polytope $P$ is called \defin{Gorenstein}\label{gorensteinPolytope}
(of index $r$) if for some $r\in\setZ_{>0}$, the dilation $rP$ contains a
unique lattice point $m$, and the translation $rP-m$ is reflexive.



\section[magicPositiveEhrhart]{Magic positivity and reflexive polytopes}

Following \cite{FerroniHigashitani2024}, a degree-$d$ polynomial $f(n)$ is
called \defin{magic positive}\label{magicPositivePolynomial}
if it can be written in the form
\[
 f(n) = \sum_{i=0}^d a_i n^i (n+1)^{d-i}
\]
with $a_i \geq 0$ for all $i$.
For Ehrhart polynomials, this basis is useful because it links coefficient
positivity to \hyperref[realRootedPolynomial]{real-rootedness} of the
corresponding \hyperref[hStarPolynomial]{$h^*$-polynomial}.

\begin{theorem}[Brändén; see \cite[Thm.~4.19]{FerroniHigashitani2024}]
If the Ehrhart polynomial $E_P(n)$ of a lattice polytope is magic positive,
then $E_P(n)$ has nonnegative coefficients and $h^*_P(t)$ is
\hyperref[realRootedPolynomial]{real-rooted}.
\end{theorem}

If $W$ denotes Wagner's transformation characterized by
\[
 \sum_{n \geq 0} p(n) z^n = \frac{W(p)(z)}{(1-z)^{1+\deg p}},
\]
then magic positivity implies that $W(p)$ is real-rooted, see
\cite[Rem.~4.21]{FerroniHigashitani2024}.
For Ehrhart polynomials, $W(E_P)$ is exactly the $h^*$-polynomial.
This is the same transformation discussed on the page about
\hyperref[interlacingPolynomials]{interlacing polynomials}.

\name{Fu Liu} surveys families of Ehrhart-positive polytopes, examples with
negative Ehrhart coefficients, and open problems around Ehrhart positivity
\cite{Liu2017}.
\name{Masato Konoike} studies how dilation affects magic positivity
\cite{Konoike2025x}.  Every polynomial with positive real coefficients becomes
magic positive after sufficiently large positive dilation, but for each
dimension $d\geq 3$, some polytopes have fixed dilates that are not magic
positive.

\begin{example*}[Pitman--Stanley polytopes]
Let $y=(y_1,\dotsc,y_n)\in \setZ_{>0}^n$.  The
\defin{Pitman--Stanley polytope}\label{pitmanStanleyPolytope} is
\[
  \Pi_n(y)
  \coloneqq
  \left\{
    x\in \setR_{\geq 0}^n :
    x_1+\dotsb+x_k \leq y_1+\dotsb+y_k
    \text{ for } k=1,\dotsc,n
  \right\}.
\]
\name{Nicolas Avila}, \name{Luis Ferroni}, and
\name{Alejandro H. Morales} prove that the Ehrhart polynomial
$E_{\Pi_n(y)}(m)$ is magic positive \cite{AvilaFerroniMorales2026}.
More precisely, if
\[
  E_{\Pi_n(y)}(m)
  =
  \sum_{i=0}^n c_i m^i(1+m)^{n-i},
\]
then $n!c_i$ counts $y$-parking functions with exactly $i$ lucky cars
in a modified parking protocol.

It follows from Brändén's theorem that the $h^*$-polynomial
$h^*_{\Pi_n(y)}(z)$ is
\hyperref[realRootedPolynomial]{real-rooted}.  In particular, its coefficients
are \hyperref[logConcavePolynomial]{log-concave} and
\hyperref[unimodalPolynomial]{unimodal}.

For computations, one may use the lattice-point function
\[
  L_y(m)
  =
  \#\left\{
    x\in\setZ_{\geq 0}^n :
    x_1+\dotsb+x_k
    \leq
    m(y_1+\dotsb+y_k)
    \text{ for } k=1,\dotsc,n
  \right\}.
\]
The corresponding $h^*$-coefficients are
\[
  h^*_j
  =
  \sum_{i=0}^j
  (-1)^{j-i}
  \binom{n+1}{j-i}
  L_y(i).
\]
For example,
\[
\begin{array}{c|c}
 y & h^*_{\Pi_n(y)}(z) \\ \hline
 (1) & 1 \\
 (2) & 1+z \\
 (1,1) & 1+2z \\
 (2,1) & 1+6z+z^2 \\
 (1,1,1) & 1+10z+5z^2 \\
 (2,1,1) & 1+24z+24z^2+z^3 .
\end{array}
\]
See the page on \hyperref[parkingFunctions]{parking functions} for the
classical parking function interpretation.
\end{example*}

\begin{example*}[Stasheff and symmetric edge polytopes]
\name{Masato Konoike} studies two reflexive families
\cite{Konoike2024x}.
Let
\[
 \operatorname{St}_d \coloneqq \operatorname{conv}\left(
 \{\pm e_i : 1\leq i \leq d\} \cup
 \{e_i+\dotsb+e_j : 1\leq i \lt j \leq d\}
 \right).
\]
Then $\operatorname{St}_d$ is the dual of the
$d$-dimensional \hyperref[stasheffPolytope]{Stasheff polytope},
and Konoike proves that the Ehrhart polynomial of the Stasheff polytope
$\operatorname{St}_d^*$ is magic positive.

For the cycle graph $C_{d+1}$, the
\hyperref[symmetricEdgePolytope]{symmetric edge polytope} is
\[
 P_{C_{d+1}} \coloneqq
 \operatorname{conv}\left(
 \{\pm(e_i-e_{i+1}) : 1\leq i \leq d-1\} \cup \{\pm(e_d-e_1)\}
 \right).
\]
Writing the Ehrhart polynomial of the dual polytope as
\[
 E_{P_{C_{d+1}}^*}(n) = \sum_{j=0}^d a_j n^j (n+1)^{d-j},
\]
the same paper proves that $a_i > 0$ and $a_{d-i} > 0$ for $i=0,1,2$.
\end{example*}





\section[polytopePapers]{Further reading}

Here we mainly discuss polytopes that arise in the area of algebraic combinatorics.

\begin{itemize}

\item \name{Tyrrell B. McAllister} and \name{Hélène O. Rochais} study periods
of Ehrhart coefficients of rational polytopes \cite{McAllisterRochais2018}.

\item \name{Robert Davis}, \name{Jesús A. De Loera}, \name{Alexey Garber},
\name{Katharina Jochemko}, and \name{Josephine Yu} develop Ehrhart theory over
abelian group rings \cite{DavisLoeraGarberJochemkoYu2025x}.

\item \name{Matias von Bell} and \name{Cesar Ceballos} relate framing lattices
and flow polytopes \cite{BellCeballos2025x}.

\item \name{Konstanze Rietsch} and \name{Lauren Williams} study root
polytopes, flow polytopes, and order polytopes \cite{RietschWilliams2024x}.

\item \name{Max Köbl} studies roots of Ehrhart polynomials of symmetric edge
polytopes from complete multipartite graphs \cite{Kolbl2024x}.

\end{itemize}
