\metatitle{Resources}
\metadescription{Useful links, suggested books and software resources related to symmetric functions and algebraic combinatorics.}


\section[general-useful-things]{Useful links}

\todo{Cleanup this page}

\begin{itemize}

\item \icon{sun} \href{https://www.doi2bib.org/}{Doi 2 bib.} This tool is extremely useful.

\item \icon{gem} Maria Gillespie's \href{http://www.mathematicalgemstones.com/}{Mathematical gemstones}.

\item \icon{file} Overview of \href{https://www.aimath.org/WWN/kostka/}{generalized Kostka Polynomials}.

\item \icon{file} \href{https://faculty.math.illinois.edu/~west/grammar.html}{Grammar and math} by D. B. West.

\item \icon{file} \href{https://arxiv.org/pdf/1612.04888.pdf}{A primer of mathematical writing} by S. G. Krantz.

\end{itemize}

\section[suggestedBooks]{Suggested books}


For an introduction to the topic of symmetric functions and algebraic combinatorics,
I suggest the following books. 
Richard Stanley, \emph{Enumerative Combinatorics II} \cite{StanleyEC2}, 
Lynne Butler, \emph{Subgroup Lattices and Symmetric Functions} \cite{Butler1994}, 
Bruce Sagan, \emph{The Symmetric Group} \cite{Sagan2001}, 
William Fulton, \emph{Young Tableaux} \cite{Fulton1997} and 
Ian Macdonald, \emph{Symmetric functions and Hall polynomials} \cite{Macdonald1995}.

Other related books:
David Bressoud, \emph{Proofs and confirmations: the story of 
the alternating sign matrix conjecture} \cite{Bressoud1999}, which is a very nice reading.



\href{http://www-math.mit.edu/~rstan/papers/ubc.pdf}{Research story: R. Stanley --- How the upper bound conjecture was proved}

 
\href{https://www.emis.de/journals/SLC/wpapers/s54Ahaglund.pdf}{Research story: J. Haglund --- The genesis of the Macdonald  polynomial statistic}


\href{https://arxiv.org/pdf/2105.07884.pdf}{R. Stanley --- Enumerative and Algebraic combinatorics in the 1960's and 1970's.}





\section[software]{Software resources}


\begin{itemize}

\item \icon{database} \href{https://math.depaul.edu/~bridget/patterns.html}{Bridget Tenner's} permutation pattern database.

\item \icon{file-code} \href{https://github.com/PerAlexandersson/Mathematica-packages}{Mathematica packages (Per Alexandersson)}
Packages for Catalan objects, Symmetric functions, posets, tableaux and unicellular chromatic calculations.


\end{itemize}