\metatitle{Schubert polynomials}
\metadescription{An introduction to Schubert polynomials, their definitions via divided difference operators, combinatorial formulas using reduced words and pipe dreams, and various properties including representation theory connections, GT polytopes, and multiplication rules.}


\section[schubert]{Schubert polynomials}

\begin{polydata}{schubert}
  Name   & Schubert polynomials \\
  Space  & All \\
  Basis  & Yes \\
  Rating & 5 \\
  Bib    & LascouxSchutzenberger1982 \\
  Year   & 1982 \\
  Category & Schubert \\
  Contains & schurFlagged | LascouxSchutzenberger1982,ReinerShimozono1995 | scope=vexillary permutations \\
  Contains & schurS | LascouxSchutzenberger1982 | scope=Grassmannian permutations \\
  StableLimit & stanleySym | Stanley1984 \\
  PositiveIn & key | LascouxSchutzenberger1990 \\
  PositiveIn & slideF | AssafSearles2016 \\
  PositiveIn & forestPolynomials | NadeauTewari2023x | proof=Thm. 1.4 \\
\end{polydata}

\todo{Legacy relation hints from old polynomialData.m/tikzdata*.txt to verify or migrate:
- PositiveIn schubert -> keyQuasi (current guess: schubert -> kohnert); Link=arxiv:1707.01172; from polynomialData.m.
}


\todo{Add this on Specializations
http://de.arxiv.org/pdf/1811.11596.pdf
}

\todo{https://arxiv.org/pdf/1812.00321.pdf}





\todo{
When coeffs are 0 or 1. Permutation-patterns
https://arxiv.org/pdf/1903.10332.pdf
Also permutation patterns:
https://arxiv.org/pdf/1705.02065.pdf
Even more patterns:
https://arxiv.org/pdf/1909.07206.pdf (this has a nice background - incorporate!)
}

Schubert polynomials were introduced by \name{Alain Lascoux} and \name{M.-P. Schützenberger} in \cite{LascouxSchutzenberger1982}.
The Schubert polynomials generalize the \hyperref[schurS]{Schur polynomials} and
represent Schubert cycles in flag varieties.
The Schubert polynomials are generalized by the \hyperref[grothendieck]{Grothendieck polynomials}.
\name{Colleen Robichaux}, \name{Harshit Yadav}, and \name{Alexander Yong}
survey edge-labeled Young tableaux as models for equivariant Schubert
calculus of Grassmannians \cite{RobichauxYadavYong2022}.  They also report on
a shifted analogue connected to isotropic Grassmannians.

The Schubert polynomials are indexed by permutations. The set 
\[
\{ \schubert_\omega(x_1,\dotsc,x_n) \}_{\omega \in \symS_n}
\]
is then a basis for the space $\setZ[x_1,\dotsc,x_n]/I_n$
where $I_n = \langle \elementaryE_1, \elementaryE_2,\dotsc, \elementaryE_n \rangle$,
i.e., the ideal generated by non-constant symmetric functions.
The Schubert polynomials $\schubert_\omega$ with $\omega \in \symS_\infty$
form an integral basis for $\setZ[x_1,x_2,\dotsc]$.


\subsection[schubertOperatorformula]{Operator definition}

Consider the ring $\setR[x_1,\dotsc,x_n]$ and let $s_i$ act by permuting $x_i$ and $x_{i+1}$.
Define $\partial_i(f)$ as $\frac{f-s_i(f)}{x_i - x_{i+1}}$ for $i=1,\dotsc,n-1$.
These \defin{divided difference operators} satisfy the \hyperref[braidRelations]{braid relations}, 
and we let $\partial_\omega = \partial_{a_1} \partial_{a_2} \dotsm \partial_{a_k}$
where $(a_1,a_2,\dotsc,a_k)$ is a \hyperref[reducedWord]{reduced word} for $\omega$.

The \defin{Schubert polynomials} are then defined as
\begin{equation*}
 \schubert_\omega(x_1,\dotsc,x_n) \coloneqq \partial_{\omega^{-1}\omega_0}(x_1^{n-1}x_2^{n-2}\dotsm x_{n-1}).
\end{equation*}


\subsection[schubertReducedwordFormula]{Reduced word formula}
Let $RW(\omega)$ be the set of reduced words of $\omega$.

We say that a $k$-tuple $\alpha_1,\dotsc,\alpha_k$ is \defin{$a$-compatible}
if 
\begin{itemize}
\item
$\alpha_1 \leq \alpha_2 \leq \dotsb \leq \alpha_k$
\item
$\alpha_j \leq a_j$ for $1 \leq j \leq k$
\item
$\alpha_j \lt \alpha_{j+1}$ whenever $a_j \lt a_{j+1}$.
\end{itemize}
Let $C(a)$ denote the set of $a$-compatible sequences.
Then 
\begin{equation*}
 \schubert_\omega(x_1,\dotsc,x_n) = \sum_{a \in RW(\omega)} \sum_{\alpha \in C(a)}
 x^\alpha
\end{equation*}



\subsection[schubertPipeDream]{Pipe dream formula}

The set of reduced words and compatible sequences can be 
described in a more combinatorial manner using rc-graphs.
Suppose $(\alpha_1,\dotsc,\alpha_k)$ is an $a$-compatible sequence, for $a\in RW(\omega)$.
Let
\[
D(a,\alpha) = \{ (\alpha_j, a_j - \alpha_j + 1) \text{ for } j=1,\dotsc,k \}.
\]
This set of points is illustrated in an \defin{rc-graph}, also known as \defin{pipe-dreams}.

\begin{example}
Let $\omega = 314652$. Then $a = 521345$ is in $RW(\omega)$,
and $\alpha = 111235$ is $a$-compatible. 
We mark the corresponding entries in $D(a,\alpha)$ with a cross, $+$, in the rc-graph:

\begin{figure}
\begin{ytableau}
\none \; & \none 1 & \none  2 & \none 3 & \none  4 & \none  5 & \none  6 \\
\none 1 & + & + & \cdot  & \cdot  & + & \cdot  \\
\none 2 & \cdot  & + & \cdot  & \cdot  & \cdot   \\
\none 3 & \cdot  & + & \cdot  & \cdot  \\
\none 4 & \cdot  & \cdot  & \cdot  \\
\none 5 & + & \cdot  \\
\none 6 & \cdot 
\end{ytableau}
\end{figure}

\end{example}

Let $RC(\omega)$ denote the set of all such diagrams.
Let $\xvec_D = \prod_{(i,j) \in D} x_i$ then the Schubert polynomial is given as
\begin{equation*}
 \schubert_\omega(x_1,\dotsc,x_n) = \sum_{D \in RC(\omega)} \xvec_D.
\end{equation*}

See also \cite{FominKirillov1996YangBaxter}.

\name{Allen Knutson} and \name{Paul Zinn-Justin} introduce
\defin{generic pipe dreams}, a tile model whose leading forms recover the
classical pipe dream and bumpless pipe dream formulas for double Schubert
polynomials \cite{KnutsonZinnJustin2024x}.  Their generic pipe dream
polynomials compute equivariant classes of lower-upper varieties and also
appear in formulas for Segre--Schwartz--MacPherson classes.


\subsection[schubertBumplessPipeDream]{Bumpless pipe dreams}

The notion of \defin{bumpless pipe dreams} was introduced by 
\name{Thomas Lam}, \name{Seung Jin Lee}, and \name{Mark Shimozono} in \cite{LamLeeShimozono2018x}.
They prove the following formula for the \hyperref[schubertDouble]{double Schubert polynomials}:
\[
\schubert_\omega(\xvec,\yvec) = \sum_{P \in RBPD(\omega)} \prod_{(i,j)\in D(P)} (x_i - y_j )
\]
where the sum runs over \emph{reduced bumpless pipe dreams}.
This formula can be lifted to a formula for \hyperref[grothendieck]{the double Grothendieck polynomials.}

\name{Anna Weigandt} uses bumpless pipe dreams to give explicit change-of-basis
rules between Grothendieck and Schubert polynomials
\cite{Weigandt2025x}.  In one direction, Grothendieck polynomials expand as
signed sums of Schubert polynomials; in the other, Schubert polynomials expand
positively in the Grothendieck basis.  The bumpless pipe dream model also
extends to a formula for expanding back stable Grothendieck polynomials in the
back stable Schubert basis \cite[Thm.~8.1]{Weigandt2025x}.
\name{Thomas Lam}, \name{Seung Jin Lee}, and \name{Mark Shimozono} develop
the corresponding back stable $K$-theoretic Schubert calculus
\cite{LamLeeShimozono2022}.  This includes back stable Grothendieck
polynomials and their role in the $K$-theory Schubert basis.
\name{Patricia Klein} and \name{Anna Weigandt} show that bumpless pipe dreams
index irreducible components of diagonal Groebner degenerations of matrix
Schubert varieties \cite{KleinWeigandt2021x}.  Thus bumpless pipe dreams have
a geometric role parallel to the classical anti-diagonal pipe-dream theory.

\name{Daoji Huang} and \name{Pavlo Pylyavskyy} study Knuth moves for the
Schubert RSK correspondence \cite{HuangPylyavskyy2023x}.  In this setting
biwords insert to bumpless pipe dreams, while recording objects are decorated
Bruhat chains.  Their plactic biwords with the same insertion bumpless pipe
dream are connected by Schubert Knuth moves.  \name{Daoji Huang} and
\name{Son Nguyen} later give a growth-diagram version for plactic biwords
\cite{HuangNguyen2023x}, recovering the Gao--Huang bijection between pipe
dreams and bumpless pipe dreams.
\name{Yibo Gao} and \name{Daoji Huang} give a direct canonical bijection
between reduced pipe dreams and reduced bumpless pipe dreams
\cite{GaoHuang2023PipeDreams}.  The bijection is defined through reduced
compatible sequences on bumpless pipe dreams and preserves Monk's formula.
\name{Daoji Huang} and \name{Pavlo Pylyavskyy} had earlier introduced left
and right Schubert RSK insertion into bumpless pipe dreams, with decorated
Bruhat chains as recording objects \cite{HuangPylyavskyy2022x}.  Their
growth-diagram framework gives a combinatorial rule for Schubert structure
constants in the separated-descent case.  \name{Adam Gregory} and
\name{Zachary Hamaker} show that, for vexillary permutations, the
Gao--Huang bijection preserves the associated semistandard tableaux
\cite{GregoryHamaker2022x}.
\name{Tianyi Yu} embeds bumpless pipe dreams as Bruhat chains
\cite{Yu2024x}, giving another bridge between the bumpless model and the
Bruhat-chain viewpoint on Schubert calculus.  \name{Christian Gaetz} and
\name{Yibo Gao} study weighted enumeration of Bruhat chains in the symmetric
group \cite{GaetzGao2020}; this is a useful source for Schubert formulas that
are naturally phrased in terms of Bruhat chains.



\subsection[schubertRepresentationTheory]{Representation theory}

This subsection follows the introduction in \cite{FanGuo2019}.

Let $D=(D_1,\dotsc,D_n)$ be an $n$-tuple of subsets of $[n]$.
Then $D$ encodes a subset of $[n]\times [n]$, where $D_j$ is considered as a 
set of row indices (seen as a strictly increasing list).
For example, the following diagram encodes $(23,14,\emptyset,124)$
\begin{figure}
\begin{ytableau}
\; & *(lightblue) & \; & *(lightblue) \; \\
*(lightblue) &  & \; & *(lightblue) \; \\
*(lightblue) & \;  & \; &  \; \\
\; & *(lightblue) & \; &  *(lightblue) \; 
\end{ytableau}
\end{figure}
We define a partial order on such diagrams, where $C \leq D$ if for every $j \in [n]$,
we have $|C_j| = |D_j|$ and $C_{jk} \leq D_{jk}$ for all $k$.

Let $B \subset GL_n(\setC)$ be the set of invertible upper-triangular $n\times n$-matrices.
Let $Y$ denote the upper-triangular matrix with indeterminate entries $\{y_{ij}\}_{i\leq j}$,
and let $\setC[Y]$ be the polynomial ring with these indeterminates.
We have that a matrix $M \in GL_n(\setC)$ act on $f \in \setC[Y]$ on the right by $f \cdot M \coloneqq f(M^{-1} Y)$.

Given a diagram $D$, the \defin{flagged Weyl module} $\mathcal{M}_D$ is a $B$-module, defined as
\[
  \mathcal{M}_D \coloneqq \mathrm{Span} \left\{   \prod_{j=1}^n \det( Y_{D_j}^{C_j} ) : C \leq D  \right\} \subseteq \setC[Y],
\]
where $ Y_{D_j}^{C_j}$ denotes the submatrix of $Y$ with row-indices from $C_j$ and column-indices from $D_j$.

The character, $char(M_D)(x_1,\dotsc,x_n)$ of $\mathcal{M}_D$ is defined as the trace of the map $X : \mathcal{M}_D \to \mathcal{M}_D$,
where $X = (x_1,\dotsc,x_n)$ is a diagonal matrix acting on $\setC[Y]$.
Note that for $C \leq D$,
\[
\left( \prod_{j=1}^n \det( Y_{D_j}^{C_j} ) \right) \cdot X = 
 \prod_{j=1}^n  \prod_{i \in C_j} x_i^{-1} \det( Y_{D_j}^{C_j} )
\]
so we see that all polynomials $\det( Y_{D_j}^{C_j} )$ with $C \leq D$ are eigenvectors, and do in fact span $\mathcal{M}_D$.
It was shown in \cite{KraskiewiczPragacz1987,KraskiewiczPragacz2004} 
that the Schubert polynomial indexed by $\omega$ is related to the character of $\mathcal{M}_{D(\omega)}$, via
\[
\schubert_\omega(x_1,\dotsc,x_n) = char(M_{D(\omega)})(x^{-1}_1,\dotsc,x^{-1}_n).
\]

\name{Zhuowei Lin}, \name[Simon C. Y. Peng]{Simon C.~Y.~Peng}, and
\name[Sophie C. C. Sun]{Sophie C.~C.~Sun} prove the
\name{Karola Mészáros}--\name[Avery St. Dizier]{Avery St.~Dizier}--
\name{Shijie Tanjaya} upper-bound conjecture for dual flagged Weyl characters
\cite{LinPengSun2023x}.  The criterion says that the natural upper bound is
attained exactly for diagrams avoiding a specified forbidden subdiagram.  Rothe
diagrams and skyline diagrams give the Schubert and key-polynomial special
cases.

\name{Yibo Gao}, \name{Reuven Hodges}, and \name{Alexander Yong} classify
Levi-spherical Schubert varieties \cite{GaoHodgesYong2023}.  Their proof uses
type $A$ Demazure characters, i.e. \hyperref[key]{key polynomials}, and gives
a formulation in terms of standard Coxeter elements.

\name{Peter L. Guo}, \name{Zhuowei Lin}, and \name[Simon C. Y. Peng]{Simon C.~Y.~Peng}
prove the zero-one criterion for dual characters of flagged Weyl modules
\cite{GuoLinPeng2025}.  A diagram $D$ has zero-one dual character if and only
if it avoids the twelve multiplicitous configurations of
\cite[Fig.~1.1]{GuoLinPeng2025}.  Since Schubert and key polynomials are
special cases of dual flagged Weyl characters, this gives a unified proof of
the known zero-one criteria for Schubert and key polynomials.

\name{David Anderson} gives a representation-theoretic interpretation of the
\name{Philippe Nadeau}--\name{Hunter Spink}--\name{Vasu Tewari} recursion for
Schubert polynomials \cite{Anderson2024x}.  Using filtrations of
Kraskiewicz--Pragacz Schubert modules, Anderson proves that the recursion
determines the characters of these modules, and extends the same mechanism to
flagged Schur modules coming from transparent and translucent diagrams.


\subsection[schubertGTFormula]{Gelfand--Tsetlin polytopes}

In \cite{LiuMeszarosDizier2019}, the authors express the Schubert polynomials
as a projection of a Minkowski sum of \hyperref[gtpolytopes]{Gelfand--Tsetlin polytopes}.
This generalizes the way Schur polynomials are expressible as 
an integer point transform of a single Gelfand--Tsetlin polytope.

\name{Federico Castillo}, \name{Yairon Cid-Ruiz}, \name{Fatemeh Mohammadi},
and \name{Jonathan Montaño} prove that double Schubert polynomials have
saturated Newton polytopes \cite{CastilloCidRuizMohammadiMontano2023}.
Thus every lattice point in the Newton polytope occurs as an exponent vector.

See also a similar result in \cite{FinkMeszarosDizier2018}, 
using the integer point transforms of generalized permutahedra.

Top-degree components of Lascoux polynomials provide another bridge to
Schubert polynomials.  In \cite{Yu2025}, \name{Tianyi Yu} shows that
top Lascoux polynomials are reverse-complement transforms of Schubert
polynomials and that their structure constants coincide with Schubert
structure constants.

\name{Sylvester W. Zhang} relates back stable Schubert calculus to the
Heisenberg algebra \cite{Zhang2024x}.  The Hilbert space with basis indexed by
infinite permutations and the ring of back symmetric functions both carry
faithful Heisenberg actions, and the map sending an infinite permutation to
its back stable Schubert polynomial is an isomorphism of representations.
The same framework interprets the pipe dream model as Hamiltonian time
evolution of two-dimensional fermions.



\subsection[schubertBlockDecomposition]{Block decomposition formula}

If $\sigma \in \symS_j$ and $\tau \in \symS_{k}$,
we can compute the \hyperref[directSumPermutations]{skew sum} $\sigma \times \tau \in \symS_{j+k}$.
We then have the factorization (see \cite{BilleyJockuschStanley1993})
\[
 \schubert_{\sigma \times \tau} = (x_1 \dotsm x_j)^k \schubert_{\sigma} \; \uparrow^j \! \schubert_{\tau}
\]
where $\uparrow^k x_i \coloneqq x_{j+k}$ is the operator which 
increases the variable index by $j$ on all variables.


\subsection[grassmannianPermutation]{Grassmannian permutations}

A permutation $\omega \in \symS_n$ is called \defin{Grassmannian} (of descent $k$) if $\omega_i \lt \omega_{i+1}$
for all $i\neq k$. All Grassmannian permutations are \hyperref[vexillary]{vexillary}, which means it is 2143-avoiding.

A Grassmannian permutation determines a partition:
\[
\lambda(\omega) = (\omega_k - k, \omega_{k-1} - k + 1, \dotsc, \omega_1 - 1 ).
\] 
We then have the identity 
\begin{equation*}
\schurS_{\lambda(\omega)}(x_1,\dotsc,x_k) = \schubert_{\omega}(\xvec).
\end{equation*}

Furthermore, the \hyperref[stanleySym]{Stanley symmetric functions} 
are obtained as a stable limit of the Schubert polynomial.


\subsection[schubertMurnaghanNakayama]{Murnaghan--Nakayama rule}

In \cite{MorrisonSottile2018} a Murnaghan--Nakayama rule is proved:
\begin{equation*}
\powerSum_r(x_1,\dotsc,x_k) \schubert_\omega(\xvec) = \sum (-1)^{\epsilon_k(\sigma)} \schubert_{\omega \cdot \sigma}(\xvec)
\end{equation*} 






\subsection[schubertMonksRule]{Monks rule}

The geometric version of this result was given by \name{David Monk} in \cite{Monk1959}.
(\href{http://www.wfnmc.org/erdmonk.html}{David Monk} is a prolific IMO problem constructor.)

Let $s_r$ be a simple transposition. Then
\[
\schubert_{s_r}(\xvec) \schubert_\omega(\xvec) = \sum \schubert_{\omega t_{ij}}(\xvec)
\]
where the sum is over all transpositions $t_{ij}$ such that $i\leq r \lt j$ and 
$\length(\omega t_{ij}) = \length(\omega)+1$.
Notice that 
\[
\schubert_{s_r}(\xvec) = x_1+x_2+\dotsb + x_r.
\]

A combinatorial proof is given in \cite{CoskunTaskin2018}, and a bijective proof using bumpless pipe
dreams can be found in \cite{Huang2020x}.
Huang also proves Monk's rule for the double Schubert polynomials with this model.

\subsection[schubertPieriRule]{Pieri rule}

\name{Frank Sottile} \cite{Sottile1996} provides the following Pieri-type formulas for Schubert polynomials.
Let $\omega \in \symS_n$, then
\[
\completeH_{m}(x_1,\dotsc,x_k) \schubert_\omega(\xvec) = \sum_{\omega'} \schubert_{\omega'}(\xvec)
\]
where the sum run over all $w = w't_{a_1b_1}\dotsm t_{a_mb_m}$ such that $a_i \leq k \leq b_i$,
$\length(w't_{a_1b_1}\dotsm t_{a_ib_i}) = \length(w')+i$ and all $b_i$ distinct.

Similarly,
\[
\elementaryE_{m}(x_1,\dotsc,x_k) \schubert_\omega(\xvec) = \sum_{\omega'} \schubert_{\omega'}(\xvec)
\]
with the same sum as above, but now the $a_i$ are distinct.


\subsection[schubertSottileRule]{Sottile rule}

There is a rule for multiplying $\schubert_\omega(\xvec)$ with a Schur polynomial indexed by a hook.



\subsection[schubertCauchyFormula]{Cauchy identity}

By letting $w=\omega_0$, in the \hyperref[schubertDoubleGiambelliFormula]{Giambelli formula} for double Schubert 
polynomials, one can prove the following \defin{Cauchy identity} for Schubert polynomials:
\[
\prod_{i+j \leq n} (x_i - y_j) = \sum_{\omega} \schubert_{\omega}(\xvec) \schubert_{\omega_0 \omega}(-\yvec).
\]


\subsection[schubertLittlewoodRichardson]{Littlewood--Richardson rule}


\begin{problem}
Give a combinatorial interpretation of the coefficients in the product
\[
\schubert_u(\xvec) \schubert_v(\xvec) = \sum_{w} c^w_{uv} \schubert_w(\xvec).
\]
\end{problem}
Due to the geometric interpretation of Schubert polynomials, 
it is known that the coefficients $c^w_{uv}$ are non-negative integers.

Note that the coefficients $c^w_{uv}$ generalize the classical Littlewood--Richardson 
coefficients in \hyperref[schurLittlewoodRichardson]{the Littlewood--Richardson rule for Schur polynomials}.
This is a major open problem in Schubert calculus and algebraic combinatorics.

This problem is closely related to \defin{the Fomin--Kirillov conjecture},
see \cite[Conjecture 8.1 and Problem 8.3]{FominKirillov1999}.
It concerns the evaluation of Schubert polynomials at Dunkl elements.


There are many special cases proved for this problem,
one particular case is treated in \cite{Huang2022},
regarding products of two permutations with separated descents.

\name{Oliver Pechenik} and \name{Anna Weigandt} introduce backstable clans and
prove an inverse Grassmannian Littlewood--Richardson rule
\cite{PechenikWeigandt2024}.  Their work also gives new linear relations among
Schubert structure constants, extending the range of products accessible by
positive combinatorial formulas.

\name{Matthew Samuel} has a python \href{https://pypi.org/project/schubmult/}{package for computing products of Schubert polynomials}.


\begin{problem}
Can one give an efficient algorithm to decide if $c^w_{uv} = 0$ or not?
\end{problem}
This problem is discussed in \cite{PakRobichaux2025xI}, where it is shown
that (assuming GRH) that this problem lies in $AM \cap coAM$,  
(see \href{https://en.wikipedia.org/wiki/Arthur%E2%80%93Merlin_protocol#AM}{Arthur--Merlin protocol} for complexity theory background).



\subsection[schubertKeyExpansion]{Expansion in key polynomials}

There are several proofs of the fact that Schubert polynomials expand positively into \hyperref[key]{key polynomials},
the first proof appearing in \cite{LascouxSchutzenberger1990}.

One recent proof is by using crystal bases, see \name{Sami Assaf} and \name{Anne Schilling} \cite{AssafSchilling2018}.
A crystal structure on RC-graphs where the key expansion is obtained is given 
by \name{Sarah Gold}, \name{Elizabeth Milićević}, and \name{Yuxuan Sun} in \cite{GoldMilicevicSun2024x}.

\subsection[schubertAsSumOfElementary]{Sum of elementary}

Let $\omega \in \symS_{n+1}$, and let $\elementaryE_{j}(m) = \elementaryE_{j}(x_1,\dotsc,x_m)$
denote the elementary symmetric functions of degree $j$ in $m$ variables.
Then there is a unique way to express $\schubert_\omega(\xvec)$ as
\[
\schubert_\omega(\xvec) = \sum a_{k_1 \dotsc k_n} \elementaryE_{k_1}(1) \elementaryE_{k_2}(2) \dotsm \elementaryE_{k_n}(n)
\]
where the sum ranges over integers such that $0 \leq k_i \leq i$ and $k_1+\dotsb + k_n = \length(\omega)$.
This expansion is the starting point for defining the quantum Schubert polynomials.
The coefficients $a_{k_1 \dotsc k_n}$ are in general not always positive.
These coefficients are studied in \cite{Winkel1998} but many questions are left unanswered.

\name{Dora Woodruff} characterizes the permutations $w$ for which the
Schubert polynomial $\schubert_w$ is a single standard elementary monomial
\cite{Woodruff2025x}.  The characterization is by pattern avoidance and also
has an analogue for complete homogeneous monomials.



\begin{problem}[See \cite{MerzonSmirnov2015}]
For each $n$, find a permutation $w\in\symS_n$
which maximizes $\schubert_w(1,1,\dotsc,1)$ and find the maximal value.
\end{problem}
The sequence is given as \oeis{A284661}.
See also \href{https://arxiv.org/pdf/1704.00851.pdf}{\name{Richard Stanley}'s Some Schubert Shenanigans}.

\name{Philippe Nadeau} and \name{Vasu Tewari} relate the permutahedral
variety, mixed Eulerian numbers, and principal specializations of Schubert
polynomials \cite{NadeauTewari2021}.
\name{Hugh Dennin} proves pattern lower bounds for principal specializations
of Schubert polynomials in the $1243$-avoiding case and extends the result to
$\beta$-Grothendieck polynomials for vexillary $1243$-avoiding permutations
\cite{Dennin2025}.  The proof gives bijective interpretations of the
coefficients appearing in these bounds.

\name{Karola Mészáros} and \name{Arthur Tanjaya} prove nonnegativity of an
inclusion--exclusion expression for Schubert polynomials of permutations
avoiding $1432$ and $1423$ \cite{MeszarosTanjaya2021x}.  This gives a partial
answer to a conjecture of \name{Yibo Gao} on principal specializations of
Schubert polynomials.
\name{Gleb Nenashev} studies differential operators on Schur and Schubert
polynomials \cite{Nenashev2020x}.
\name{Peter L. Guo} and \name{Zhuowei Lin} give lower bounds for the number
of support monomials of a Schubert polynomial, equivalently lattice points in
its Newton polytope, in terms of permutation-pattern occurrences
\cite{GuoLin2024x}.  Their framework is the more general setting of dual
characters of flagged Weyl modules.



\subsection[schubertComplexity]{Complexity}

Deciding if a monomial coefficient, or if a Schubert structure constant coefficient $c^w_{uv}$
is non-zero or not, can be done in probabilistic polynomial time, see \cite{AdveRobichauxYong2021x,PakRobichaux2025xII}.
See \href{https://www.youtube.com/watch?v=41UCZFIxZpY}{Colleen Robichaux --- Deciding positivity of Schubert coefficients - IPAM at UCLA}
for a YouTube video on this topic.
\name[Avery St. Dizier]{Avery St.~Dizier} and \name{Alexander Yong} connect
generalized permutahedra with Schubert calculus
\cite{DizierYong2020SchubertPermutahedra}.  They obtain sufficient vanishing
criteria for Schubert intersection numbers, using Schubitopes as Newton
polytopes of Schubert polynomials, and give a polynomial-time tableau test.
\name{Igor Pak} and \name{Colleen Robichaux} prove that the vanishing problem
for Schubert coefficients lies in $\mathrm{coAM}$ assuming the generalized
Riemann hypothesis \cite{PakRobichaux2024Vanishing}.  They also formulate a
conditional positive rule for deciding Schubert coefficient positivity
\cite{PakRobichaux2024Positivity}.
\name{Igor Pak} and \name{Colleen Robichaux} also give a signed puzzle rule
for Schubert coefficients based on Knutson's recurrence
\cite{PakRobichaux2025SignedPuzzles}.  They use it to prove polynomiality
for sums of Schubert coefficients with bounded number of inversions.


\section[schubertBCD]{Type B/C/D Schubert polynomials}

\begin{polydata}{schubertBCD}
  Name     & Type B/C/D Schubert polynomials \\
  Space    & All \\
  Basis    & True \\
  Rating   & 2 \\
  Bib      & FominKirillov1996 \\
  Year     & 1996 \\
  Symbol   & $\schubert^B_\omega(\xvec)$ \\
  Keywords & divided-difference  \\
  Category & Schubert \\
\end{polydata}


In type $B$, there are combinatorial analogs of Schubert polynomials introduced in \cite{FominKirillov1996}.
They also introduce \hyperref[stanleySymBC]{type $B$ Stanley symmetric functions}.

Divided difference formulas for Schubert polynomials in types $B$, $C$ and $D$ were 
introduced by \name{Billey} and \name{Haiman} \cite{BilleyHaiman1995}.
See also \name{Sara Billey}'s PhD thesis on Schubert polynomials for classical groups \cite{Billey1994Thesis}.

Type $B$, $C$ and $D$ double Schubert polynomials are defined in \cite{IkedaMihalceaNaruse2011}.
\name{Tomoo Matsumura} gives a tableau formula for vexillary Schubert
$P$-polynomials in type $C$ \cite{Matsumura2023}.  The formula uses flagged
factorial Schur $Q$-functions and a tableau model adapted to vexillary signed
permutations.
\name{Evgeny Smirnov} and \name{Anna Tutubalina} construct pipe-dream models
for Schubert polynomials of the classical groups \cite{SmirnovTutubalina2023},
complementing the divided-difference definitions in types $B$, $C$, and $D$.



\section[schubertSkew]{Skew Schubert polynomials}


\begin{polydata}{schubertSkew}
  Name   & Skew Schubert polynomials \\
  Space  & All \\
  Basis  & No \\
  Rating & 1 \\
  Bib    & LenartSottile2003 \\
  Year   & 2003 \\
  Category & Schubert \\
  Contains & schubert | LenartSottile2003 | scope=trivial skewing permutation \\
  PositiveIn & schubert | LenartSottile2003 \\
\end{polydata}

In \cite{LenartSottile2003}, a skew version of Schubert polynomials are defined.
They expand positively into Schubert polynomials, and a formula in the monomial basis is given.
\name{Harry Tamvakis} gives tableau formulas for skew Schubert polynomials
\cite{Tamvakis2023}, including formulas in classical types.


\section[schubertDual]{Dual Schubert polynomials}

The dual Schubert polynomials of \name{Alexander Postnikov} and
\name{Richard P. Stanley} are defined by a weighted-chain formula in Bruhat
order \cite{PostnikovStanley2008}.  \name{Zachary Hamaker} gives a
combinatorial proof that this formula is equivalent to a classical Cauchy
identity for polynomials \cite{Hamaker2023x}.  This also gives a natural
interpretation of the insertion algorithms of \name{Daoji Huang} and
\name{Pavlo Pylyavskyy}.


\section[schubertDouble]{Double Schubert polynomials}

\begin{polydata}{schubertDouble}
  Name     & Double Schubert polynomials \\
  Space    & All \\
  Basis    & True \\
  Rating   & 1 \\
  Bib      & LascouxSchutzenberger1982 \\
  Year     & 1982 \\
  Symbol   & $\schubert_\omega(\xvec;\yvec)$ \\
  Keywords & divided-difference  \\
  Category & Schubert \\
  SpecializesTo & schubert | LascouxSchutzenberger1982 | map=y=0 \\
\end{polydata}

\todo{Legacy relation hints from old polynomialData.m/tikzdata*.txt to verify or migrate:
- SpecializesTo doubleSchubert -> doubleSchur (current guess: schubertDouble -> schurFactorial); from polynomialData.m, tikzdataRelations.txt.
}


The double Schubert polynomials were introduced by
\name{Alain Lascoux} and \name[M.-P. Schützenberger]{Marcel-Paul Schützenberger}
\cite{LascouxSchutzenberger1982}, and by \name{Ian Macdonald} \cite{Macdonald1991}.

They are defined as
\begin{equation*}
 \schubert_\omega(x_1,\dotsc,x_n,y_1,\dotsc,y_n) = \partial_{\omega^{-1}\omega_0} \prod_{i + j \leq n} (x_i-y_j).
\end{equation*}
where the divided difference operator act on the $\xvec$-variables.
By letting $y_i=0$, we get the usual Schubert polynomials. 
The double Schubert polynomials generalize the double Schur polynomials.

\name{David Anderson} and \name{William Fulton} construct enriched type $A$
Schubert polynomials with coefficients in a polynomial ring
\cite{AndersonFulton2021x}.  Specializations recover double Schubert
polynomials, back-stable double Schubert polynomials, and type $C$ double
Schubert polynomials.  \name{David Anderson} also studies Schubert
polynomials through infinite-dimensional flag varieties and degeneracy loci
\cite{Anderson2025}, including positivity in equivariant coproduct and
back-stable/enriched Schubert expansions.

\todo{Later relation pass: decide whether enriched Schubert polynomials need a separate polydata node, and record their specializations to double, back-stable double, and type $C$ double Schubert polynomials. Add a small worked specialization example.}

There is a pipe-dream formula, expressing $\schubert_\omega(\xvec,\yvec) $
as a sum over \hyperref[schubertPipeDream]{pipe dreams}:
\begin{equation*}
 \schubert_\omega(\xvec,\yvec) = \sum_{D \in RC(\omega) } \prod_{(i,j) \in D} \left(x_{i} - y_{j} \right).
\end{equation*}
See also the survey by \name{Nantel Bergeron} and \name{Sara Billey} \cite{BergeronBilley1993}.

In a recent paper, \name{Allen Knutson} \cite{Knutson2019} provides new proofs that 
several combinatorial formulas for the double Schubert polynomials.

\name{Matthew J. Samuel} proves a Molev--Sagan type formula for products
of double Schubert polynomials under descent hypotheses \cite{Samuel2024}.
The same paper gives a Pieri rule for multiplying a double Schubert polynomial
by a factorial elementary symmetric polynomial, and gives positivity results
after specializing the two equivariant alphabets to be equal.
\name{Yibo Gao} and \name{Rui Xiong} prove Graham positivity results for
triple Schubert calculus \cite{GaoXiong2025x}.  As a corollary, they prove
Kirillov's conjectured positivity for skew divided difference operators
applied to Schubert polynomials.
\name{Serena An}, \name{Katherine Tung}, and \name{Yuchong Zhang} prove that
Postnikov--Stanley polynomials are \hyperref[lorentzianPolynomial]{Lorentzian}
\cite{AnTungZhang2024Lorentzian}.  These polynomials generalize skew dual
Schubert polynomials to arbitrary Weyl groups, and the proof realizes them as
degree polynomials of Richardson varieties.


\subsection[schubertDoubleGiambelliFormula]{Giambelli formula}


\todo{Explain the giambelli formula https://projecteuclid.org/download/pdf_1/euclid.em/1048516036 }


The Giambelli formula for double Schubert polynomials state that
\begin{equation*}
 \schubert_w(\xvec;\yvec) = \sum_{\substack{ v^{-1}u=w \\ \length(u)+\length(v) = \length(w) }}
 (-1)^{\length(v)}\schubert_u(\xvec) \schubert_v(\yvec).
\end{equation*}



\section[schubertAffine]{Affine Schubert polynomials}

\begin{polydata}{schubertAffine}
  Name & Affine Schubert polynomials \\
  Space  & All \\
  Basis  & No \\
  Rating & 1 \\
  Bib    & Lam2006Schubert \\
  Keywords & affine, schubert \\
  Year   & 2006 \\
  Category & Schubert \\
\end{polydata}


In \cite{Lee2015}, an affine extension of Schubert polynomials (indexed by affine permutations) 
is defined via divided difference operators after being introduced by \name{Thomas Lam} in \cite{Lam2006Schubert}.

In the \href{https://www.youtube.com/watch?v=c7aY09b7XjI}{YouTube video, Seungjin Lee introduces the affine Schubert polynomials}.
See 47:16 for the definition.

\todo{ add definition }

In \cite[Thm. 6.1]{LamLeeShimozono2019}, the authors prove that the affine Schubert polynomials 
expand positively in the monomial basis.


\section[schubertInvolution]{Involution Schubert polynomials}

\begin{polydata}{schubertInvolution}
 Name & Involution Schubert polynomials \\
  Space  & All \\
  Basis  & No \\
  Rating & 1 \\
  Symbol & $\ischubert_y(x_1,\dotsc,x_n)$ \\
  Bib    & HamakerMarbergPawlowski2018 \\
  Keywords & divided-difference, schubert \\
  Year   & 2018 \\
  Category & Schubert \\
  StableLimit & stanleySymInvolution | HamakerMarbergPawlowski2018 \\
\end{polydata}

The involution Schubert polynomials were introduced in \cite{WyserYoung2016}.
Additional properties are proved in \cite{HamakerMarbergPawlowski2018}, 
where the authors introduce the name of the family of polynomials.
The following definition is from that paper. 

Let $y \in I_n$, the set of involutions in $\symS_n$, and let $A(y)$ denote the set of 
minimal-length permutations $w$ such that $y = w^{-1} \circ w$.
That is, we consider all minimal factorizations of $y$ into simple transpositions 
of the form $(s_{i_1}\dotsc s_{i_k})\circ (s_{i_k}\dotsc s_{i_1})$.
Then let
\[
\ischubert_y(x_1,\dotsc,x_n) \coloneqq \sum_{w \in A(y)} \schubert_w(x_1,\dotsc,x_n)
\]
There is also a recursive way to produce these polynomials via divided difference operators.

The stable limit of these yield the \hyperref[stanleySymInvolution]{involution Stanley symmetric functions}.

A combinatorial model using \emph{involution pipe dreams} for these polynomials
is given by \name{Zachary Hamaker}, \name{Eric Marberg}, and \name{Brendan Pawlowski} in \cite{HamakerMarbergPawlowski2019}.

\todo{Add more info}


\section[schubertQuantum]{Quantum Schubert polynomials}


\begin{polydata}{schubertQuantum}
 Name & Quantum Schubert polynomials \\
  Space    & All \\
  Basis    & Yes \\
  Rating   & 3 \\
  Symbol   & $\schubert^q_\omega(\xvec)$ \\
  Bib      & FominGelfandPostnikov1997 \\
  Keywords & schubert \\
  Year     & 1997 \\
  Category & Schubert \\
  SpecializesTo & schubert | FominGelfandPostnikov1997 | map=q=0 \\
\end{polydata}


The quantum Schubert polynomials $\schubert^q_\omega(\xvec)$
is a deformation of the Schubert polynomials by a  vector $q=(q_1,\dotsc,q_{n-1})$.
These were introduced in \cite{FominGelfandPostnikov1997}.

Recall the \hyperref[schubertAsSumOfElementary]{formula}
that expresses the Schubert polynomials as sums of products of elementary symmetric functions:
\[
\schubert_\omega(\xvec) = \sum a_{k_1 \dotsc k_n} \elementaryE_{k_1}(1) \elementaryE_{k_2}(2) \dotsm \elementaryE_{k_n}(n)
\]
The quantum Schubert polynomials are then defined as
\[
\schubert_\omega(\xvec) = \sum a_{k_1 \dotsc k_n} \elementaryE^{q}_{k_1}(1) \elementaryE^{q}_{k_2}(2) \dotsm \elementaryE^{q}_{k_n}(n)
\]
where 
\[
\sum_{i=0}^k \elementaryE^{q}_{k}(i)t^i = \det(1+tG_k) \text{ and } G_k=
\begin{vmatrix}
x_1 & q_1 & 0 & \ldots & 0 \\
-1 & x_2 & q_2 & \ldots & 0 \\
0 & -1 & x_3 & \ldots & 0 \\
\vdots & \vdots &   & \ddots & \vdots \\
0 & 0 & 0 & \ldots & x_k \\
\end{vmatrix}
\]
By letting $q_i=0$, one recovers the classical Schubert polynomials.


There is a bumpless pipedream model for the quantum double Schubert polynomials, see \cite[Thm. 3.3]{LeOuyangTaoRestivoZhang2024x}.


\subsection[schubertQuantumMonksRule]{Quantum Monks rule}

The following refinement of Monk's rule is given in \cite[Theorem 7.1]{FominGelfandPostnikov1997}.
Let $s_r$ be a simple transposition. Then
\[
\schubert^q_{s_r}(\xvec) \schubert^q_\omega(\xvec) = 
(x_1+x_2+\dotsb + x_r) \schubert^q_\omega(\xvec) 
=
\sum \schubert^q_{\omega t_{ab}}(\xvec) +
\sum q_{cd} \schubert^q_{\omega t_{cd}}(\xvec)
\]
where the first sum is over all transpositions $t_{ab}$ such that $a\leq r \lt b$ and 
$\length(\omega t_{ab}) = \length(\omega)+1$,
and the second sum runs over all
transpositions $t_{cd}$ such that $c\leq r \lt d$ and 
$\length(\omega t_{cd}) = \length(\omega)-\length(t_{cd}) =
\length(\omega)-2(d-c)+1$.

Here, $q_{cd} = q_c q_{c+1} \dotsm q_d$.


\todo{https://arxiv.org/pdf/2601.03437}


\subsection[schubertQuantumPieriRule]{Quantum Pieri rule}


\subsection[schubertQuantumSottileRule]{Quantum Sottile rule}

\todo{This is by BBCSS, 2017}

\subsection[schubertQuantumMurnaghanNakayama]{Quantum Murnaghan--Nakayama}

\name{Carolina Benedetti Velásquez}, \name{Nantel Bergeron},
\name{Laura Colmenarejo}, \name{Franco Saliola}, and \name{Frank Sottile}
prove a quantum Murnaghan--Nakayama rule for the flag manifold
\cite{BenedettiVelasquezBergeronColmenarejoSaliolaSottile2025}.  The result
describes multiplication by a tautological class in small quantum cohomology
and is obtained from a hook quantum Schur multiplication formula together
with a detailed analysis of the quantum Bruhat order.




\section[schubertQuantumDouble]{Quantum Double Schubert polynomials}


\begin{polydata}{schubertQuantumDouble}
 Name & Quantum Double Schubert polynomials \\
  Space    & All \\
  Basis    & Yes \\
  Rating   & 1 \\
  Symbol   & $\schubert^q_\omega(\xvec;\yvec)$ \\
  Bib      & KirillovMaeno2000 \\
  Keywords & schubert \\
  Year     & 2000 \\
  Category & Schubert \\
\end{polydata}


The double quantum Schubert polynomials were introduced by \name{Anatoly Kirillov} and \name{Toshiaki Maeno} \cite{KirillovMaeno2000}.


\subsection[schubertQuantumDoubleCauchyIdentity]{Cauchy identity}

In \cite{KirillovMaeno2000}, the following Cauchy identity for
quantum double Schubert polynomials is given:
\[
\sum_{\omega \in \symS_n} \schubert^q_\omega(\xvec) \schubert^q_{\omega \omega_0}(\yvec)
= \schubert^q_{\omega_0}(\xvec;\yvec)
\]


\section[schubertUniversal]{Universal Schubert polynomials}


\begin{polydata}{schubertUniversal}
  Name & Universal Schubert polynomials \\
  Space    & All \\
  Basis    & Yes \\
  Rating   & 1 \\
  Symbol   & $\schubert_w(c;d)$ \\
  Bib      & Fulton1999 \\
  Keywords & schubert \\
  Year     & 1999 \\
  Category & Schubert \\
  SpecializesTo & schubert | Fulton1999 \\
\end{polydata}


The universal Schubert polynomials were introduced by \name{William Fulton}, \cite{Fulton1999} and 
generalize the (double) quantum Schubert polynomials.


The double universal Schubert polynomials are defined via
\begin{equation*}
 \schubert_w(c;d) = \sum_{u,v} (-1)^{\length(v)}\schubert_u(c) \schubert_v(d)
\end{equation*}
as for \hyperref[schubertDoubleGiambelliFormula]{double Schubert polynomials}.


\section[schubertTwisted]{Twisted Schubert polynomials}

\begin{polydata}{schubertTwisted}
 Name & Twisted Schubert polynomials \\
  Space    & All \\
  Basis    & Yes \\
  Rating   & 1 \\
  Symbol   & $\tilde{\schubert}_\omega$ \\
  Bib      & Liu2019  \\
  Keywords & schubert \\
  Year     & 2019 \\
  Category & Schubert \\
\end{polydata}

The \defin{twisted Schubert polynomials}
are defined via $\tilde{\schubert}_{\omega_0}=x_1^{n-1} x_2^{n-2}\dotsm x_{n-1}$,
and the recursion $\tilde{\schubert}_{\omega s_i}= (s_i + \partial_i)\tilde{\schubert}_{\omega}$.

It was recently proved in \cite{Liu2019} that these are positive in the monomial basis.
This is the earliest reference that I could find that explicitly define these polynomials.
They are implicitly studied in earlier works, and are related to
the Chern--Schwartz--MacPherson classes of Schubert cells in flag varieties. 
