
\metatitle{K-theoretic Schur P/Q polynomials}
\metadescription{K-theoretic analogs of Schur P and Schur Q functions, introduced by \name{Takeshi Ikeda} and \name{Hiroshi Naruse}, with properties and Littlewood--Richardson rules.}

\section[schurKPQ]{K-theoretic Schur P/Q polynomials}
\begin{polydata}{schurKP}
  Name   & K-theoretic Schur P-functions \\
  Space    &  Sym   \\
  Basis    &  True   \\
  Rating   &  1      \\
  Bib      &  IkedaNaruse2013 \\
  Year     &  2013 \\
  Keywords &  jacobi-trudi,tableaux,pieri,littlewood-richardson \\
  Symbol   &  $\schurKP_{\lambda}(\xvec)$ \\
  Category &  Schur \\
  KTheoreticAnalogueOf & schurP | IkedaNaruse2013 \\
\end{polydata}

\begin{polydata}{schurKQ}
  Name   & K-theoretic Schur Q-functions \\
  Space    &  Sym   \\
  Basis    &  True   \\
  Rating   &  1      \\
  Bib      &  IkedaNaruse2013 \\
  Year     &  2013 \\
  Keywords &  jacobi-trudi,tableaux,pieri,littlewood-richardson \\
  Symbol   &  $\schurKQ_{\lambda}(\xvec)$ \\
  Category &  Schur \\
  KTheoreticAnalogueOf & schurQ | IkedaNaruse2013 \\
  SignedIn & schurKP | ChiuMarberg2024 | sign_rule=Theorem 1.1; note=NN[beta]-positive in the cases of Corollary 1.2 \\
\end{polydata}


In \cite{IkedaNaruse2013}, \name{Takeshi Ikeda} and \name{Hiroshi Naruse} introduce $K$-theoretic analogs of the \hyperref[schurP]{Schur $P$}
and \hyperref[schurQ]{Schur $Q$} functions, as well as factorial versions of these.
The factorial versions are denoted $\schurKP_{\lambda}(\xvec|b)$ and $\schurKQ_{\lambda}(\xvec|b)$,
where $b = (b_1,b_2,\dotsc)$ is a vector of parameters.
When all these are set to $0$, we obtain $\schurKP_{\lambda}(\xvec)$ and $\schurKQ_{\lambda}(\xvec)$. 
These are $K$-theoretic analogs of $\schurP_{\lambda}(\xvec)$ and  $\schurQ_{\lambda}(\xvec)$.
Ikeda and Naruse give several formulas for these functions, including a
ratio-of-Pfaffians formula, a shifted set-valued tableau formula, and an
excited Young diagram formula.

They give a presentation for equivariant $K$-theory classes of the orthogonal and Lagrangian Grassmannian.

\name{Eric Marberg} and \name{Brendan Pawlowski} give $K$-theory formulas
for orthogonal and symplectic orbit closures \cite{MarbergPawlowski2020}.
Their stable Grothendieck analogues for fixed-point-free involutions expand
positively in $K$-theoretic Schur $P$-functions, and related involution
families are connected to $K$-theoretic Schur $Q$-functions.
\name{Shinsuke Iwao} gives a boson--fermion correspondence for dual
$K$-theoretic $P$- and $Q$-functions \cite{Iwao2027}.  The construction uses
$\beta$-deformed neutral fermions and vertex operators, and recovers generating
functions for the dual $K$-theoretic Schur $P$- and $Q$-functions.

\name{Joel B. Lewis} and \name{Eric Marberg} prove the Nakagawa--Naruse
conjecture that the dual $K$-theoretic Schur $P$- and $Q$-functions are
generating functions for \hyperref[planePartitions]{shifted plane partitions}
\cite{LewisMarberg2024}.
They also describe the images of these functions under the involution $\omega$
and verify an Ikeda--Naruse basis conjecture for the $GQ$-functions.

\name{Yu-Cheng Chiu} and \name{Eric Marberg} prove identities expanding
$K$-theoretic Schur $Q$-functions in terms of $K$-theoretic Schur
$P$-functions \cite{ChiuMarberg2024}.  Their main formula is a signed finite
expansion, and they characterize when its coefficients lie in
$\setN[\beta]$; see \cite[Thm.~1.1 and Cor.~1.2]{ChiuMarberg2024}.  Their
formulas extend to skew and dual versions, and they also prove a shifted skew
Cauchy identity for symmetric Grothendieck polynomials.
\name{Eric Marberg} proves that the $K$-Stanley symmetric functions of
classical types expand positively into these $K$-theoretic Schur $P$- and
$Q$-functions \cite{Marberg2025x}.

% Related Rust:
% /home/paxinum/Dropbox/AI-projects/rust/sym-poly/sym/examples/k_schur_pq_site_example.rs
\begin{example*}[A $Q$-to-$P$ expansion]
In the notation of \cite[Thm.~1.1]{ChiuMarberg2024}, the coefficient of
$\schurKP_\lambda$ in $\schurKQ_\mu$ is obtained by summing over strict
partitions $\lambda \supseteq \mu$ with $\ell(\lambda)=\ell(\mu)$ such that
the shifted skew diagram $\lambda/\mu$ has at most one box in each row.
The coefficient is
\[
  2^{\ell(\mu)}(-1)^{\operatorname{col}(\lambda/\mu)}
  \left(-\frac{\beta}{2}\right)^{|\lambda/\mu|},
\]
where $\operatorname{col}(\lambda/\mu)$ is the number of occupied shifted
columns.  For $\mu=(3,2)$, the only strict partitions which contribute are
$(3,2)$, $(4,2)$, and $(4,3)$, so
\[
  \schurKQ_{(3,2)}
  =4\schurKP_{(3,2)}
   +2\beta\schurKP_{(4,2)}
   -\beta^2\schurKP_{(4,3)}.
\]
The negative term occurs because the two boxes of the shifted skew diagram
$(4,3)/(3,2)$ lie in the same shifted column.  In contrast, every part of
$(4,2)$ differs from the next by at least two, and the same rule gives only
nonnegative $\beta$-coefficients:
\[
  \schurKQ_{(4,2)}
  =4\schurKP_{(4,2)}
   +2\beta\schurKP_{(4,3)}
   +2\beta\schurKP_{(5,2)}
   +\beta^2\schurKP_{(5,3)}.
\]
\end{example*}

\subsection[schurKPQProps]{Properties}

The fundamental quasisymmetric expansion of $\schurKP_{\lambda}(\xvec)$
and $\schurKQ_{\lambda}(\xvec)$ can be found in \cite{HamakerKeilthyPatriasWebsterZhangZhou2017}.

\name{Takahiko Nobukawa} and \name{Tatsushi Shimazaki} compute special values
of skew $K$-theoretic Schur $P$- and $Q$-functions
\cite{NobukawaShimazaki2024x}.  For strict partitions $\lambda \supset \mu$,
their main evaluation gives
$\schurKP_{\lambda/\mu}(\beta,\dotsc,\beta\mid -\beta^{-1})
= \schurKQ_{\lambda/\mu}(\beta,\dotsc,\beta\mid -\beta^{-1})
= \beta^{|\lambda/\mu|}$ in their notation.  The proof uses sign-reversing
involutions on shifted set-valued tableaux and implies oddness results for the
corresponding tableau counts.


\subsection[schurKPQLRRule]{Littlewood--Richardson rule}

The Littlewood--Richardson rule for $\schurKP_{\lambda}(\xvec)$ is due to 
Clifford--Thomas--Yong \cite{CliffordThomasYong2014}.
Their result is based on a Pieri rule due to \name{Anders Buch} and \name{Vijay Ravikumar}, \cite{BuchRavikumar2012}.

\begin{problem}[Littlewood--Richardson rule]
Find a Littlewood--Richardson rule for $\schurKQ_{\lambda}(\xvec)$.
\end{problem}
