\metatitle{Specht modules}
\metadescription{An introduction to Specht modules, their construction using
tableaux, and their role in representing the symmetric group.}
\metakeywords{Specht modules,symmetric group,tableaux,row-stabilizer,column-stabilizer,polytabloid}

\section[spechtModules]{Specht modules}


\subsection[spechtModulesIntro]{Introduction to Specht modules}

The \defin{Specht module} $S^\lambda$ is the basic irreducible
$\symS_n$-module attached to a partition $\lambda\vdash n$.
Over $\setC$, the modules $S^\lambda$, as $\lambda$ ranges over all
partitions of $n$, form a complete set of pairwise non-isomorphic irreducible
$\symS_n$-modules.  Their characters are sent by the
\hyperref[frobeniusCharacteristic]{Frobenius characteristic} to the
\hyperref[schurFunction]{Schur functions}:
\[
  \frobChar(S^\lambda)=\schurS_\lambda.
\]
The construction below is the standard tableau construction; see
\cite{Fulton1997,Sagan2001,JamesKerber1984}.

\subsection[spechtTabloids]{Tabloids}

In order to construct an $\symS_n$-module, we first need a vector space on
which $\symS_n$ acts.  A \defin{tableau} of shape $\lambda$ is a filling of
the Young diagram $\lambda$ with $\{1,2,3,\dotsc,n\}$, using each entry once.
The \defin{row-stabilizer} $R_T$ of a tableau $T$ is the set of permutations
that keep entries within rows.  Similarly, the
\defin{column-stabilizer} $C_T$ of $T$ is the set of permutations that keep
entries within columns.

For example, the following tableaux are in the same orbit under $R_T$, and
there are $3! \cdot 2!$ permutations total in $R_T$.
\begin{figure}
\begin{ytableau}
1 & 2 & 4 \\
3 & 5
\end{ytableau}
\begin{ytableau}
4 & 1 & 2 \\
3 & 5
\end{ytableau}
\end{figure}

Similarly, the tableaux which can be obtained from the first one by some
element of $C_T$ are the following.
\begin{figure}
\begin{ytableau}
1 & 2 & 4 \\
3 & 5
\end{ytableau}
\begin{ytableau}
3 & 2 & 4 \\
1 & 5
\end{ytableau}
\begin{ytableau}
1 & 5 & 4 \\
3 & 2
\end{ytableau}
\begin{ytableau}
3 & 5 & 4 \\
1 & 2
\end{ytableau}
\end{figure}

Two tableaux are considered equivalent if they are in the same orbit under
the row-stabilizer.  An equivalence class is called a
\defin{tabloid}\label{tabloidDefinition} and is denoted by $\{T\}$.
Equivalently, a tabloid remembers the set of entries in each row, but forgets
the order inside each row.

The \defin{permutation module} $M^\lambda$ is the $\setC$-vector space with
basis all tabloids of shape $\lambda$.  The symmetric group acts by permuting
the entries:
\[
  \sigma \{T\} = \{\sigma T\}.
\]
This action is well-defined because row-equivalent tableaux remain
row-equivalent after applying $\sigma$.

\subsection[polytabloids]{Polytabloids}

A \defin{polytabloid}\label{polytabloidDefinition} is obtained by
antisymmetrizing a tabloid in its columns.  For a tableau $T$, define
\[
  e_T
  \coloneqq
  \sum_{\tau\in C_T}\sign(\tau)\{\tau T\}.
\]
The \defin{Specht module}\label{spechtModuleClassicalDefinition}
of shape $\lambda$ is
\[
  S^\lambda \coloneqq
  \langle e_T : T\text{ is a tableau of shape }\lambda\rangle_{\setC}
  \subseteq M^\lambda.
\]
Since $\sigma e_T=e_{\sigma T}$, the space $S^\lambda$ is stable under the
action of $\symS_n$.

\begin{example}
Let $\lambda=(2,1)$ and let $T$ be the tableau
\begin{figure}
\begin{ytableau}
1 & 2\\
3
\end{ytableau}
\end{figure}
The column-stabilizer is $C_T=\{e,(13)\}$, so the corresponding polytabloid is
\[
  e_T
  =
  \left\{\begin{array}{cc}
  1&2\\
  3
  \end{array}\right\}
  -
  \left\{\begin{array}{cc}
  2&3\\
  1
  \end{array}\right\}.
\]
The row order is not part of the tabloid data: for instance
\[
  \left\{\begin{array}{cc}1&2\\3\end{array}\right\}
  =
  \left\{\begin{array}{cc}2&1\\3\end{array}\right\}.
\]
For this shape, $S^{(2,1)}$ is the standard two-dimensional irreducible
representation of $\symS_3$.
\end{example}

\subsection[spechtFrobenius]{Frobenius characteristic}

The Specht modules explain why Schur functions are the natural character
basis for $\symS_n$-representations.  If $\chi^\lambda$ is the irreducible
character of $S^\lambda$, then
\[
  \frobChar(\chi^\lambda)=\schurS_\lambda.
\]
Consequently, if an $\symS_n$-module decomposes as
\[
  M \cong \bigoplus_{\lambda\vdash n}
  (S^\lambda)^{\oplus c_\lambda},
\]
then its Frobenius characteristic is
\[
  \frobChar(M)=\sum_{\lambda\vdash n} c_\lambda\schurS_\lambda.
\]
Thus Schur positivity is the statement that a symmetric function is the
Frobenius characteristic of an honest $\symS_n$-module, or at least behaves
as if it were one.

The dimension of $S^\lambda$ is the number $f^\lambda$ of standard Young
tableaux of shape $\lambda$, given by the
\hyperref[hookFormula]{hook formula}.

\subsection[spechtZeroHeckeComparison]{Comparison with the 0-Hecke picture}

The \hyperref[zeroHeckeAlgebra]{0-Hecke algebra} gives a useful parallel to
the classical Specht-module story.  The group algebra $\setC[\symS_n]$ has
irreducible modules indexed by partitions, and its Frobenius characteristic
lands in symmetric functions.  The 0-Hecke algebra $H_n(0)$ has simple
modules indexed by compositions, and its quasisymmetric characteristic lands
in \hyperref[quasiSymmetricFunctions]{quasisymmetric functions}.

\[
\begin{array}{lll}
\toprule
\text{Feature} & \setC[\symS_n] & H_n(0)\\
\midrule
\text{Simple modules} & S^\lambda,\ \lambda\vdash n
  & F^\alpha,\ \alpha\vDash n\\
\text{Characteristic} & \frobChar(S^\lambda)=\schurS_\lambda
  & \mathrm{Ch}(F^\alpha)=\gessel_\alpha\\
\text{Indexing objects} & \text{partitions and standard tableaux}
  & \text{compositions and descent sets}\\
\text{Dual Hopf algebra} & \spaceSym & \mathrm{NSym}\\
\bottomrule
\end{array}
\]

This dictionary is one reason that the
\hyperref[qsymNSymDuality]{QSym--NSym duality} is best understood together
with ordinary representation theory of $\symS_n$.

\subsection[spechtRecentWork]{Related directions}

For related work, \name{Rowena Paget} and \name{Mark Wildon} construct
homomorphisms from Specht modules into Foulkes modules \cite{PagetWildon2011}.
\name{Reuven Hodges} and \name{Hanzhang Yin} give a non-iterative
straightening algorithm for skew Schur modules, a closely related
tableau-based construction \cite{HodgesYin2025x}.

\name{Jesse Kim} constructs a rotation-invariant web basis for the Specht
modules $S^{(d^3,1^{n-3d})}$, indexed by normal plabic graphs or augmented
$\mathrm{SL}_3$ webs \cite{Kim2024x}.  The construction gives skein
relations for the $\symS_n$-action and leads to a cyclic-sieving result for
the corresponding $q$-hook length formula.

\name{Maria Gillespie} studies higher Specht polynomials in two sets of
variables under the diagonal action of $\symS_n$ \cite{Gillespie2024x}.  In
particular, she constructs a higher Specht basis for Garsia--Haiman modules
for hook shapes.  In a complementary direction, \name{Shaul Zemel}
generalizes higher Specht polynomials to homogeneous representations and
quotients related to the rings $R_{n,k,s}$ \cite{Zemel2025x}.
