\metatitle{Stable polynomials}

\metadescription{Multivariate stable polynomials, strongly Rayleigh measures, stability-preserving operations, and same-phase stability.}


\section[stablePolynomial]{Stable polynomials}



Let $\mathcal{H} \subset \setC$ denote the
upper half-plane $\{z \in \setC: \mathrm{im}(z) > 0 \}$.

\begin{definition}[Stability]

A multivariate polynomial $P \in \setC[z_1,\dotsc,z_n]$
is called \defin{stable}\label{stablePolynomialDefinition}
if it does not vanish on $\mathcal{H}^n$.
That is, $P$ is stable if
\[
 \zvec^* \in \mathcal{H}^n \implies P(\zvec^*) \neq 0.
\]
One can easily show that $P \in \setR[z_1,\dotsc,z_n]$ is stable if and only
if $P(\alpha+\lambda t)=0$ has only real zeros for all
$\alpha \in \setR^n$, $\lambda \in \setR_+^n$.
A univariate polynomial with real coefficients
is stable if and only if all its roots are real.
\end{definition}





\begin{definition}[Strongly Rayleigh]\label{stronglyRayleigh}
A measure $\mu$ on the set of subsets of $[n]$ is called
\defin{strongly Rayleigh} if
\[
 \sum_{S \subseteq [n]} \mu(S) \prod_{s\in S} x_s
\]
is real stable.
\end{definition}


\subsection[stabilityOperators]{Operators preserving stability}

\begin{theorem}[Stability preserving operators]
The following operations on polynomials preserve stability.
Let $p,q \in \setR[z_1,\dotsc,z_n]$ be stable polynomials.
Then the following are also stable.
\begin{itemize}
\item $p\cdot q$ (Product)
\item $\partial_{z_j} p$ (Differentiation)
\item $z_j \partial_{z_j} p$ (Degree-preserving differentiation)
\item $p(z_1,\dotsc,z_{j-1},r,z_{j+1},\dotsc,z_n)$ (Real specialization)
\item $p(z_1,\dotsc,z_{j-1},z_1,z_{j+1},\dotsc,z_n)$ (Projection)
\item $z_j^d p(z_1,\dotsc,z_{j-1},-1/z_{j},z_{j+1},\dotsc,z_n)$ (Inversion)
\end{itemize}
Here, $d$ is the degree of $z_j$ in $p$.
\end{theorem}


\subsection[borceaBrandenSymbols]{Borcea--Brändén symbols}

The Borcea--Brändén classification packages a linear operator into a
polynomial, or into an entire power series, called its
\defin{symbol}\label{stabilitySymbol}.  This is often the most practical way
to prove that an operator preserves stability: instead of testing the operator
on all stable inputs, one tests one universal object.  For real stability one
must allow both signs.  For instance,
\(f(z)\mapsto f(-z)\) preserves real-rootedness, but it is detected by the
minus symbol rather than the plus symbol.

\begin{definition}[Algebraic symbols]
Let $\mathbf{d}=(d_1,\dotsc,d_n) \in \setN^n$, and let
$\setR_{\leq \mathbf{d}}[z_1,\dotsc,z_n]$ denote the polynomials whose degree
in $z_j$ is at most $d_j$.
For a linear operator
\[
 T:\setR_{\leq \mathbf{d}}[z_1,\dotsc,z_n]\to \setR[x_1,\dotsc,x_m],
\]
define the two \defin{algebraic symbols}\label{algebraicSymbol} of $T$ by
\[
 G_T^{+}(\xvec,\mathbf{w})
 \coloneqq
 T\left( \prod_{j=1}^n (z_j+w_j)^{d_j} \right),
 \qquad
 G_T^{-}(\xvec,\mathbf{w})
 \coloneqq
 T\left( \prod_{j=1}^n (z_j-w_j)^{d_j} \right).
\]
Here $T$ acts on the $z$-variables and the $w_j$'s are auxiliary variables.
\end{definition}

\begin{theorem}[Borcea--Brändén finite-degree classification, see \cite{BorceaBranden2009I}]
\label{borceaBrandenFiniteSymbol}
Let \(T:\setR_{\leq \mathbf{d}}[z_1,\dotsc,z_n]\to
\setR[x_1,\dotsc,x_m]\) be linear.  Then \(T\) sends every real stable
polynomial to a real stable polynomial or to zero if and only if one of the
following holds.
\begin{itemize}
\item \(G_T^{+}\) is real stable.
\item \(G_T^{-}\) is real stable.
\item The range of \(T\) has dimension at most two and can be written as
  \(T(f)=\alpha(f)P+\beta(f)Q\), where \(\alpha,\beta\) are real linear
  functionals and \(P,Q\) are real stable polynomials in proper position.
  In one common convention this means that \(Q+iP\) is stable.
\end{itemize}
\end{theorem}

For operators over complex coefficients, the corresponding stability-preserver
classification has only the plus symbol, apart from the rank-one exceptional
case.  The extra sign in the real theorem records the distinction between
preserving the upper half-plane and preserving real-rootedness.

\begin{example*}[Two elementary symbols]
For \(T=\partial_{z_i}\), the plus symbol is
\[
  G_T^{+}(\xvec,\mathbf{w})
  =
  d_i(x_i+w_i)^{d_i-1}
  \prod_{j\neq i}(x_j+w_j)^{d_j},
\]
which is stable.  This recovers the stability-preserving property of
partial differentiation.

For the univariate reflection \(T(f)(x)=f(-x)\) on
\(\setR_{\leq d}[z]\), the plus symbol \((w-x)^d\) is not stable, but the
minus symbol
\[
  G_T^{-}(x,w)=(-x-w)^d
\]
is stable.  Hence the sign alternative is genuinely needed for real
stability preservers.
\end{example*}

\begin{definition}[Laguerre--Pólya class and transcendental symbol]
Let \(\mathcal{LP}_r\) be the
\defin{Laguerre--Pólya class}\label{laguerrePolyaClass}
of entire functions in \(r\) variables
that are limits, uniformly on compact sets, of real stable polynomials.
This is the multivariate Laguerre--Pólya class.

For a linear operator
\[
  T:\setR[z_1,\dotsc,z_n]\to \setR[x_1,\dotsc,x_m],
\]
its \defin{transcendental symbol}\label{transcendentalSymbol}
is the formal power series
\[
  \overline{G}_T(\xvec,\mathbf{w})
  \coloneqq
  T\!\left(e^{-\zvec\cdot\mathbf{w}}\right)
  =
  \sum_{\alpha\in\setN^n}
    (-1)^{|\alpha|}
    T(\zvec^\alpha)
    \frac{\mathbf{w}^{\alpha}}{\alpha!}.
\]
\end{definition}

\begin{theorem}[Borcea--Brändén transcendental classification, see \cite{BorceaBranden2009I}]
\label{borceaBrandenTranscendentalSymbol}
Up to the same rank-at-most-two exception as above, a linear operator
\[
  T:\setR[z_1,\dotsc,z_n]\to \setR[x_1,\dotsc,x_m]
\]
preserves real stability if and only if
\[
  \overline{G}_T(\xvec,\mathbf{w})\in \mathcal{LP}_{m+n}
  \qquad\text{or}\qquad
  \overline{G}_T(\xvec,-\mathbf{w})\in \mathcal{LP}_{m+n}.
\]
\end{theorem}

This is the infinite-degree analogue of the algebraic-symbol test.  It also
explains why the classical \hyperref[multiplierSequences]{multiplier
sequence} theorem has an exponential generating function.  If
\[
  T_\gamma(z^k)=\gamma_k z^k,
\]
then
\[
  \overline{G}_{T_\gamma}(x,w)
  =
  \sum_{k\geq 0}\gamma_k\frac{(-xw)^k}{k!}
  =
  \Phi_\gamma(-xw).
\]
Thus the Pólya--Schur condition on \(\Phi_\gamma\) is the diagonal-operator
case of the Borcea--Brändén symbol criterion.  Compare this with
\hyperref[polyaFrequency]{Pólya frequency sequences}: PF sequences are
ordinary coefficient sequences controlled by Toeplitz total positivity,
whereas multiplier sequences are diagonal operators controlled by exponential
symbols.

\name{Ming-Jian Ding} and \name{Bao-Xuan Zhu} give a systematic treatment of
stability for combinatorial polynomials \cite{DingZhu2021Stability}.  They
apply Borcea--Br\"and\'en type characterizations to real and Hurwitz stability,
Tur\'an expressions, alternating-runs polynomials, and several interlacing and
alternatingly-increasing phenomena.

\name{Hao Fang} and \name{Biao Ma} introduce
\defin{G{\aa}rding polynomials}\label{gardingPolynomial}, a class strictly
containing real stable
polynomials \cite{FangMa2026x}.  The definition is phrased in terms of
positivity regions invariant under translation by positive vectors; they prove
equivalent formulations by polarization and by a recursive criterion using
partial derivatives.  Multi-affine G{\aa}rding polynomials with non-negative
coefficients satisfy the Rayleigh property, and their positive univariate
specializations have ultra-log-concave coefficient sequences.



\subsection[stabilitymultiaffine]{Multi-affine stability}

A polynomial $P(z_1,\dotsc,z_n)$ is
\defin{multi-affine}\label{multiAffinePolynomial}
if the degree in each variable is at most $1$.



\begin{theorem}[See \cite{Branden2007}]
A homogeneous multi-affine (real) polynomial $f(x_1,\dotsc,x_n)$ is real
stable if and only if
\[
 \frac{\partial f}{\partial x_i} \cdot \frac{\partial f}{\partial x_j} -
 \frac{\partial^2 f}{\partial x_i \partial x_j} \cdot f \geq 0
\]
whenever $x_1,x_2,\dotsc,x_n \in \setR$.
\end{theorem}
A polynomial satisfying this inequality is said to have the
\defin{strongly Rayleigh property}\label{stronglyRayleighProperty}.
The \defin{weak Rayleigh property}\label{weakRayleighProperty}
holds if the above inequality is true whenever the variables are positive
real numbers.


\begin{theorem}[Grace--Walsh--Szeg\H{o}, see \cite{Grace1902,Szego1922,Walsh1922,RahmanSchmeisser2002}]
Let $P(z_1,\dotsc,z_n)$ be a \emph{symmetric} multi-affine polynomial.
Then $P$ is stable if and only if the univariate diagonal specialization
$P(t,\dotsc,t)$ is stable.
If $P$ has real coefficients, this is equivalent to the univariate
specialization $P(t,\dotsc,t)$ being real-rooted.
\end{theorem}

\begin{definition}[Apolar polynomials]
Let
\[
  F(z)=\sum_{k=0}^n \binom{n}{k} a_k z^k
  \qquad\text{and}\qquad
  G(z)=\sum_{k=0}^n \binom{n}{k} b_k z^k.
\]
The polynomials $F$ and $G$ are \defin{apolar}\label{apolarPolynomial} if
\[
  \sum_{k=0}^n (-1)^k \binom{n}{k} a_k b_{n-k}=0.
\]
\end{definition}

\begin{theorem}[Grace's apolarity theorem, see \cite{Grace1902,Szego1922,RahmanSchmeisser2002}]
Let $C$ be a circular domain, and let $F$ and $G$ be apolar
polynomials of degree $n$.  If all zeros of $F$ lie in $C$, then $G$
has at least one zero in $C$.
\end{theorem}

A useful way to reduce questions to the multi-affine case is
\defin{polarization}\label{polarizationDefinition}.
If $0 \leq k \leq d_j$, set
\[
 \operatorname{Pol}_{d_j}(z_j^k)
 \coloneqq \binom{d_j}{k}^{-1}
 \elementaryE_k(z_{j,1},\dotsc,z_{j,d_j}).
\]
For monomials, define
\[
 \operatorname{Pol}_{\mathbf{d}}(\zvec^\alpha)
 \coloneqq
 \prod_{j=1}^n \operatorname{Pol}_{d_j}(z_j^{\alpha_j}),
\]
and extend linearly to polynomials.  This gives the polarization
$\operatorname{Pol}_{\mathbf{d}}(P)$ of a polynomial whose degree in $z_j$
is at most $d_j$.

\begin{theorem}[Polarization, see \cite{BorceaBranden2009II}]
Let $P \in \setC[z_1,\dotsc,z_n]$ have degree at most $d_j$ in the variable
$z_j$.  Then $P$ is stable if and only if
$\operatorname{Pol}_{\mathbf{d}}(P)$ is stable.
\end{theorem}


\section[stabilityConnections]{Jump systems and matroids}


For $P = \sum_{\alpha} c_\alpha \zvec^\alpha$, let
$\supp(P) \coloneqq \{\alpha : c_\alpha \neq 0\}$.

\begin{theorem}[See \cite{Branden2007}]
If $P$ is stable, then $\supp(P)$ is a jump system.
\end{theorem}


\begin{theorem}[See \cite{ChoeOxleySokalWagner2004}]
If $P$ is homogeneous and multi-affine stable,
then its support is the set of bases of a \hyperref[matroid]{matroid}.
See also \cite[Cor. 3.4]{Branden2007}.
\end{theorem}
The converse is in general not true but some classes of \hyperref[matroid]{matroids}
always have a \hyperref[basisGeneratingPolynomial]{basis-generating polynomial}
which is stable.
For example, \hyperref[matroidRegular]{regular matroids}
(which include \hyperref[matroidGraphic]{graphic matroids}),
\hyperref[matroidUniform]{uniform matroids} and
a subclass of \hyperref[transversalMatroids]{transversal matroids}
(see \cite{ChoeOxleySokalWagner2004}).


\hyperref[matroidOperations]{Deletion and contraction} on matroids preserve
stability of basis-generating polynomials.

For a connected graph $G$ on vertex set $V$, the spanning-tree degree
enumerator
\[
  \sum_T \prod_{v \in V} x_v^{\deg_T(v)},
\]
where the sum is over all spanning trees $T$ of $G$, is real stable if and
only if $G$ is distance-hereditary \cite{CherkashinPetrovProzorov2023}.


\begin{theorem}[See \cite{BrandenHuh2020}]
Let $P \in \setR[z_1,\dotsc,z_n]$ be homogeneous, real stable, and have
non-negative coefficients.  Then $P$ is \hyperref[lorentzianPolynomial]{Lorentzian}.
\end{theorem}


\begin{theorem}[Determinantal stability, see \cite[Prop. 2.4]{BorceaBranden2008MixedDeterminants}]
Let $A_0,A_1,\dotsc,A_n$ be real symmetric matrices, and suppose that
$A_1,\dotsc,A_n$ are positive semidefinite.  Then
\[
 \det(A_0+z_1A_1+\dotsb+z_nA_n)
\]
is real stable or identically zero.
\end{theorem}


\begin{theorem}[See \cite{BorceaBranden2010}]
A polynomial in two variables $P(z,w) \in \setR[z,w]$ is stable if and only
if it can be expressed as $\pm \det(zA+wB+C)$, where $A,B,C$ are real
$n \times n$ matrices, $A$ and $B$ are positive semidefinite, and $C$ is
symmetric.
\end{theorem}


\begin{example*}[The monotone column permanent conjecture]

Let $A$ be a real $n\times n$-matrix in which entries along columns decrease.
Let $Z_n = \mathrm{diag}(z_1,\dotsc,z_n)$ be a diagonal matrix with indeterminates,
and let $J_n$ be the matrix with all entries equal to $1$.
Then
\[
 \mathrm{per}(J_nZ_n + A) = \mathrm{per}\left( (z_j + a_{ij})_{1 \leq i,j \leq n} \right)
\]
is stable, see \cite{BrandenHaglundVisontaiWagner2011}.
By noting that the \hyperref[ex:eulerianSturm]{Eulerian polynomials}
are generated by excedances,
they give a new proof that (a multivariate version of) Eulerian polynomials
are stable.
\end{example*}


\begin{example*}[Multivariate Eulerian polynomials]
\label{ex:stableMultivariateEulerianOperator}
\name{Petter Brändén} (see \cite[Thm. 2.5]{HaglundVisontai2012}) proved
stability of the following multivariate refinement of Eulerian polynomials.
Set
\[
 \widetilde A_n(\xvec,\yvec)
 =
 \sum_{\pi \in \symS_n}
 \prod_{\substack{1\leq i<n\\ \pi_i \gt \pi_{i+1}}} x_{\pi_i}
 \prod_{\substack{1\leq i<n\\ \pi_i \lt \pi_{i+1}}} y_{\pi_{i+1}}.
\]
Thus descent tops receive \(x\)-variables and ascent tops receive
\(y\)-variables.  The proof uses a stability-preserving operator.  If
\[
  \partial
  =
  \sum_{i=1}^n \frac{\partial}{\partial x_i}
  +
  \sum_{i=1}^n \frac{\partial}{\partial y_i},
\]
then
\[
  T_{n+1}
  \coloneqq
  (x_{n+1}+y_{n+1}) + x_{n+1}y_{n+1}\partial
\]
preserves real stability on multi-affine polynomials, by the
\hyperref[borceaBrandenFiniteSymbol]{Borcea--Brändén symbol criterion}.
Equivalently, if
\[
  U=\prod_{i=1}^n (x_i+u_i)(y_i+v_i),
\]
then the algebraic symbol
\[
  T_{n+1}(U)
  =
  \left(
    x_{n+1}+y_{n+1}
    +x_{n+1}y_{n+1}
    \sum_{i=1}^n
    \left(\frac{1}{x_i+u_i}+\frac{1}{y_i+v_i}\right)
  \right)U
\]
is stable.
The recurrence
\[
  \widetilde A_{n+1}(\xvec,\yvec)
  =
  T_{n+1}\widetilde A_n(\xvec,\yvec)
\]
then proves stability by induction from \(\widetilde A_1=1\).
Specializing \(y_i=1\) and then \(x_i=t\) gives
\[
  t\,\widetilde A_n(t,\dotsc,t;1,\dotsc,1)=A_n(t),
\]
so the classical Eulerian polynomial has only real roots.

In fact, \name{Jim Haglund} and \name{Mirko Visontai} also treat Stirling
permutations, where plateaus occur.  If \(Q_n\) is the set of Stirling
permutations and \(D(\sigma),A(\sigma),P(\sigma)\) are the descent-top,
ascent-top, and plateau sets, then
\[
 C_n(\xvec,\yvec,\zvec)
 =
 \sum_{\sigma \in Q_n}
 \prod_{i\in D(\sigma)} x_{\sigma_i}
 \prod_{i\in A(\sigma)} y_{\sigma_i}
 \prod_{i\in P(\sigma)} z_{\sigma_i}
\]
is stable.  Specializing \(y_i=z_i=1\) recovers real-rootedness of the
second-order Eulerian polynomials.
\end{example*}




\begin{example*}[Multivariate matching polynomials]
The multivariate \hyperref[realRootedMatching]{matching polynomial} is real
stable, see \cite{BorceaBranden2009I,BorceaBranden2009II}.
See also \cite{LeakeRyder2019}; there are two types of multivariate matching
polynomials, recording either vertex labels or edge labels in the matchings:
\begin{align}
\mu_V(G;x_1,\dotsc,x_n)
  & \coloneqq
  \sum_{\substack{M \subseteq E(G) \\ \text{$M$ matching}}}
  \prod_{uv \in M} -x_u x_v, \\
\mu_E(G;x_1,\dotsc,x_n)
  & \coloneqq
  \sum_{\substack{M \subseteq E(G) \\ \text{$M$ matching}}}
  \prod_{e \in M} x_e.
\end{align}
\end{example*}

The \defin{multi-affine vertex matching polynomial}
\label{multiAffineVertexMatchingPolynomial}
$\mu_V(G;x_1,\dotsc,x_n)$ is real stable, see \cite{BorceaBranden2009I}.

\name[D.G. Wagner]{David Wagner}'s survey \cite{Wagner2009StableSurvey}
is a useful reference for the theory and applications of multivariate
stable polynomials, including the Borcea--Brändén classification of
stability preservers, matrix applications, and probabilistic applications.

\begin{theorem}[Strongly Rayleigh measures are negatively associated, see
\cite{Wagner2009StableSurvey}]
Let $\mu$ be a probability measure on subsets of $[n]$ whose generating
polynomial
\[
  \sum_{S\subseteq[n]}\mu(S)\prod_{i\in S}x_i
\]
is real stable.  Then every measure obtained from $\mu$ by applying
positive external fields is negatively associated: increasing functions
depending on disjoint coordinate sets have non-positive covariance.
\end{theorem}




\subsection[samePhaseStable]{Same phase stability}

The following notion was introduced in \cite{LeakeRyder2019}
and is a strictly weaker notion than stability.

\begin{definition}[Same-phase stability]
 A polynomial $P(z_1,\dotsc,z_n) \in \setR[z_1,\dotsc,z_n]$ is said
 to be \defin{same-phase stable}\label{samePhaseStabilityDefinition}
 if for every $\lambda \in \setR_+^n$, the univariate polynomial
 $P(\lambda_1 t,\lambda_2 t,\dotsc,\lambda_n t) \in \setR[t]$ is real-rooted.
\end{definition}

A family of polynomials $\{p_j(\xvec) \}_{j=1}^m$ is
\defin{same-phase compatible}\label{samePhaseCompatible}
if for every vector of positive numbers $\lambda$, the univariate polynomials
$\{p_j(\lambda t) \}_{j=1}^m$ are
\hyperref[compatiblePolynomials]{compatible}.


Let $P, P_0, \dotsc, P_m$ be polynomials such that
$P = P_0 + \sum_{j=1}^m z_{i_j} P_{j}$.
This is a \defin{proper splitting of $P$}\label{properSplitting}
if none of the polynomials $P_j$ depends on the variables
$\{z_{i_1}, \dotsc, z_{i_m} \}$.


\begin{theorem}[See \cite{LeakeRyder2019}]
Suppose $P$ is a multi-affine polynomial with non-negative coefficients.
Then the following are equivalent:
\begin{itemize}
\item $P$ is same-phase stable;
\item For any proper splitting,
$P = P_0 + \sum_{j=1}^m z_{i_j} P_{j}$,
the polynomials $P_0, z_{i_1} P_{1}, \dotsc, z_{i_m} P_{m}$
are same-phase compatible.

\item There is some proper splitting
$P = P_0 + \sum_{j=1}^m z_{i_j} P_{j}$
such that $P_0, z_{i_1} P_{1}, \dotsc, z_{i_m} P_{m}$ are same-phase
compatible.

\end{itemize}

\end{theorem}


The \defin{multi-affine edge matching polynomial}
\label{multiAffineEdgeMatchingPolynomial}
$\mu_E(G;x_1,\dotsc,x_n)$ is same-phase stable \cite[Cor. 3.2]{LeakeRyder2019}.

Let $G$ be a graph on the vertex set $[n]$, and define the
\defin{multivariate independence polynomial}\label{multivariateIndependencePolynomial}
\[
I(G;x_1,\dotsc,x_n)
\coloneqq
\sum_{\substack{A \subseteq V(G) \\ A \text{ independent}}}
\prod_{v \in A} x_v.
\]
In \cite{LeakeRyder2019}, it is proved that $I(G;x_1,\dotsc,x_n)$
is same-phase stable \emph{if and only if} $G$ is claw-free.
This is a reinterpretation of Engström's weighted result \cite{Engstrom2007},
but Leake and Ryder give a much sleeker proof, and the if-and-only-if
condition explains
that the claw-free requirement is sharp.


\begin{example*}[Multivariate Eulerian polynomials II]
In \cite{AlexanderssonNabawanda2021}, we show that the following generalization
of the Eulerian polynomial is same-phase stable, but not stable:
\[
 \tilde{A}_n(\xvec) = \sum_{\pi \in \symS_n} \prod_{\pi_i \gt \pi_{i+1}} x_{i}.
\]
Note that we now keep track of the descent index, rather than the descent
bottom value.
Moreover, we showed that
\[
 \tilde{A}_{n-1}(\lambda_1 t, \dotsc, \lambda_{n-1} t)
 \interl
 \tilde{A}_n(\lambda_1 t, \dotsc, \lambda_n t)
\]
for all positive real numbers $\lambda_1,\dotsc,\lambda_n$.
\end{example*}
