\metatitle{Todo}
\metadescription{All topics and families of polynomials that are to be added at some point.}


%Papers citing SYMCAT: 
% https://link.springer.com/article/10.1007/s13226-021-00185-7
% https://link.springer.com/article/10.1134/S0016266321030023
% https://www.sciencedirect.com/science/article/pii/S0196885821001470
% https://doi.org/10.1093/imrn/rnab165
% https://arxiv.org/pdf/2004.03928.pdf
% http://homepages.ulb.ac.be/~anvdwyng/pdfs/Delta%20conjectures%20and%20Theta%20refinements.pdf
% http://www-math.mit.edu/~rstan/ec/ch7suppsol.pdf
% \cite{CorteelGitlinKeatingMeza2021}
% https://www.cip.ifi.lmu.de/~grinberg/algebra/petriesym.pdf
% \cite{Grinberg2020x}


\section[metaWriteFamilies]{Missing families of functions}

\begin{itemize}

\item \url{https://arxiv.org/pdf/1203.4465.pdf}  Strong Schur functions (similar to k-Schur)

\item  \url{http://www.lorentzcenter.nl/lc/web/2010/423/presentations/Noumi.pdf} Elliptic Schur

\item \url{https://pdfs.semanticscholar.org/4e50/3a263890c8d3b0fbcc49d935e21cf1f8be50.pdf}
\url{http://doc.sagemath.org/html/en/reference/combinat/sage/combinat/sf/witt.html} Witt symmetric functions,


% \url{https://idea.library.drexel.edu/islandora/object/idea%3A4274/datastream/OBJ/download/Combinatorial_aspects_of_generalizations_of_Schur_functions.pdf}
\item  Factorial Schur

\item \url{http://fpsac2019.fmf.uni-lj.si/resources/Posters/34poster.pdf} All these K-theoretic stuff

\item  \url{https://alco.centre-mersenne.org/item/ALCO_2019__2_4_447_0/} More K-k stuff

\item Rational Schubert, key, Grothendieck \url{https://dergipark.org.tr/en/download/article-file/128037}
\item Something about Schubert structure constants \url{https://sites.math.washington.edu/~billey/classes/schubert.library/bergeron.sottile.pdf}
\item Type B-q-Schur functions, Chow's qsym funcs and LLTs:
\url{http://fpsac2019.fmf.uni-lj.si/resources/Proceedings/159.pdf}


\item Flag-symmetric functions (connections with Posets, Kostka-polynomials and Hecke algebra).
\url{http://www-math.mit.edu/~rstan/pubs/pubfiles/98.pdf}


\end{itemize}

\section[metaWriteTopics]{Topics to be covered}


\begin{itemize}


\item Kazhdan--Lusztig polynomials

\item 
	Make page with resources (oeis, Findstat, doi2bib, latex tips).
	Link Viennot videos: \url{http://www.viennot.org/abjc-course.html}

\item More Cauchy identities for Hall--Littlewood functions, \url{http://www.labri.fr/perso/marckert/wheeler.pdf}

\item Grassmannian and Schubert cells.

\end{itemize}





\section[metaPapers]{Misc papers on algebraic combinatorics}


\begin{enumerate}

\item \url{https://www.ams.org/journals/mcom/0000-000-00/S0025-5718-2025-04166-8/} CSP on Findstat

\item \url{https://escholarship.org/uc/combinatorial_theory/5/3} Many nice papers here.

\item \url{https://doi.org/10.5070/C65365563} LLT polynomials, partition function, lattice paths.
\item \url{https://arxiv.org/pdf/2504.05123} Order polytopes, Ehrhart, and some nice open problems

\item \url{https://arxiv.org/pdf/2503.19344} Chromatic symmetric glued at a vertex, e-positivity.


\item \url{https://arxiv.org/pdf/2502.15586} Symplectic and odd orthogonal characters, JT, Schur

\item \url{https://arxiv.org/pdf/2112.09633v2} 

\item \url{http://arxiv.org/pdf/2405.01166v3} Cycle-chords are e-pos

\item \url{https://arxiv.org/pdf/2411.13371} Fundamental quasisymmetric functions in superspace.

\item \url{https://arxiv.org/pdf/2411.10933} Dual flagged Weyl characters (generalizes Key and Schubert)

\item \url{https://arxiv.org/pdf/2411.11447} Symplectic Murnaghan--Nakayama rule

\item \url{https://arxiv.org/pdf/2411.02897} Real-rootedness, multi-dimensional permutations

\item \url{https://arxiv.org/pdf/2506.06951} RSK for type C, Berele insertion and King tableaux
\item \url{https://arxiv.org/pdf/2410.18343} Refined canonical stable grothendieck polynomials and tableaux

\item \url{https://arxiv.org/pdf/2410.13642} Macdonald stuff
\item \url{https://arxiv.org/pdf/2410.19189} Machine Learning Stanley-Stembridge

\item \url{https://arxiv.org/pdf/2410.07960} Kirillov polynomials, and Hecke-Grothendieck polynomials

\item \url{https://arxiv.org/pdf/2410.03916} Lascoux polynomials and Kohnert moves.
\item \url{https://arxiv.org/pdf/2211.02993.pdf} Hecke insertion, Grothendieck and Lascoux polys.
\item \url{https://arxiv.org/pdf/2302.04226.pdf} Key and Lascoux
\item \url{https://arxiv.org/pdf/2306.04159.pdf} Top Lascoux polynomials (very related to Schubert polys)
\item \url{https://arxiv.org/pdf/2302.03643.pdf} Grothendieck and Lascoux polynomials.
\item \url{https://arxiv.org/pdf/2312.01417.pdf} Lascoux polynomials from GT-patterns (and faces of GT polytopes).



\item \url{https://arxiv.org/pdf/2409.20478} Kneser Chromatic sym funcs

\item \url{https://arxiv.org/pdf/2409.17842} Skew hook formula via Yang-Baxter

\item \url{https://arxiv.org/pdf/2409.04621} Asymptotics of Macdonald and Jack symmetric functions.

\item \url{https://arxiv.org/pdf/2409.06175} Sn-action on some matrices, cool Frobenius series (MOXUAN (JASPER) LIU, YICHEN MA, BRENDON RHOADES, AND HAI ZHU)

\item \url{https://arxiv.org/pdf/2408.15111} Big descents and Schur positivity (nice conjectures!)


\item \url{https://arxiv.org/pdf/2408.07863} Puzzles, and interlacing triangles

\item \url{https://arxiv.org/pdf/2408.01390} K-theoretic key-thing

\item \url{https://arxiv.org/pdf/2407.10792} Delta springer varieties


\item \url{https://arxiv.org/pdf/2407.04810} super-symmetric Macdonald

\item \url{https://arxiv.org/pdf/2406.12751} Quasisymmetric Q-Schur expands positively into Young QSym Schur.

\item \url{https://arxiv.org/pdf/2406.05311} Quantum Murnaghan--Nakayama rule for Schubert polynomials

\item \url{https://arxiv.org/pdf/2405.13137} Schur-Q Pfaffian formulas

\item \url{https://arxiv.org/pdf/2404.04512.pdf} New method for schur positivity from Gessel expansion.

\item \url{https://arxiv.org/pdf/2402.16251.pdf} CSP on many things in Findstat. Alexandersson-Kebede map :)

\item \url{https://arxiv.org/pdf/2401.11060.pdf} Molev-Sagan rule for Double Schubert


\item \url{https://www.cambridge.org/core/journals/forum-of-mathematics-pi/article/skew-rsk-dynamics-greene-invariants-affine-crystals-and-applications-to-qwhittaker-polynomials/40AA4466D994D85B2BF1F6549E633F52}

\item \url{https://arxiv.org/pdf/2312.17469.pdf} Macdonald polys of type C.

\item \url{https://arxiv.org/pdf/2312.15409.pdf} Also crystals for set-valued stuff.
\item \url{https://arxiv.org/pdf/2312.16776.pdf} Crystals for set-valued decomp. tableaux.

\item \url{https://arxiv.org/pdf/2312.13675.pdf} Spin analog of plethysm MN-rule.
\item \url{https://arxiv.org/pdf/2212.08412.pdf} Vertex operators and a plethystic MN-rule.

\item \url{https://arxiv.org/pdf/2312.13681.pdf} Murnaghan-Nakayama and new iterative formula for Schur

\item \url{https://arxiv.org/pdf/2312.14348.pdf} 6-vertex model stuff



\item \url{https://arxiv.org/pdf/2312.02383.pdf} Homomesy on permutation rotation

\item \url{https://arxiv.org/pdf/2311.10659.pdf} Bender--Knuth involutions in type B and C.




\item \url{https://arxiv.org/pdf/2307.10852.pdf} Thm 4.10: real-rooted+pos Ehrhart -> real-rooted h*.
\item \url{https://arxiv.org/pdf/2303.09614.pdf} Weighted Ehrhart, cool result
\item \url{https://arxiv.org/pdf/2310.15730.pdf} Macdonald Murnaghan nakayama + lots of nice formulas

\item \url{https://arxiv.org/pdf/2310.17362.pdf} Intermediate Macdonald polynomials

\item \url{https://arxiv.org/pdf/2309.05903.pdf} Sturm sequences from Dyck paths counting UD and UUD instances.


\item \url{https://arxiv.org/pdf/2303.09605.pdf} Free action(?) CSP symplectic tableau
\item \url{https://arxiv.org/pdf/2212.13588.pdf} New CSP on r-fans of Dyck paths, and promotion.

\item \url{https://arxiv.org/pdf/2308.10456.pdf} QSym powersum expansion of $Y_\alpha$ (Young's qsym schur)

\item \url{https://arxiv.org/pdf/2212.07343.pdf} Twisted characters. (Similar to N. Kumari's results).
\item \url{https://arxiv.org/pdf/2212.12477.pdf} twisted character factorizations, Nishu


\item \url{https://arxiv.org/pdf/2306.00336.pdf} Crystals for shifted keys

\item \url{https://doi.org/10.1016/j.ejc.2023.103688} P-partitions, weighted. (New)

\item \url{https://arxiv.org/pdf/2102.00935.pdf} Kostka cone
\item \url{https://arxiv.org/pdf/2304.07439.pdf} Q-Kostka polynomials
\item \url{https://arxiv.org/pdf/2303.10664.pdf} Spin Kostka polynomials
\item \url{https://arxiv.org/pdf/2210.10158.pdf} Degrees of stretched Kostka polys in all types

\item \url{https://arxiv.org/pdf/2303.09019.pdf} Back stable quasisymmetric functions and P-partitions. Also a new basis.

\item \url{https://arxiv.org/pdf/2302.07239.pdf} Jacobi-Trudi over finite fields.

\item \url{https://arxiv.org/pdf/2302.05903.pdf} Averaging symmetric functions


\item \url{https://arxiv.org/pdf/2312.16824.pdf} Nabla of monomial
\item \url{https://arxiv.org/abs/2302.04136} delta conj at q=-1
\item \url{https://arxiv.org/pdf/2206.11760.pdf} Delta conj.


\item \url{https://www.mat.univie.ac.at/~slc/wpapers/FPSAC2021/52Kushwaha.pdf} Interesting LR-coeffs from key polys, and saturation.



\item \url{https://arxiv.org/pdf/2301.12110.pdf} Free Fermionic Schur functions (unifies lots of known families!)

\item \url{https://www.sciencedirect.com/science/article/pii/S0021869320302982} Symplectic Kostka-Foulkes

\item \url{https://www.combinatorics.org/ojs/index.php/eljc/article/view/v30i1p5} Flagged Schur polynomial duality.

\item \url{https://www.emis.de/journals/SIGMA/2020/130/sigma20-130.pdf} Plane partition CSP (Sam Hopkins)


\item \url{https://arxiv.org/pdf/2301.02203.pdf} Almost every character value in Sn is divisible by k, as n tend to infinity.

\item \url{https://arxiv.org/pdf/2301.00175.pdf} Littlewood identities, ASM.

\item \url{https://arxiv.org/pdf/2301.00225.pdf} Kiritchenko - Mitosis, Kogan faces.

\item \url{https://arxiv.org/pdf/2212.09419.pdf} Butler's conjecture, (modified Macdonald polys)


\item \url{https://arxiv.org/pdf/2212.06885.pdf} Parking function polytope


\item \url{https://arxiv.org/pdf/2212.05665.pdf} 3D Jack polynomials.

\item \url{https://arxiv.org/pdf/2212.04932.pdf} Wachs permutations.


\item \url{https://doi.org/10.37236/10941} Skew Schur CSP.

\item \url{https://arxiv.org/pdf/1204.2484.pdf} LR coeffs stuff

\item \url{https://arxiv.org/pdf/2210.17476.pdf} Q-sym powersum, type 3.

\item \url{https://doi.org/10.1006/jabr.1994.1064} GT-patterns for classical groups (Proctor)

\item \url{https://www.mat.univie.ac.at/~slc/wpapers/s86jackson.html} RSK and reverse of permutation.

\item \url{https://arxiv.org/pdf/2206.03054.pdf} LR-coeffs properties.

\item \url{https://arxiv.org/pdf/2311.12216.pdf} m-Symmetric Macdonalds 


\item \url{https://link.springer.com/article/10.1007/s00026-021-00563-2?utm_source=toc&utm_medium=email&utm_campaign=toc_26_26_1&utm_content=etoc_springer_20220413} P,w-partitions ad positive differences.

\item \url{https://doi.org/10.4153/CMB-2011-105-7} Symplectic hook-content

\item \url{https://www.mat.univie.ac.at/~slc/wpapers/FPSAC2021/58Grinberg.pdf} New bases for QSym, Grinberg, Vassilieva.

\item \url{https://www.mat.univie.ac.at/~slc/wpapers/FPSAC2021/37Pfannerer.pdf} Descent refinement of Murnaghan--Nakayama.


\item \url{https://www.combinatorics.org/ojs/index.php/eljc/article/view/v28i2p45/pdf} Lattice polytopes 
from Schur and Grothendieck polynomials.

\item \url{https://www.combinatorics.org/ojs/index.php/eljc/article/view/v8i1r16/pdf}

\item \url{https://www.math.ucla.edu/~pak/papers/Ribbon1.pdf} Skew shape asymptotics---Survey by Igor Pak.

\item \url{http://www.birs.ca/workshops/2019/19w5131/files/Frieden.pdf} q=-1 on Kostka-Foulkes.

\item \url{http://www.math.columbia.edu/~woit/notes12.pdf} Nice Weyl character formula notes, for Lie stuff.

\item \url{https://www.combinatorics.org/ojs/index.php/eljc/article/view/v26i2p28/7842} Cyclic Demazure modules

\item \url{https://www.combinatorics.org/ojs/index.php/eljc/article/view/v26i2p27} Super RSK

\item \url{https://doi.org/10.1023/A:1008662732304} Nice paper with plethysm, Kostka--Foulkes at roots of unity, etc.
See also \cite{DoranIV1997}.


\todo{
Does this have CSP?
https://arxiv.org/pdf/1911.10703.pdf
Kreweras: https://arxiv.org/pdf/2001.01449.pdf
}


\item \url{https://www.mat.univie.ac.at/~slc/wpapers/FPSAC2019/96.pdf} CSP on double cosets.


\item \url{https://www.combinatorics.org/ojs/index.php/eljc/article/view/v30i1p7} Spanning Configurations and Representation Stability

\item \url{https://www.sciencedirect.com/science/article/pii/S0001870820300256} Positive specializations of symmetric Grothendieck polynomials

\item \url{https://alco.centre-mersenne.org/item/10.5802/alco.102.pdf} Set-partition tableaux and representations of diagram algebras

\item \url{https://alco.centre-mersenne.org/articles/10.5802/alco.178/} affine Demazure crystals for q-whittaker.


\item \url{https://www.combinatorics.org/ojs/index.php/eljc/article/view/v29i2p49} A Categorification for the Signed Chromatic Polynomial 

\end{enumerate}
