\metatitle{Topics}
\metadescription{Preliminaries and topics related to symmetric functions}


\section[topicsIndex]{Topics related to symmetric functions}


Here is an overview of various topics which are related to symmetric functions,
but do not fit together with a specific family. Some of the pages 
give a brief overview of the definitions and notation that is used.

We suggest that the reader is somewhat familiar with algebraic combinatorics,
although a few topics is covered under preliminaries below.
Furthermore, if you are a PhD student or researcher interested in symmetric functions,
join our \href{https://www.facebook.com/groups/SymmetricFunctions}{Facebook group}!

\subsection[preliminaries-Links]{Preliminaries}

\begin{itemize}

\item
\hyperref[preliminaries]{Partitions and tableaux}

\item
\hyperref[permutationMatrix]{Permutation matrices and patterns}

\item
\hyperref[q-analogs]{$q$-analogs}

\item
\hyperref[plethysm]{Plethysm}

\item
\hyperref[planePartitions]{Plane partitions}

\item
\hyperref[polynomialRootIntro]{Unimodality, realrootedness and stable polynomials}

\item 
\hyperref[matroid]{Matroids}


\item
\hyperref[general-representation-theory]{Representation theory}

\item
\hyperref[root-systems]{Root systems}

\item
\hyperref[schurPositivity]{Schur positivity}

\item
\hyperref[littlewoodRichardsonModels]{Models for Littlewood--Richardson coefficients}

\item
\hyperref[borderStripTableaux]{Border-strip tableaux and ribbon tableaux}

\item
\hyperref[diagonalHarmonics]{Diagonal harmonics}


\item
\hyperref[latticeModel]{Vertex models and the Yang--Baxter equation}

\item
\hyperref[lagrangeInversion]{Lagrange inversion}

\end{itemize}



\subsection[tableau-op-Links]{Operations on tableaux}

\begin{itemize}

\item
\hyperref[rsk]{The Robinson--Schensted--Knuth correspondence}

\item
\hyperref[crystals]{Crystals}

\item
\hyperref[tableau-operations]{Operations on Young tableaux} (Bender--Knuth, promotion, evacuation, jeu-de-taquin)


\end{itemize}



\subsection[kostka-Links]{Kostka coefficients}

\begin{itemize}

\item
\hyperref[gtpatterns]{Gelfand--Tsetlin polytopes}

\item
\hyperref[kostantPartitionFunction]{The Kostant partition function}

\item
\hyperref[kostkaFoulkes]{Kostka--Foulkes polynomials}

\item
\hyperref[riggedConfigurations]{Rigged configurations}

\end{itemize}



\subsection[csp-Links]{Cyclic sieving}

\begin{itemize}
\item
\hyperref[cyclicSieving]{The cyclic sieving phenomenon}

\end{itemize}



\subsection[misc-links]{Miscellaneous}

\begin{itemize}

\item 
\hyperref[unitIntervalAO]{Acyclic orientations and Touchard--Riordan polynomials}

\item 
\hyperref[personal-minor-problems]{Minor problems related to my research}

\item 
\href{http://www-math.mit.edu/~rstan/papers/ubc.pdf}{Research story: R. Stanley --- How the upper bound conjecture was proved}

\item 
\href{https://www.emis.de/journals/SLC/wpapers/s54Ahaglund.pdf}{Research story: J. Haglund --- The genesis of the Macdonald  polynomial statistic}

\item
\href{https://arxiv.org/pdf/2105.07884.pdf}{R. Stanley --- Enumerative and Algebraic combinatorics in the 1960's and 1970's.}

\end{itemize}


\section[software]{Software resources}


\begin{itemize}

\item \icon{database} \href{https://math.depaul.edu/~bridget/patterns.html}{Bridget Tenner's} permutation pattern database.

\item \icon{file-code} \href{https://github.com/PerAlexandersson/Mathematica-packages}{Mathematica packages (Per Alexandersson)}
Packages for Catalan objects, Symmetric functions, posets, tableaux and unicellular chromatic calculations.


\end{itemize}


\subsection[general-useful-things]{Useful links}

\begin{itemize}

\item \icon{sun} \href{https://www.doi2bib.org/}{Doi 2 bib.} This tool is extremely useful.

\item \icon{gem} Maria Gillespie's \href{http://www.mathematicalgemstones.com/}{Mathematical gemstones}.

\item  \icon{file} Overview of \href{https://www.aimath.org/WWN/kostka/}{generalized Kostka Polynomials}.

\item \icon{file} \href{https://faculty.math.illinois.edu/~west/grammar.html}{Grammar and math} by D. B. West.

\item \icon{file} \href{https://arxiv.org/pdf/1612.04888.pdf}{A primer of mathematical writing} by S. G. Krantz.

\end{itemize}



\section[suggestedBooks]{Suggested books}


For an introduction to the topic of symmetric functions and algebraic combinatorics,
I suggest the following books. 
Richard Stanley, \emph{Enumerative Combinatorics II} \cite{StanleyEC2}, 
Lynne Butler, \emph{Subgroup Lattices and Symmetric Functions} \cite{Butler1994}, 
Bruce Sagan, \emph{The Symmetric Group} \cite{Sagan2001}, 
William Fulton, \emph{Young Tableaux} \cite{Fulton1997} and 
Ian Macdonald, \emph{Symmetric functions and Hall polynomials} \cite{Macdonald1995}.

Other related books:
David Bressoud, \emph{Proofs and confirmations: the story of 
the alternating sign matrix conjecture} \cite{Bressoud1999}, which is a very nice reading.


% 
% \section[files-links]{Files}
% 
% These files are rather old --- I intend to update them some time in the future
% when the catalog include most of the quasisymmetric families.
% 
% \begin{itemize}
% \item \icon{fa-file} \href{polynomial-graph-relations.pdf}{Relations graph}
% 
% \item \icon{fa-file} \href{polynomial-graph-positivity.pdf}{Positivity graph}
% \end{itemize}

