\metatitle{Tutte symmetric functions}
\metadescription{Background on Tutte symmetric functions.}



\section[tutteSymmetric]{Tutte symmetric functions}


\begin{polydata}{tutteSymmetric}
  Name   & Tutte symmetric functions \\
  Space  & Sym \\
  Basis  & False \\
  Rating & 1 \\
  Bib    & Stanley1998 \\
  Year   & 1998 \\
  SignedIn & completeMultipartite | CrewSpirkl2020x | sign_rule=i-maximal partitions; proof=Thm. 2 \\
\end{polydata}

The \defin{Tutte symmetric functions} were introduced by
\name{Richard Stanley} \cite[Def.~3.1]{Stanley1998}, and generalize the
chromatic symmetric functions.
See also \cite{AlistePrietoCrewSpirklZamora2020x} for a vertex-weighted generalization.

The Tutte symmetric functions are indexed by graphs, and defined as
\begin{equation}
 \tutte_G(\xvec;t) \coloneqq \sum_{\pi \vdash V(G)} (1+t)^{a(\pi)} \tilde{\monomial}_{\lambda(\pi)}
\end{equation}
where the $\tilde{\monomial}$ denote augmented monomial symmetric functions,
and the sum is taken over all set-partitions of the vertex set of $G$.
Here, $a(\pi)$ denotes the number of \defin{attacking edges} --- edges where both endpoints are in the
same block of $\pi$.

Alternatively, for a graph on $n$ vertices, we have
\begin{equation}
 \tutte_G(\xvec;t) \coloneqq \sum_{\kappa : V(G) \to \setN} (1+t)^{m(\kappa)} x_{\kappa(1)} \dotsm
 x_{\kappa(n)}
\end{equation}
where the sum is over all vertex colorings of $G$, and $m(\kappa)$ counts
the number of monochromatic edges in the coloring.

Note that $\tutte_G(\xvec; -1) = \chrom_G(\xvec)$, that is, we recover the
\hyperref[chromaticQuasisymmetric]{chromatic symmetric function} at $t=-1$.

\begin{example}
For the graph $K_2$ with one edge,
\[
 \tutte_{K_2}(\xvec;t) =
 \tilde{\monomial}_{11} + (1+t)\tilde{\monomial}_{2}.
\]
At $t=-1$, only the proper colorings remain, giving
$\chrom_{K_2}(\xvec) = \tilde{\monomial}_{11}$.
\end{example}


A quasisymmetric version of the Tutte symmetric functions
was introduced in \cite{AwanBernardi2016}.
In \cite[Thm. 7.15]{AlexanderssonSulzgruber2019},
we give the expansion of these polynomials in the quasisymmetric powersum basis.


See \cite{CrewSpirkl2020x} for more results on the Tutte symmetric functions.
\name{Logan Crew} and \name{Sophie Spirkl} also give graph-theoretic
interpretations for plethysms based on chromatic and Tutte symmetric
functions \cite{CrewSpirkl2022Plethysm}.  This yields short proofs of several
plethystic identities and chromatic-symmetric-function identities.


\subsection[pExpansionTutte]{Powersum expansion}

Stanley's powersum expansion is
\[
 \tutte_G(\xvec;t) = \sum_{S \subseteq E(G)} t^{|S|} \powerSum_{\lambda(S)}(\xvec)
\]
where $\lambda(S)$ denotes the sizes of the connected components induced by $S$.


\subsection[tutteSpanningTrees]{Spanning tree expansion}

A formula for computing $\tutte_G(\xvec;t)$ using spanning trees and spanning
forests is given in \cite{McDonaldMoffatt2012}.
This generalizes the classical formula for computing Tutte polynomials using
spanning trees.
A vertex-weighted version is stated in \cite[Eq. (15)]{AlistePrietoCrewSpirklZamora2020x}.
