
\metatitle{Extended chromatic symmetric functions}
\metadescription{Extended chromatic symmetric functions and connections with Tutte symmetric functions.}


\section[extendedChromatic]{Extended chromatic symmetric functions}


\begin{polydata}{extendedChromatic}
  Name   & Extended chromatic symmetric functions \\
  Space  & Sym \\
  Basis  & False \\
  Rating & 2 \\
  Bib    & CrewSpirkl2019 \\
  Year   & 2019 \\
\end{polydata}

\begin{polydata}{completeMultipartite}
  Name     & Complete multipartite basis \\
  Space    & Sym \\
  Basis    & True \\
  Rating   & 1 \\
  Bib      & CrewSpirkl2020x \\
  Year     & 2021 \\
  Symbol   & $r_\lambda$ \\
  Category & Other \\
\end{polydata}


\name{Logan Crew} and \name{Sophie Spirkl} \cite{CrewSpirkl2019x} introduce a vertex-weighted version of 
the \hyperref[chromaticQuasisymmetric]{chromatic symmetric functions}.
This has the advantage that it fulfills a deletion-contraction relation.
Furthermore, this family of polynomials is also $\omega \powerSum$-positive. 
This can easily be seen from \cite{AlexanderssonSulzgruber2019}.

\name{Azzurra Ciliberti} extends chromatic symmetric homology to
vertex-weighted graphs \cite{Ciliberti2024}.  This categorifies the
Crew--Spirkl invariant and gives a deletion--contraction long exact sequence
lifting the deletion--contraction relation for vertex-weighted chromatic
symmetric functions.

In \cite{CrewSpirkl2020x}, the authors examine a new basis, the
\defin{complete multipartite basis}, whose elements are chromatic symmetric
functions of complete multipartite graphs.  They give combinatorial
interpretations for the change-of-basis coefficients with the monomial basis,
and for chromatic and Tutte symmetric functions expanded in this basis.
More explicitly, if $G_\lambda$ is the complete multipartite graph with
stable-set sizes $\lambda$, then
\[
  r_\lambda \coloneqq X_{G_\lambda}.
\]
For each $d$, the functions $\{r_\lambda:\lambda\vdash d\}$ form a basis for
the degree-$d$ part of the ring of symmetric functions
\cite[Sec.~3]{CrewSpirkl2020x}.

\begin{example*}
In degree $2$, the graph $G_{(2)}$ is edgeless and $G_{(1,1)}$ is the
complete graph $K_2$.  Thus
\[
  r_{(2)}=\monomial_2+2\monomial_{11},
  \qquad
  r_{(1,1)}=2\monomial_{11}.
\]
In degree $3$, direct enumeration of proper colorings gives
\[
\begin{aligned}
  r_{(3)} &= \monomial_3+3\monomial_{21}+6\monomial_{111},\\
  r_{(2,1)} &= \monomial_{21}+6\monomial_{111},\\
  r_{(1,1,1)} &= 6\monomial_{111}.
\end{aligned}
\]
These triangular expansions are the first instances of the fact that the
$r_\lambda$ form a basis.
\end{example*}

The deletion-contraction relation does not extend to the $q$-weighted version with ascents.

\name{Yosuke Sato} introduces a common generalization of vertex-weighted
chromatic symmetric functions and chromatic functions coming from universal
graph series \cite{Sato2024x}.  This gives complete invariants for finite
vertex-weighted graphs when the target graph, or target graph series, is
universal, and it includes a power-sum expansion for the complete-graph
series.

In \cite{AliniaeifardWangvanWilligenburg2021}, the authors study weighted paths with equal extended
chromatic symmetric functions.

In \cite{AliniaeifardWangvanWilligenburg2021}, the authors consider a Tutte-symmetric extension of the
vertex-weighted chromatic symmetric functions.
This generalizes both \hyperref[tutteSymmetric]{Tutte symmetric functions}
and the \hyperref[chromaticQuasisymmetric]{chromatic symmetric functions}. 
They provide a spanning-tree formula for these, which then provides a new
spanning-tree formula for the chromatic symmetric functions.

\name{Farid Aliniaeifard}, \name{Shu Xiao Li}, and
\name{Stephanie van Willigenburg} introduce generalized chromatic functions
of edge-coloured digraphs \cite{AliniaeifardLiWilligenburg2024}.  This
framework contains Stanley's chromatic symmetric functions, extended
chromatic symmetric functions, chromatic quasisymmetric functions, and
\hyperref[pPartition]{$P$-partition} generating functions as special cases.
It also realizes many standard symmetric and quasisymmetric bases as
generalized chromatic functions and gives product and coproduct formulas.

\name{Jeremy L. Martin} and \name{May B. Trist} introduce chromatic
MacMahon symmetric functions for vertex-weighted graphs
\cite{MartinTrist2025x}.  These take values in MacMahon symmetric functions
on two alphabets and refine weighted chromatic information by recording both
cardinalities and weights of vertex subsets.

\name{Nancy Mae Eagles}, \name{Angele Foley}, \name{Alice Huang},
\name{Elene Karangozishvili}, and \name{Annan Yu} introduce
\defin{$H$-chromatic symmetric functions} \cite{EaglesFoleyHuangKarangozishviliYu2022}.
Here a fixed graph $H$ restricts the colorings of a graph $G$.  This gives
new graph invariants, new uniqueness questions, and realizations of the
monomial, power-sum, and elementary bases as $H$-chromatic symmetric
functions of complete multipartite graphs.

\name{Shao Yuan Lin} and \name{Laura Pierson} study distinguishability and
linear independence phenomena for $H$-chromatic symmetric functions
\cite{LinPierson2025HChromaticx}.  For the self-chromatic functions $X_G^G$,
they give evidence and structural results related to tree distinguishability,
prove results for complete bipartite targets, and construct bases of
$\Lambda^n$ from $H$-chromatic symmetric functions for fixed non-complete
target graphs.

\todo{
Is there a quasi-symmetric chromatic version of this spanning-tree formula?
}
